### 2018-08-13 paper.tex
first commit

%%%% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
\documentclass[10pt, sigplan]{acmart}
%%\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For double-blind review submission, w/ CCS and ACM Reference
%\documentclass[sigplan,10pt,review,anonymous]{acmart}\settopmatter{printfolios=true}
%% For single-blind review submission, w/o CCS and ACM Reference (max submission space)
%\documentclass[sigplan,10pt,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For single-blind review submission, w/ CCS and ACM Reference
%\documentclass[sigplan,10pt,review]{acmart}\settopmatter{printfolios=true}
%% For final camera-ready submission, w/ required CCS and ACM Reference
%\documentclass[sigplan,10pt]{acmart}\settopmatter{}

%% Conference information
%% Supplied to authors by publisher for camera-ready submission;
%% use defaults for review submission.
%\acmConference[PL'17]{ACM SIGPLAN Conference on Programming Languages}{January 01--03, 2017}{New York, NY, USA}
%\acmYear{2017}
%\acmISBN{} % \acmISBN{978-x-xxxx-xxxx-x/YY/MM}
%\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
%\startPage{1}

%% Copyright information
%% Supplied to authors (based on authors' rights management selection;
%% see authors.acm.org) by publisher for camera-ready submission;
%% use 'none' for review submission.
\setcopyright{none}
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\copyrightyear{2017}           %% If different from \acmYear

%% Bibliography style

%\bibliographystyle{ACM-Reference-Format}

\citestyle{acmnumeric}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Note: Authors migrating a paper from traditional SIGPLAN
%% proceedings format to PACMPL format must update the
%% '\documentclass' and topmatter commands above; see
%% 'acmart-pacmpl-template.tex'.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% Some recommended packages.
\usepackage{booktabs}   %% For formal tables:
                        %% http://ctan.org/pkg/booktabs
\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
                        %% http://ctan.org/pkg/subcaption
\usepackage{xspace}
\usepackage{graphicx}
\usepackage{ifthen}
\usepackage{pgfplots}
\usepackage{listings}
\usepackage{multirow}
\usepgfplotslibrary{statistics}
\usepackage{dblfloatfix} %enable fig at bottom of page
%\input{macros.tex}

\usepackage{xcolor}
\newcommand{\todo}[1]{\color{orange}\fbox{\bfseries\sffamily\scriptsize TODO:}{\sf\small$\blacktriangleright$\textit{#1}$\blacktriangleleft$}\color{black}}
\newcommand{\sd}[1]{\color{red}\fbox{\bfseries\sffamily\scriptsize Stef:}{\sf\small$\blacktriangleright$\textit{#1}$\blacktriangleleft$}\color{black}}
\newcommand{\sk}[1]{\color{blue}\fbox{\bfseries\sffamily\scriptsize Sophie:}{\sf\small$\blacktriangleright$\textit{#1}$\blacktriangleleft$}\color{black}}
\newcommand{\cba}[1]{\color{purple}\fbox{\bfseries\sffamily\scriptsize Clement:}{\sf\small$\blacktriangleright$\textit{#1}$\blacktriangleleft$}\color{black}}
%\newcommand*{rotatebox{75}}
\input{macros}

\begin{document}

%%%
% Legend definition (so I can change it once for all)
%%%

\def\legBase{Baseline\xspace}
\def\legSmartSyntaxPlug{Syntax\xspace}
\def\legSlangPlug{Plugin\xspace}
\def\legSlang{Slang\xspace}
\def\legRTL{Slang+RTL\xspace}
\def\LLVMMac{LLVM-Mac\xspace}
\def\GCCLinux{GCC-Linux\xspace}
\def\plotHeight{6.8cm}

%%%
% End Legend
%%%

%% Title information
\title[Over Twenty Years of Virtual Machine Development Through Simulation]{Over Twenty Years of Virtual Machine Development and Debugging Through Simulation}

%% Author with single affiliation.
\author{Eliot Miranda}
                                        %% can be repeated if necessary
\affiliation{
  %\position{Position1}
 % \department{VM team}              %% \department is recommended
  \institution{Feenk}            %% \institution is required
 % \streetaddress{Street1 Address1}
  \city{San Francisco}
  %\state{France}
  %\postcode{Post-Code1}
  \country{California}                    %% \country is recommended
}
\email{eliot.miranda@gmail.com}          %% \email is recommended

%% Author with two affiliations and emails.
\author{Cl\'ement B\'era}
\affiliation{
  % \position{}
	\department{Software Languages Lab}              %% \department is recommended
	\institution{Vrije Universiteit Brussel}            %% \institution is required
	\city{Brussel}
  % \state{}
  % \postcode{}
	\country{Belgium}                    %% \country is recommended
}
\email{clement.bera@vub.ac.be}          %% \email is recommended

%% Author with two affiliations and emails.
\author{Elisa Gonzalez Boix}
\affiliation{
  % \position{}
	\department{Software Languages Lab}              %% \department is recommended
	\institution{Vrije Universiteit Brussel}            %% \institution is required
	\city{Brussel}
  % \state{}
  % \postcode{}
	\country{Belgium}                    %% \country is recommended
}
\email{egonzale@vub.ac.be}          %% \email is recommended

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
Fork from Dan Ing VM
Dev in Restr St, Comp through C, simulation by interpreting the Slang.
Evolved through the years with processor simulator and various extensions
We explain how we generate the prod VM and how we simulate it for dev and debugging.
We go through 2 experience example, GC and JIT, and show why we believe how valuable it the infra

\end{abstract}

%% 2012 ACM Computing Classification System (CSS) concepts
%% Generate at 'http://dl.acm.org/ccs/ccs.cfm'.
%\begin{CCSXML}
%<ccs2012>
%<concept>
%<concept_id>10011007.10011006.10011008</concept_id>
%<concept_desc>Software and its engineering~General programming languages</concept_desc>
%<concept_significance>500</concept_significance>
%</concept>
%<concept>
%<concept_id>10003456.10003457.10003521.10003525</concept_id>
%<concept_desc>Social and professional topics~History of programming languages</concept_desc>
%<concept_significance>300</concept_significance>
%</concept>
%</ccs2012>
%\end{CCSXML}

%\ccsdesc[500]{Software and its engineering~General programming languages}
%\ccsdesc[300]{Social and professional topics~History of programming languages}
%% End of generated code

%% Keywords
%% comma separated list
\keywords{Just-in-Time compiler, Virtual machine, Managed runtime, Tools}  %% \keywords are mandatory in final camera-ready submission

%% \maketitle
%% Note: \maketitle command must come after title commands, author
%% commands, abstract environment, Computing Classification System
%% environment and commands, and keywords command.
\maketitle

\section{Introduction}
\label{sec:intro}

\section{Virtual Machine Simulation \& Production Infrastructure}
\label{sec:VMSimulation}

\subsection{Generating the Production Virtual Machine}

\subsection{Simulating the Virtual Machine}

\section{Garbage Collection Development}
\label{sec:GCExp}

default behavior, cloneOnSavenge/GC
Lemming debugging

Levels of Assertions: ST only (debugging code in St), Slang/C, Node.

\section{Just-in-Time Compiler Development}
\label{sec:JITExp}
Back-in-time debugging of mahcine state
Conditional stepping.

\section{Virtual Machine Analysis}
\label{sec:Analysis}
Analysis of caches

Analysis of machine code zone

Analysis of the heap

\section{Discussion, Related Work and Conclusion}

\subsection{Discussion}
Simulation perf and its limitations
Out of the simulation, FFI calls and call-backs

\subsection{Related Work}
Maxine inspectors

RPython toolchain, though it seems they don't do it (Balance abstraction and time to compile/simulate)

\subsection*{Conclusion}

%% Acknowledgments
%%\begin{acks}                            %% acks environment is optional
                                        %% contents suppressed with 'anonymous'
  %% Commands \grantsponsor{<sponsorID>}{<name>}{<url>} and
  %% \grantnum[<url>]{<sponsorID>}{<number>} should be used to
  %% acknowledge financial support and will be used by metadata
  %% extraction tools.
%  This material is based upon work supported by the
%  \grantsponsor{GS100000001}{National Science
%    Foundation}{http://dx.doi.org/10.13039/100000001} under Grant
%  No.~\grantnum{GS100000001}{nnnnnnn} and Grant
%  No.~\grantnum{GS100000001}{mmmmmmm}.  Any opinions, findings, and
%  conclusions or recommendations expressed in this material are those
%  of the author and do not necessarily reflect the views of the
%  National Science Foundation.
%\end{acks}

%% Bibliography
\bibliographystyle{alpha}
\bibliography{sista,rmod,others}

\end{document}

### 2018-08-14 paper.tex
section 3-4 to write. Everything is raw (no pass on the text at all)

@@
 %%%
-% Legend definition (so I can change it once for all)
+% Keyword definition (so I can change it once for all)
 %%%
 
-\def\legBase{Baseline\xspace}
-\def\legSmartSyntaxPlug{Syntax\xspace}
-\def\legSlangPlug{Plugin\xspace}
-\def\legSlang{Slang\xspace}
-\def\legRTL{Slang+RTL\xspace}
-\def\LLVMMac{LLVM-Mac\xspace}
-\def\GCCLinux{GCC-Linux\xspace}
-\def\plotHeight{6.8cm}
-
+\def\openSmalltalkVM{Open-Smalltalk-VM\xspace}
 %%%
@@
 \begin{abstract}
-Fork from Dan Ing VM
-Dev in Restr St, Comp through C, simulation by interpreting the Slang.
-Evolved through the years with processor simulator and various extensions
-We explain how we generate the prod VM and how we simulate it for dev and debugging.
-We go through 2 experience example, GC and JIT, and show why we believe how valuable it the infra
+
+\openSmalltalkVM was originally written in Smalltalk to specify and example the virtual machine (VM) implementation.
+The same code base was changed a little bit to allow Smalltalk-to-C compilation, generating effectively the production VM.
+Two execution models are effectively available, the simulation mode, executing the Smalltalk code on top of a Smalltalk VM, and the production mode, compiling the code to executable through C.
+Simulation is used to develop and debug the VM.
+Production is used to release the VM.
+
+As the VM evolved, by introducing better garbage collector algorithms or a just-in-time compiler, both execution models co-evolved.
+For example, the simulation mode was extended with a processor simulator to simulate the code generated by the just-in-time compiler.
+
+In this paper, we detail the VM simulation infrastructure and we report our experience developing and debugging the VM within it.
+We discuss some of the limitations and how we dealt with it.
+Then, we focus on two specific use-cases, a bug in the just-in-time compiler and the development of a new full garbage collector compaction algorithm, and we show how the simulation infrastructure helped us.
+Lastly, we discuss how we use the simulation mode to perform analysis on the runtime, directing some design decisions we make to tune the VM performance.
 
@@
 
-\section{Virtual Machine Simulation \& Production Infrastructure}
+To specify the language and explain how to write the virtual machine (VM) for it, the Smalltalk-80 crew wrote a Smalltalk VM entirely in Smalltalk \cite{blueBook}.
+In 1995, the same team built an open-source Smalltalk and VM, Squeak \cite{SqueakByExample}.
+The VM was ported from the original specifications however part of the code base was narrowed down to a subset of Smalltalk, called \emph{Slang}, to allow Smalltalk to C compilation \cite{BackToTheFuture}.
+Effectively, the VM could be both simulated by executing the Slang code on top of the Smalltalk VM or compiled to native code to produce the production VM.
+At this point, the VM consisted mainly in an interpreter, a memory manager with a generational garbage collector (GC) and BitBlt, an extension for the user interface bit-based engine.
+These tree parts of the VM were written entirely in Slang.
+A few extra features, such as file management, were written both in Smalltalk for simulation purposes and in C for the production VM.
+
+Over the years, the Squeak VM evolved to give birth recently to \openSmalltalkVM\footnote{https://github.com/OpenSmalltalk/opensmalltalk-vm/}, the default VM for different Smalltalk and Smalltalk-like runtimes (Pharo \cite{PharoByExample}, Squeak \cite{SqueakByExample}, Cuis, Croquet and NewSpeak \cite{NewspeakOopsla}).
+As the VM evolved, the simulator co-evolved as a tool to develop and debug the VM.
+The most significant evolution of the simulator came with the introduction of the Just-In-Time compiler (JIT), \emph{Cogit}, a template-based JIT.
+The existing simulator was not able to interpreter the machine code generated by the JIT and had to be extended by binding various processor simulators (Bochs for x86 \& x64, SkyEye for ARMv6).
+
+In the following section, we explain the VM infrastructure with both the compilation pipeline to generate the production VM and the simulation infrastructure used to develop and debug the VM.
+Section \ref{sec:GCExp} reports our experience developing the full GC with our infrastructure.
+Section \ref{sec:JITExp} explains how we fixed a bug in the Just-in-Time compiler using single stepping in machine code in the simulator.
+Section \ref{sec:Analysis} shows how we abuse the simulator to analyse the runtime and direct our performance decisions.
+Lastly, we discuss some related work and conclude.
+
+\section{Virtual Machine Infrastructure}
 \label{sec:VMSimulation}
 
-\subsection{Generating the Production Virtual Machine}
+As shown in Figure \ref{fig:VMCompilation}, the VM code base is written both in Slang and in C.
+The Core VM code, also known as the object engine, is written in Slang and includes mainly the interpreter, the template-based JIT and the memory manager.
+The platform code, \emph{i.e.,} Operating System dependent code such as file management or I/O is written directly in C.
+%EXPLAIN scientifically in this part - no history. + FIGURE, basic with design.
+
+\begin{figure}[bth!]
+		\centering
+		\includegraphics[width=0.62\linewidth]{figures/VMCompilation}
+		\caption{Cog VM compilation}
+		\label{fig:VMCompilation}
+\end{figure}
+
+The VM executable is generated in a two step process.
+Firstly, the Slang-to-C compiler translates the Slang code to C code in a few files.
+This first compilation pass takes a few seconds.
+Secondly, the C compiler (depending on the platform, LLVM or GCC), translates the C code into an executable.
+This second compilation pass may take several dozens of seconds the first time, then, it depends on what the programmer changed and what the C compiler can cache or not, but it usually also takes a few seconds to recompile the C code.
+The VM can be compiled in two main flavors, interpreter-only or interpreter+JIT.
+Although the version with the JIT is used in production, the interpreter version is convenient for special purposes, for example to debug the garbage collector or to evaluate new language features without dealing with the JIT complexity.
+
+There are multiple reasons why the object engine is written in Slang and not in C.
+One of the reason is that the Slang-to-C compiler, using différents annotations, generates C code slightly different from the Slang code: for example, it duplicates the implementation of specific methods with specific constant operands in Slang to generate more efficient code in the interpreter.
+The main reason is VM simulation.
+By interpreting the Slang code as Smalltalk code, emulating native code using an external processor simulator and simulating the memory using a large byte array, we are able to simulate the whole VM execution.
 
@@
 
+in this section, we first describe briefly the start-up sequence of \openSmalltalkVM using snapshots.
+Then, we detail the simulation of the interpreter-only VM.
+The simulation of the full VM is a superset and is explained in the last subsection.
+
+\paragraph{Start-up from snapshots.} 
+Smalltalk is a snapshot-based language.
+A Smalltalk program is started from a snapshot, \emph{i.e.,} a memory dump of all live objects at a given point in the execution of a program.
+The snapshot includes objects such as the classes, the complied methods in the form of bytecodes and the running processes.
+At start-up, the VM restores the state of all objects in memory and resumes execution in the active process at snapshot time.
+
+When programming with Smalltalk, the programmers usually start ups from a snapshot which contains the core libraries, the development environment and the application developed.
+Development of the application consists essentially in writing and editing code, which effectively installs, modifies and removes compiled methods and classes from the snapshots.
+When the changes are done, the developer takes a new snapshot, which includes its changes.
+Deployment is done from a snapshot containing the deployed application and required core librairies (unused librairies and development tools may be removed from the deployed snapshot).
+
+Starting up the VM for source files has never been possible.
+However, recent work~\cite{PharoBootstrap} allowed one of the Smalltalk runtime to recreate a snapshot from sources, indirectly allowing to start the VM from source files.
+
+\paragraph{Interpreter-only simulation.} 
+As the VM can be compiled with and without the JIT, the VM can also be simulated with or without the JIT.
+
+The heap is simulated as a large contiguous byte array.
+References between objects are effectively indexes inside the byte array instead of pointers.
+All the C variables are simulated as Smalltalk objects.
+They used specific classes, such as \emph{CArrayAccessor} over normal Smalltalk classes, to emulate the C behavior (only array accesses are available in C, not high level iterator APIs for example).
+One of the most complex component is the Stack.
+The Stack is represented in \openSmalltalkVM as a double linked list of stack pages which are maintained by the VM.
+Each stack page is represented as a Smalltalk object.
+
+All the Slang code is implemented in multiple Smalltalk classes, to organise the code.
+For production, at Slang-to-C compilation time, all the code is compiled in a single C file.
+No polymorphism is available at runtime.
+However, polymorphism can be abused for debugging purposes.
+For example, a class is available holding all the garbage collection compaction logic.
+This class might be subclassed with a simulation only version: this version can use normal Smalltalk code over Slang, allowing to express easily specific constraints which leads to assertion failures if not met, such constraints are convenient for debugging.
+
+In Simulation, one of the core feature is to be able to re-use the whole Smalltalk IDE, including the browser, the inspectors and the debugger to develop and debug the VM.
+Most new features can be develop and debug interactively, adding code to the VM at runtime, in the simulation environment, as for normal Smalltalk programs.
+
+\paragraph{JIT simulation.}
+
+In addition to the interpreter simulator, simulating the JIT requires to simulate the execution of native code it generates.
+The JIT itself is written in Slang and simulated with the Smalltalk execution model.
+To simulate the machine code, the start of the byte array representing the memory now holds the machine code generated and installed by the JIT.
+Bindings to processor simulator libraires (Bochs for x86 and x64, Skyeye for ARMv6) were implemented so that the machine code can be simulated.
+Calls in-between slang code and direct machine code are a little bit trickier to simulate.
+Calls from machine code to slang code are implemented by using multiple invalid processor instructions, leading to a trap in the processor simulator.
+This trap is caught is the VM simulator, which then resumes Slang simulation by calling the method corresponding to the invalid processor instruction.
+Calling machine code from Slang requires to start the simulation and then launch an exception to stop Slang simulation.
+The processor simulator can be start in two different ways.
+It can either start simulating code until it meets an invalid instruction or simulate one instruction at a time.
+The second version is slower, but allows to implement specific debugging features, such as conditional breakpoints in-between each machine instruction.
+
+\subsection{In-image compilation}
+
+Simulating the whole VM requires going through the whole start-up sequence: loading the snapshot, running code registered in the start-up sequence and resuming the user interface.
+The whole start-up takes around 15 seconds on a recent Macbook pro.
+While developing the template JIT, this start-up time may still be too long and move the live programming experience to an edit-compile-run cycle.
+To work around this problem, we implemented a tool called \emph{In-image compilation}.
+In-image compilation basically allows to call the JIT as a Smalltalk library on a given bytecode compiled method to generate the corresponding machine code and display the disassembly.
+Since the JIT is template-based, in-image compilation is very convenient to develop and optimize each of the JIT templates.
+To generate the machine code, the JIT has to access specific objects (the compiled method, the literals, known objects such as true, false or nil) as if they were in the simulated memory.
+To work around this, we built a facade, which is started on the compiled method and only includes the subset of objects required by the JIT to generate the machine code.
+
+\subsection{Virtual Machine Simulation Limitations}
+
+\paragraph{Performance.}
+The first limitation is due to the simulation performance.
+The interpreter-only is simulator is around 200 times slower to execute code than the normal VM.
+With the JIT and processor simulation enabled, without specific debugging options such as conditional breakpoints in between machine instructions, simulation drops to around 500 times slower than the normal VM.
+This means for example that if a GC bug happens in an application 15 minutes after start-up, it will take 50 hours to reproduce in the interpreter-only simulator.
+Bugs in the jitted code are worst.
+Fortunately, we work around this problem by using snapshots and the interpreter-only simulator for GC bugs.
+In general, once we are able to reproduce a bug in the production VM, we try to snapshot the runtime just before it crashes.
+The VM simulator can then be started just before the crash and the debugging tools can be used after only several dozens of seconds.
+If the bug is unrelated to the JIT, the interpreter-only simulator can be used and it is a little bit quicker to execute code.
+
+\paragraph{Calls to external code.}
+Although most of the GC and JIT development and bug resolution can be done in the simulator, specific developments and bugs cannot.
+Basically, any calls outside of the machine code generated by the JIT and the Slang code cannot be simulated.
+For specific small parts of the VM, such as file management, we extended the simulator, effectively duplicating the code base with the C code, to support those features in simulation.
+However, there is no solution in the general case: we cannot afford to simulate both the compiled C code and the jitted code on the processor simulator, that would be horribly slow, and specific behaviors in the machine code not present in the code generated by the JIT cannot be simulated (Access to C variables, OS, etc.).
+
+The main limitation we have is with Foreign Function Interfaces (FFI).
+Most bugs we have with FFI are due to specific interaction between call-backs, low-level assembly FFI specific glue code and moving objects.
+Such bugs cannot be debugged, so far, with our simulation infrastructure and we have to rely on gdb/lldb.
+
 \section{Garbage Collection Development}
 \label{sec:GCExp}
+
+REf to papers
+
+Example of the GC impl.
+Debugging mainly from ST, standard dev tools.
 
@@
 \label{sec:JITExp}
-Back-in-time debugging of mahcine state
+
+\subsection{Debugging crashes with conditional breakpoints in machine code}
+Back-in-time debugging of machine state
 Conditional stepping.
 
-\section{Virtual Machine Analysis}
+Ref Sista
+
+\subsection{Optimizing the templates with in-image compilation}
++ ref low overhead RB AND explain template optimized with in-image compilation.
+
+\section{Virtual Machine Analysis: Directing performance decisions}
 \label{sec:Analysis}
-Analysis of caches
 
-Analysis of machine code zone
+A side-effect of VM simulation, and specifically to be able to interrupt the simulation and introspect the simulated memory and simulation specific objects, is to be able to analyse the runtime with scripts written on-the-fly.
 
-Analysis of the heap
+\subsection{Analysis example}
 
-\section{Discussion, Related Work and Conclusion}
+One of the first analysis we run is to stop the simulation when the machine code zone reached 1Mb.
+We then iterated over the machine code zone and investigated what was in.
+As show in Table \ref{tab:generalAnalysis}, 1752 were compiled to machine code by the template JIT, 6352 sends\footnote{We use the Smalltalk terminology, send, to discuss virtual calls since we are talking about Smalltalk.} are present but 2409 of them are not linked (basically, they have never been used).
 
-\subsection{Discussion}
-Simulation perf and its limitations
-Out of the simulation, FFI calls and call-backs
+\begin{table} [th]
+\centering
+\begin{tabular}{l|r}
+	Number of methods  & 1752 \\
+  \hline
+	Number of sends  &  6352 \\
+  \hline
+	Average number of sends per method  & 3.63  \\
+  \hline
+	Number of unlinked sends  &  2409 \\
+  \hline
+	Percentage of unlinked sends  & 37.9\%  \\
+  \end{tabular}
+\caption{General Machine Code Zone Analysis\vspace{-0.5cm}}
+\label{tab:generalAnalysis}
+\end{table}
 
-\subsection{Related Work}
-Maxine inspectors
+Further analysis, in Table \ref{tab:polyAnalysis}, confirms Urs H\"{o}lzle analysis \cite{PICSelf}: around 90\% of used send sites are monomorphic, 9\% are polymorphic (up to 6 different cases in our implementation) and the remaining \% is megamorphic.
 
-RPython toolchain, though it seems they don't do it (Balance abstraction and time to compile/simulate)
+\begin{table} [th]
+\centering
+\begin{tabular}{l|c|c}
+  	& Number of sends & \% of linked sends \\
+  \hline
+	Monomorphic & 3566 & 90.4 \% \\
+	Polymorphic & 307 & 07.8 \% \\
+	Megamorphic & 70 & 01.8 \% \\
+  \end{tabular}
+\caption{Polymorphism Inline Cache Analysis\vspace{-0.5cm}}
+\label{tab:polyAnalysis}
+\end{table}
+
+The code for these analysis are detailed in the Section "Let Me Tell You All About It, Let Me Quantify" of the blog post "Build me a JIT as fast as you can"\footnote{http://www.mirandabanda.org/cogblog/2011/03/01/build-me-a-jit-as-fast-as-you-can/}.
+
+\subsection{Directing the VM behavior}
+
+The results of the analysis are used to direct performance design decisions on the VM. In this section we describe how the analysis impacted a design called "Early Polymorphic inline cache promotion".
+
+We designed the polymorphic inline caches (PICs) with two implementations:
+\begin{itemize}
+	\item \emph{Closed PICs:} Such caches can deal with up to 6 cases, and are basically implemented as a jump table.
+	\item \emph{Open PICs:} Such caches can deal with any number of cases, they consist of three probes searching the global look-up cache, a hash map shared with the interpreter, and fall back into a standard look-up routine if there is a cache miss.
+\end{itemize}
+
+One idea we had was to promote a monomorphic inline cache straight to an open PIC if available, and create the closed PIC only if no open PIC is available for the given selector.
+The benefit is avoiding lots of code space modifications and an allocation.
+The downside is replacing faster closed PIC dispatch with slower open PIC dispatch.
+The question is how many send sites would be prematurely promoted to megamorphic, or how many closed PICs have selectors for which there are open PICs.
+Analysing the question is easy in our context.
+
+The analysis result showed that 17\% of polymorphic send sites would get prematurely promoted.
+So we have implemented a simple sharing scheme.
+The JIT maintains a linked list of open PICs, and before it creates a closed PIC for a send site it will patch it to an open PIC if the list contains one for the send?s selector.
+
+\section{Related Work and Conclusion}
+
+Many VM developers developed different tools to help them being more efficient, but they rarely publish about it. We focus in this section on two related work.
+
+\paragraph{Maxine Inspectors.} The Maxine inspectors \cite{MaxineInspector} were demonstrated at OOPSLA'18.
+They allow to inspect the running state of the VM while it runs for debugging purposes.
+One of the main difference with out design is that the Maxine VM is metacircular, they do not have a simulation and a production mode as we do.
+We believe having two different modes allows us to easily generate a production artifact while still having nice debugging features.
+Having a full metacircular VM would be interesting.
+However, so far, most VMs used in production (Java, Javascript, etc.), even after the huge recent investments in the Javascript VMs by the four major web vendors, are still compiling through the C/C++ compiler and not metacircular.
+Hence, although a metacircular VM has interesting advantages, it is not clear it is that convenient to build a VM in such a way.
+
+\paragraph{RPython toolchain.}  The RPython toolchain \cite{RPythonToolchain} was designed and implemented quite similarly to \openSmalltalkVM.
+Most of the VM code is written in RPython, a restricted Python, instead of Slang, and some leftovers are written in plain C.
+The main difference is that RPython is much closer to Python than Slang is to Smalltalk, RPython allows higher level structures such as dictionaries to be used.
+The design decision comes with its set of advantages and drawbacks.
+The key advantage is that the RPython code feels like Python code and is relatively quite easy to read write.
+The main drawback is that RPython to C compilation takes way longer than the Slang to C compilation (up to 40 minutes in a recent Macbook pro for the RSqueak VM \cite{RSqueak}, instead of several seconds for Slang).
+
+Although the RPython code can be executed as normal Python code, for some reasons, the developers seem to think it is not worth to do it, mostly because executing code in such a way is very slow.
+The overall architecture of the RPython toolchain is different, which leads to a longer time to reach peak performance (though their peak performance is at least in theory better than with a template JIT as \openSmalltalkVM features).
+This time may be very significant in simulation mode.
+In addition, RPython was originally designed for Python, which does not feature snapshot by default, so they cannot abuse snapshots to work around the simulation slow performance.
 
 \subsection*{Conclusion}
+
+We introduced and discussed the \openSmalltalkVM simulation infrastructure, used to develop and debug the VM.
+We believe it is a powerful tool allowing to reduce our development time and to allow to fix bugs quickly.
+In the near future, we plan to extend the simulator with customizable development tools, especially the moldable inspectors and debuggers \cite{MoldableInspector, MoldableDebugger}, to have a fancy user interface on top of the current simulation model, currently quite tricky to apprehend by new developers.
 
@@
 \bibliographystyle{alpha}
-\bibliography{sista,rmod,others}
+\bibliography{sista}
 

### 2018-08-14 paper.tex
added most sections. I do not really like the paper right nwo though

@@
 		\centering
-		\includegraphics[width=0.62\linewidth]{figures/VMCompilation}
+		\includegraphics[width=0.6\linewidth]{figures/VMCompilation}
 		\caption{Cog VM compilation}
@@
 
-REf to papers
+Recently, to evaluate new garbage collection algorithm against standard algorithm, we implemented a Mark-Sweep in addition to the existing Mark-Compact for old space garbage collection.
+The whole implementation was done in the simulator, and only when it was working there, we compiled it to C and it was working out of the box.
+To describe the implementation process, we need to discuss briefly first the assertion levels in the VM.
+Then we will show how we debugged the algorithm.
 
-Example of the GC impl.
-Debugging mainly from ST, standard dev tools.
+\paragraph{Assertions.}To stabilise our code base and find easily production bugs, all the code based is annotated with assertion, ensuring a specific state is as the VM developer would expect it to be, and stopping the program execution if not.
+We have basically three levels of assertions.
+Assertions written in plain Smalltalk are convenient since more complex constraints can be expressed, but they can be performed only in the simulator.
+Assertions written in Slang are also verified in simulation mode, but they are compiled to C.
+The C compiler, based on a compilation flag, then picks if it wants to compile the VM with or without assertions.
+The VM with assertions is used for debugging and to recreate snapshots of runtime juste before they crash.
+The VM without assertions is the production VM.
 
-default behavior, cloneOnSavenge/GC
-Lemming debugging
-
-Levels of Assertions: ST only (debugging code in St), Slang/C, Node.
+\paragraph{Simulation.} Once we had partially written the new Sweep algorithm, we started the simulator.
+Since the algorithm was partially written, we could just write the missing pieces inside the debugger, installing the new code at runtime, as one can do it in Smalltalk.
+In addition, the simulator has an interesting property: each time it performs a GC, either a scavenge or an old space collection, it first copies the simulated memory (\emph{i.e.,} the heap), performs the GC there, and if no assertion fails, it then performs the GC on the original version.
+This means that if the algorithm was not working, an assertion would fail or an error such as incorrect memory access would happen in a copy of the heap.
+The simulator can then reproduce the exact same crash as many times as we want duplicating again and again the same original heap.
+This is very convenient to debug specific GC bugs because often in GC bugs, when the bug happens, the memory is already corrupted and it's quite difficult to track back where the bug comes from without rewinding the memory state.
 
@@
 
+We firstly discuss how we use the simulator to debug crashes in deployed applications, then how we use in-image compilation to develop the JIT itself.
+
 \subsection{Debugging crashes with conditional breakpoints in machine code}
+
+TOWRITE FROM MY YOUTUBE VIDEO.
 Back-in-time debugging of machine state
@@
 \subsection{Optimizing the templates with in-image compilation}
-+ ref low overhead RB AND explain template optimized with in-image compilation.
+
+A few years ago, we added support in the VM for read-only objects \cite{ROObject}.
+Read-only objects were critical performance-wise for specific customers using them in the context of object databases.
+To maximize the performance, we changed the templates in the JIT compiler for the different memory stores.
+Since the template JIT has to generate very quickly machine code, it goes through the code three times: (1) it scans the bytecode, (2) it generates abstract instructions from the bytecode and (3) it generates the native instructions from the abstract instructions.
+The abstract instructions are mapped almost one to one to machine instructions, their main purpose is to compact the generated code by removing Nops used for jump targets or finding out out the size of jumps.
+
+To optimize the template, we use the in-image compilation framework.
+We selected a method with a single store to make it simple.
+We requested the JIT to generate the machine code and changed the store template until the machine code generated was the exact instructions we wanted.
+It is possible, in in-image compilation, to use the Smalltalk debugger on the JIT code itself to inspect the JIT state and fix the code on -the-fly without any recompilation process.
+Once we went through the few store templates (there are a few different templates for optimizations purposes, for example, storing a constant integer does not require a garbage collector write barrier check),
 
@@
 So we have implemented a simple sharing scheme.
-The JIT maintains a linked list of open PICs, and before it creates a closed PIC for a send site it will patch it to an open PIC if the list contains one for the send?s selector.
+The JIT maintains a linked list of open PICs, and before it creates a closed PIC for a send site it will patch it to an open PIC if the list contains one for the send's selector.
 

### 2018-08-14 paper.tex
pass over the text

@@
 \openSmalltalkVM was originally written in Smalltalk to specify and example the virtual machine (VM) implementation.
-The same code base was changed a little bit to allow Smalltalk-to-C compilation, generating effectively the production VM.
-Two execution models are effectively available, the simulation mode, executing the Smalltalk code on top of a Smalltalk VM, and the production mode, compiling the code to executable through C.
+The same code base was changed a little bit to allow Smalltalk-to-C compilation to generate the production VM through the C compiler.
+Two execution models are effectively available, simulation, where the Smalltalk code is executed on top of a Smalltalk VM, and production, where the same code is compiled to executable code through the C compiler.
 Simulation is used to develop and debug the VM.
@@
 As the VM evolved, by introducing better garbage collector algorithms or a just-in-time compiler, both execution models co-evolved.
-For example, the simulation mode was extended with a processor simulator to simulate the code generated by the just-in-time compiler.
+For example, the simulation infrastructure was extended with a processor simulator to simulate the code generated by the just-in-time compiler.
 
 In this paper, we detail the VM simulation infrastructure and we report our experience developing and debugging the VM within it.
-We discuss some of the limitations and how we dealt with it.
-Then, we focus on two specific use-cases, a bug in the just-in-time compiler and the development of a new full garbage collector compaction algorithm, and we show how the simulation infrastructure helped us.
-Lastly, we discuss how we use the simulation mode to perform analysis on the runtime, directing some design decisions we make to tune the VM performance.
+We mention some of the limitations and how we worked around them.
+We discuss specifically how we use the VM simulator to develop and debug two core VM components, the garbage collector and the just-in-time compiler.
+Then, we discuss how we use the simulation infrastructure to perform analysis on the runtime, directing some design decisions we make to tune the VM performance.
 
@@
 In 1995, the same team built an open-source Smalltalk and VM, Squeak \cite{SqueakByExample}.
-The VM was ported from the original specifications however part of the code base was narrowed down to a subset of Smalltalk, called \emph{Slang}, to allow Smalltalk to C compilation \cite{BackToTheFuture}.
-Effectively, the VM could be both simulated by executing the Slang code on top of the Smalltalk VM or compiled to native code to produce the production VM.
-At this point, the VM consisted mainly in an interpreter, a memory manager with a generational garbage collector (GC) and BitBlt, an extension for the user interface bit-based engine.
-These tree parts of the VM were written entirely in Slang.
-A few extra features, such as file management, were written both in Smalltalk for simulation purposes and in C for the production VM.
+The VM was ported from the original specifications.
+Part of the code base was however narrowed down to a subset of Smalltalk, called \emph{Slang}, to allow Smalltalk to C compilation \cite{BackToTheFuture} and generate an efficient VM through the C compiler.
+Effectively, the VM could be both simulated by executing the Slang code on top of the Smalltalk VM or compiled through the C compiler to native code to generate the production VM.
+At this point, the VM consisted mainly in:
+\begin{itemize}
+\item an interpreter with a spaghetti stack, 
+\item a memory manager with a generational garbage collector (GC) and
+\item BitBlt, an extension for the user interface bit-based engine.
+\end{itemize} 
+These three parts of the VM were written entirely in Slang. A few extra features, such as file management, were written both in Smalltalk for simulation purposes and in C for the production VM.
 
-Over the years, the Squeak VM evolved to give birth recently to \openSmalltalkVM\footnote{https://github.com/OpenSmalltalk/opensmalltalk-vm/}, the default VM for different Smalltalk and Smalltalk-like runtimes (Pharo \cite{PharoByExample}, Squeak \cite{SqueakByExample}, Cuis, Croquet and NewSpeak \cite{NewspeakOopsla}).
+Over the years, the Squeak VM evolved to give birth recently to \openSmalltalkVM\footnote{https://github.com/OpenSmalltalk/opensmalltalk-vm/}, the default VM for different Smalltalk and Smalltalk-like runtimes such as Pharo \cite{PharoByExample}, Squeak \cite{SqueakByExample}, Cuis, Croquet and NewSpeak \cite{NewspeakOopsla}.
 As the VM evolved, the simulator co-evolved as a tool to develop and debug the VM.
-The most significant evolution of the simulator came with the introduction of the Just-In-Time compiler (JIT), \emph{Cogit}, a template-based JIT.
-The existing simulator was not able to interpreter the machine code generated by the JIT and had to be extended by binding various processor simulators (Bochs for x86 \& x64, SkyEye for ARMv6).
+The most significant evolution of the simulator came with the introduction of the Just-In-Time compiler (JIT).
+The existing simulator was not able to simulate the machine code generated by the JIT and had to be extended by binding multiple processor simulators (Bochs for x86 \& x64, SkyEye for ARMv6).
 
 In the following section, we explain the VM infrastructure with both the compilation pipeline to generate the production VM and the simulation infrastructure used to develop and debug the VM.
-Section \ref{sec:GCExp} reports our experience developing the full GC with our infrastructure.
-Section \ref{sec:JITExp} explains how we fixed a bug in the Just-in-Time compiler using single stepping in machine code in the simulator.
-Section \ref{sec:Analysis} shows how we abuse the simulator to analyse the runtime and direct our performance decisions.
+Section \ref{sec:GCExp} and Section \ref{sec:JITExp}  reports our experience developing respectively the full GC and the JIT with our infrastructure.
+Section \ref{sec:Analysis} shows how we abuse the simulator to analyse the runtime and direct design decisions to improve performance.
 Lastly, we discuss some related work and conclude.
@@
 As shown in Figure \ref{fig:VMCompilation}, the VM code base is written both in Slang and in C.
-The Core VM code, also known as the object engine, is written in Slang and includes mainly the interpreter, the template-based JIT and the memory manager.
+The Core VM code (the interpreter, the JIT and the memory manager), is written in Slang.
 The platform code, \emph{i.e.,} Operating System dependent code such as file management or I/O is written directly in C.
@@
 The VM executable is generated in a two step process.
-Firstly, the Slang-to-C compiler translates the Slang code to C code in a few files.
-This first compilation pass takes a few seconds.
+Firstly, the Slang-to-C compiler translates the Slang code to C code, generating a few files.
+This first step takes several seconds.
 Secondly, the C compiler (depending on the platform, LLVM or GCC), translates the C code into an executable.
-This second compilation pass may take several dozens of seconds the first time, then, it depends on what the programmer changed and what the C compiler can cache or not, but it usually also takes a few seconds to recompile the C code.
-The VM can be compiled in two main flavors, interpreter-only or interpreter+JIT.
-Although the version with the JIT is used in production, the interpreter version is convenient for special purposes, for example to debug the garbage collector or to evaluate new language features without dealing with the JIT complexity.
+This second step may take several dozens of seconds the first time, then, depending on what the programmer changed and what the C compiler have cached, it usually only takes a few seconds to recompile the C code.
 
-There are multiple reasons why the object engine is written in Slang and not in C.
-One of the reason is that the Slang-to-C compiler, using différents annotations, generates C code slightly different from the Slang code: for example, it duplicates the implementation of specific methods with specific constant operands in Slang to generate more efficient code in the interpreter.
-The main reason is VM simulation.
-By interpreting the Slang code as Smalltalk code, emulating native code using an external processor simulator and simulating the memory using a large byte array, we are able to simulate the whole VM execution.
+The VM can be compiled in two main flavours, interpreter-only or interpreter+JIT.
+Although the version with the JIT is used in production, the interpreter version is convenient for special purposes.
+For example, debugging the garbage collector or evaluating new language features without dealing can be done in the interpreter-only VM, avoiding the JIT complexity.
+
+There are multiple reasons why we keep writing the core components in Slang and not in C/C++.
+Most of them are small details.
+For instance, the Slang-to-C compiler generates C code slightly different from the Slang code using different annotations.
+It can for example duplicate the implementation of specific methods with specific constant operands to generate more efficient code in the interpreter.
+The main reason why we keep using Slang over C/C++ is VM simulation.
+By interpreting the Slang code as Smalltalk code, emulating native code using an external processor simulator and simulating the memory using a large byte array, we are able to simulate the whole VM execution, allowing to develop and debug it with the Smalltalk development tools and a live programming experience.
 
@@
 
-in this section, we first describe briefly the start-up sequence of \openSmalltalkVM using snapshots.
+In this section, we firstly describe briefly the start-up sequence of \openSmalltalkVM using snapshots, a key aspect of the VM simulator success.
 Then, we detail the simulation of the interpreter-only VM.
@@
 A Smalltalk program is started from a snapshot, \emph{i.e.,} a memory dump of all live objects at a given point in the execution of a program.
-The snapshot includes objects such as the classes, the complied methods in the form of bytecodes and the running processes.
-At start-up, the VM restores the state of all objects in memory and resumes execution in the active process at snapshot time.
+The snapshot includes objects such as the classes, the compiled methods in the form of bytecodes and the running processes.
+At start-up, the VM restores the state of all objects in memory and resumes execution in the process active at snapshot time.
 
-When programming with Smalltalk, the programmers usually start ups from a snapshot which contains the core libraries, the development environment and the application developed.
-Development of the application consists essentially in writing and editing code, which effectively installs, modifies and removes compiled methods and classes from the snapshots.
+When programming with Smalltalk, the programmers usually starts from a snapshot which contains the core libraries, the development environment and the application developed.
+Development of the application consists essentially in writing and editing code, which effectively installs, modifies and removes compiled methods and classes from the set of live objects.
 When the changes are done, the developer takes a new snapshot, which includes its changes.
-Deployment is done from a snapshot containing the deployed application and required core librairies (unused librairies and development tools may be removed from the deployed snapshot).
+Deployment is performed using a snapshot containing the deployed application and the required core librairies (unused librairies and development tools may be removed from the deployed snapshot to decrease memory footprint and for security purposes).
 
-Starting up the VM for source files has never been possible.
+Starting up \openSmalltalkVM from source files has never been possible.
 However, recent work~\cite{PharoBootstrap} allowed one of the Smalltalk runtime to recreate a snapshot from sources, indirectly allowing to start the VM from source files.
@@
 All the C variables are simulated as Smalltalk objects.
-They used specific classes, such as \emph{CArrayAccessor} over normal Smalltalk classes, to emulate the C behavior (only array accesses are available in C, not high level iterator APIs for example).
-One of the most complex component is the Stack.
-The Stack is represented in \openSmalltalkVM as a double linked list of stack pages which are maintained by the VM.
+They used specific classes, such as \emph{CArrayAccessor} over normal Smalltalk classes, to emulate the C behavior (only array accesses are available in C, not high level iterator APIs, etc.).
+An interesting example is the execution stack.
+The stack is represented in \openSmalltalkVM as a double linked list of stack pages which are maintained by the VM.
 Each stack page is represented as a Smalltalk object.
 
-All the Slang code is implemented in multiple Smalltalk classes, to organise the code.
+All the Slang code is implemented in multiple Smalltalk classes, to organise the code and to add flexibility through polymorphism.
+For example, the AbstractCompactor class has two subclasses, one implements a sweep algorithm and the other a compact one.
 For production, at Slang-to-C compilation time, all the code is compiled in a single C file.
+All the flexibility is removed, using the same example, the VM developer chooses at this moment if he wants to compile a VM with a sweep or compact algorithm.
 No polymorphism is available at runtime.
-However, polymorphism can be abused for debugging purposes.
-For example, a class is available holding all the garbage collection compaction logic.
-This class might be subclassed with a simulation only version: this version can use normal Smalltalk code over Slang, allowing to express easily specific constraints which leads to assertion failures if not met, such constraints are convenient for debugging.
+However, since polymorphism is available in the simulator, it can be abused for debugging purposes.
+Still re-using the same example, the AbstractCompactor class has also simulation specific subclasses.
+Such versions typically express additional constraints in the form of assertions which can be written in plain Smalltalk without restrictions to easily express complex constraints.
+They also keep specific values live so they can be accessed at debugging time.
 
@@
 The JIT itself is written in Slang and simulated with the Smalltalk execution model.
-To simulate the machine code, the start of the byte array representing the memory now holds the machine code generated and installed by the JIT.
+To simulate the machine code, the start of the byte array representing the memory holds, when the JIT is enabled, the machine code generated at runtime.
+
 Bindings to processor simulator libraires (Bochs for x86 and x64, Skyeye for ARMv6) were implemented so that the machine code can be simulated.
@@
 Calls from machine code to slang code are implemented by using multiple invalid processor instructions, leading to a trap in the processor simulator.
-This trap is caught is the VM simulator, which then resumes Slang simulation by calling the method corresponding to the invalid processor instruction.
-Calling machine code from Slang requires to start the simulation and then launch an exception to stop Slang simulation.
-The processor simulator can be start in two different ways.
-It can either start simulating code until it meets an invalid instruction or simulate one instruction at a time.
-The second version is slower, but allows to implement specific debugging features, such as conditional breakpoints in-between each machine instruction.
+Such traps can be caught is the VM simulator, which then resumes Slang simulation by calling the correct method based on the invalid processor instruction.
+Calling machine code from Slang requires to start the machine code simulation but also to raise an exception to stop Slang simulation.
+Indeed, in the production VM, the processor can execute the C code or the native code generated by the JIT, but not both at the same time.
+
+Machine code simulation can be performed in two different ways.
+The processor simulator can start simulating code until it meets an invalid instruction.
+That version is convenient because it is the fastest to execute, while still catching errors such as invalid memory accesses.
+Alternatively, it can simulate one instruction at a time.
+This second version is slower, but it allows to implement specific debugging features, such as conditional breakpoints in-between each machine instruction.
 
@@
 The whole start-up takes around 15 seconds on a recent Macbook pro.
-While developing the template JIT, this start-up time may still be too long and move the live programming experience to an edit-compile-run cycle.
+While developing the JIT, this start-up time may still be too long and move the live programming experience to an edit-compile-run cycle, which we want to avoid.
 To work around this problem, we implemented a tool called \emph{In-image compilation}.
-In-image compilation basically allows to call the JIT as a Smalltalk library on a given bytecode compiled method to generate the corresponding machine code and display the disassembly.
-Since the JIT is template-based, in-image compilation is very convenient to develop and optimize each of the JIT templates.
+In-image compilation basically allows to call the JIT as a Smalltalk library on a given bytecode compiled method to generate the corresponding machine code and display the disassembly\footnote{Note that this apply for the baseline JIT, which translate a single bytecode method into machine code, adaptive optimizations and speculative optimizations are debugged differently.}.
+HERE Since the JIT is template-based, in-image compilation is very convenient to develop and optimize each of the JIT templates.
 To generate the machine code, the JIT has to access specific objects (the compiled method, the literals, known objects such as true, false or nil) as if they were in the simulated memory.

### 2018-08-14 paper.tex
kept second pass

@@
 In-image compilation basically allows to call the JIT as a Smalltalk library on a given bytecode compiled method to generate the corresponding machine code and display the disassembly\footnote{Note that this apply for the baseline JIT, which translate a single bytecode method into machine code, adaptive optimizations and speculative optimizations are debugged differently.}.
-HERE Since the JIT is template-based, in-image compilation is very convenient to develop and optimize each of the JIT templates.
+
+The JIT basically generates a sequence of machine instructions for a given pattern of bytecodes.
+We will use the term template to describe the sequence of machine instructions generated for a pattern of bytecodes, even though we do not like to reduce the JIT to a simple template-based JIT since each bytecode generates slightly different machine instructions based on register pressure, a simulated stack and
+a few heuristics.
+Since the JIT is template-based, in-image compilation is very convenient to develop and optimize each of the JIT templates.
 To generate the machine code, the JIT has to access specific objects (the compiled method, the literals, known objects such as true, false or nil) as if they were in the simulated memory.
-To work around this, we built a facade, which is started on the compiled method and only includes the subset of objects required by the JIT to generate the machine code.
+To work around this, we built a facade, which includes mock addresses for all the object the JIT may require to generate the machine code of a given method.
 
 \subsection{Virtual Machine Simulation Limitations}
+
+The simulator has several limitations.
+Due to the simulation infrastructure being different from the actual hardware, code is run differently and one could think it leads to strange bugs happening only in simulation or in production.
+However, we have been using this infrastructure since 1995, and this kind of bugs are really rare and usually easy to fix.
+We have however two main limitations: simulation performance which is quite slow and calls to external C/C++/machine code which cannot be simulated.
 
@@
 With the JIT and processor simulation enabled, without specific debugging options such as conditional breakpoints in between machine instructions, simulation drops to around 500 times slower than the normal VM.
+We usually enabled conditional breakpoints in-between machine instructions only when we reach a point in the simulator where the bug is about to happen since it is way slower, so the overall performance in this context is not really relevant.
+
 This means for example that if a GC bug happens in an application 15 minutes after start-up, it will take 50 hours to reproduce in the interpreter-only simulator.
-Bugs in the jitted code are worst.
+Bugs in the jitted code are worse.
 Fortunately, we work around this problem by using snapshots and the interpreter-only simulator for GC bugs.
@@
 The VM simulator can then be started just before the crash and the debugging tools can be used after only several dozens of seconds.
-If the bug is unrelated to the JIT, the interpreter-only simulator can be used and it is a little bit quicker to execute code.
+If the bug is unrelated to the JIT, the interpreter-only simulator can also be used and it is a little bit quicker to execute code.
 
 \paragraph{Calls to external code.}
-Although most of the GC and JIT development and bug resolution can be done in the simulator, specific developments and bugs cannot.
+Although most of the GC and JIT development and debugging can be done in the simulator, specific tasks cannot be done this way.
 Basically, any calls outside of the machine code generated by the JIT and the Slang code cannot be simulated.
 For specific small parts of the VM, such as file management, we extended the simulator, effectively duplicating the code base with the C code, to support those features in simulation.
-However, there is no solution in the general case: we cannot afford to simulate both the compiled C code and the jitted code on the processor simulator, that would be horribly slow, and specific behaviors in the machine code not present in the code generated by the JIT cannot be simulated (Access to C variables, OS, etc.).
+However, there is no solution in the general case: we cannot afford to simulate both the compiled C code and the jitted code on the processor simulator, that would be horribly slow, and specific behaviors in the machine code not present in the code generated by the JIT can hardly be simulated (Access to C variables, OS variables, etc.).
 
-The main limitation we have is with Foreign Function Interfaces (FFI).
-Most bugs we have with FFI are due to specific interaction between call-backs, low-level assembly FFI specific glue code and moving objects.
-Such bugs cannot be debugged, so far, with our simulation infrastructure and we have to rely on gdb/lldb.
+The main limitation we have right now is with Foreign Function Interfaces (FFI).
+We have a significant amount of bugs in FFI, often due to specific interaction between call-backs, low-level assembly FFI specific glue code and moving objects.
+Such bugs cannot be debugged with our simulation infrastructure so far and we have to rely on gdb/lldb.
 
@@
 
+HERE
 Recently, to evaluate new garbage collection algorithm against standard algorithm, we implemented a Mark-Sweep in addition to the existing Mark-Compact for old space garbage collection.

### 2018-08-14 paper.tex
keep pass

@@
 
-HERE
-Recently, to evaluate new garbage collection algorithm against standard algorithm, we implemented a Mark-Sweep in addition to the existing Mark-Compact for old space garbage collection.
-The whole implementation was done in the simulator, and only when it was working there, we compiled it to C and it was working out of the box.
+Recently, to evaluate new old space garbage collection algorithms we designed against standard algorithms, we implemented a Mark-Sweep in addition to the existing Mark-Compact collector.
+The whole implementation was done in the simulator, and only when it was working there, was it compiled to C.
+Using this process, the compiled C code was kind of working out of the box.
 To describe the implementation process, we need to discuss briefly first the assertion levels in the VM.
@@
 
-\paragraph{Assertions.}To stabilise our code base and find easily production bugs, all the code based is annotated with assertion, ensuring a specific state is as the VM developer would expect it to be, and stopping the program execution if not.
-We have basically three levels of assertions.
-Assertions written in plain Smalltalk are convenient since more complex constraints can be expressed, but they can be performed only in the simulator.
-Assertions written in Slang are also verified in simulation mode, but they are compiled to C.
-The C compiler, based on a compilation flag, then picks if it wants to compile the VM with or without assertions.
-The VM with assertions is used for debugging and to recreate snapshots of runtime juste before they crash.
+\paragraph{Assertions.}To stabilise our code base and find easily production bugs, all the code based is annotated with assertions.
+Each assertion ensures a specific state is as the VM developer would expect it to be and stops or log the incorrect result if not.
+We have a multi-level assertion system.
+Assertions written in plain Smalltalk are convenient since more complex constraints can be expressed easily using high level structures (Sets, etc.), but they can be performed only in the simulator.
+Assertions written in Slang are executed in the simulator but they are also compiled to C.
+The C compiler, based on a compilation flag, chooses to compile the VM with or without assertions.
+The VM with assertions is used for debugging and to recreate snapshots just before a VM crash.
 The VM without assertions is the production VM.
 
+HERE
 \paragraph{Simulation.} Once we had partially written the new Sweep algorithm, we started the simulator.
-Since the algorithm was partially written, we could just write the missing pieces inside the debugger, installing the new code at runtime, as one can do it in Smalltalk.
+Since the algorithm was partially written, we had to write the missing pieces inside the debugger, installing the new code at runtime, as one can do it in Smalltalk.
 In addition, the simulator has an interesting property: each time it performs a GC, either a scavenge or an old space collection, it first copies the simulated memory (\emph{i.e.,} the heap), performs the GC there, and if no assertion fails, it then performs the GC on the original version.

### 2018-08-14 paper.tex
fix;

@@
 
-HERE
 \paragraph{Simulation.} Once we had partially written the new Sweep algorithm, we started the simulator.
-Since the algorithm was partially written, we had to write the missing pieces inside the debugger, installing the new code at runtime, as one can do it in Smalltalk.
+Since the algorithm was partially written, we had to write the missing pieces inside the debugger, installing the new code at runtime, as one can do it in any Smalltalk application.
 In addition, the simulator has an interesting property: each time it performs a GC, either a scavenge or an old space collection, it first copies the simulated memory (\emph{i.e.,} the heap), performs the GC there, and if no assertion fails, it then performs the GC on the original version.
-This means that if the algorithm was not working, an assertion would fail or an error such as incorrect memory access would happen in a copy of the heap.
-The simulator can then reproduce the exact same crash as many times as we want duplicating again and again the same original heap.
-This is very convenient to debug specific GC bugs because often in GC bugs, when the bug happens, the memory is already corrupted and it's quite difficult to track back where the bug comes from without rewinding the memory state.
+This means that if the algorithm was not working, and likely an assertion would fail or an error such as incorrect memory access would be raised, the erratic behavior would have happened in a copy of the heap.
+The simulator can then reproduce the exact same erratic behavior as many times as the VM developer wants duplicating again and again the same original heap.
+This is very convenient to debug specific GC bugs.
+Indeed, often in GC bugs, when the bug happens, the memory is already corrupted and it's quite difficult to track back where the bug comes from without rewinding the memory state, which is usually very difficult or impossible.
 
@@
 
+HERE
 A few years ago, we added support in the VM for read-only objects \cite{ROObject}.

### 2018-08-14 paper.tex
finish pass

@@
 
+MEMORY FIGURE?
+
 All the Slang code is implemented in multiple Smalltalk classes, to organise the code and to add flexibility through polymorphism.
@@
 
-TOWRITE FROM MY YOUTUBE VIDEO.
+TO WRITE FROM MY YOUTUBE VIDEO.
 Back-in-time debugging of machine state
@@
 
-Ref Sista
+Ref Sista ManLang
+\cite{SistaFastStartUp}
+END TODO
 
@@
 
-HERE
 A few years ago, we added support in the VM for read-only objects \cite{ROObject}.
@@
 To maximize the performance, we changed the templates in the JIT compiler for the different memory stores.
-Since the template JIT has to generate very quickly machine code, it goes through the code three times: (1) it scans the bytecode, (2) it generates abstract instructions from the bytecode and (3) it generates the native instructions from the abstract instructions.
+%Since the JIT has to generate very quickly machine code, it goes through the code three times: (1) it scans the bytecode, (2) it generates abstract instructions from the bytecode and (3) it generates the native instructions from the abstract instructions.
 The abstract instructions are mapped almost one to one to machine instructions, their main purpose is to compact the generated code by removing Nops used for jump targets or finding out out the size of jumps.
-
-To optimize the template, we use the in-image compilation framework.
+To optimize each template, we use the in-image compilation framework.
 We selected a method with a single store to make it simple.
-We requested the JIT to generate the machine code and changed the store template until the machine code generated was the exact instructions we wanted.
-It is possible, in in-image compilation, to use the Smalltalk debugger on the JIT code itself to inspect the JIT state and fix the code on -the-fly without any recompilation process.
-Once we went through the few store templates (there are a few different templates for optimizations purposes, for example, storing a constant integer does not require a garbage collector write barrier check),
+We requested the JIT to generate the machine code and changed the template to optimize until the machine code generated was the exact instructions we wanted.
+It is possible, in in-image compilation, to use the Smalltalk debugger on the JIT code itself to inspect the JIT state and fix the JIT code on-the-fly without any major compilation pause.
+Once we went through the few store templates (there are a few different templates for optimizations purposes, for example, storing a constant integer does not require a garbage collector write barrier check), we've just had to evaluate performance and correctness through benchmarks and tests to validate the implementation.
 
@@
 A side-effect of VM simulation, and specifically to be able to interrupt the simulation and introspect the simulated memory and simulation specific objects, is to be able to analyse the runtime with scripts written on-the-fly.
+Indeed, the simulator can be stopped at any given point and arbitrary Smalltalk code can be written and evaluated similarly to the \emph{eval} Javascript construct to analyse the simulated memory, including the machine code zone, the heap or any Smalltalk object representing the VM state.
 
@@
 
-One of the first analysis we run is to stop the simulation when the machine code zone reached 1Mb.
-We then iterated over the machine code zone and investigated what was in.
-As show in Table \ref{tab:generalAnalysis}, 1752 were compiled to machine code by the template JIT, 6352 sends\footnote{We use the Smalltalk terminology, send, to discuss virtual calls since we are talking about Smalltalk.} are present but 2409 of them are not linked (basically, they have never been used).
+One of the first analysis we ran was on the machine code zone.
+We stopped the simulation when the machine code zone reached 1Mb.
+We then iterated over it and investigated what was in.
+As show in Table \ref{tab:generalAnalysis}, 1752 methods were compiled to machine code by the JIT, 6352 sends\footnote{We use the Smalltalk terminology, send, to discuss virtual calls since we are talking about Smalltalk.} are present but 2409 of them are not linked (basically, they have never been used).
 
@@
 
-Further analysis, in Table \ref{tab:polyAnalysis}, confirms Urs H\"{o}lzle analysis \cite{PICSelf}: around 90\% of used send sites are monomorphic, 9\% are polymorphic (up to 6 different cases in our implementation) and the remaining \% is megamorphic.
+Further analysis, in Table \ref{tab:polyAnalysis}, confirms Urs H\"{o}lzle statement \cite{PICSelf}: around 90\% of used send sites are monomorphic, around 9\% are polymorphic (up to 6 different cases in our implementation) and the remaining \% is megamorphic.
 
@@
 
-The code for these analysis are detailed in the Section "Let Me Tell You All About It, Let Me Quantify" of the blog post "Build me a JIT as fast as you can"\footnote{http://www.mirandabanda.org/cogblog/2011/03/01/build-me-a-jit-as-fast-as-you-can/}.
+The code used for these analysis is detailed in the Section "Let Me Tell You All About It, Let Me Quantify" of the blog post "Build me a JIT as fast as you can"\footnote{http://www.mirandabanda.org/cogblog/2011/03/01/build-me-a-jit-as-fast-as-you-can/}.
 
@@
 
-The results of the analysis are used to direct performance design decisions on the VM. In this section we describe how the analysis impacted a design called "Early Polymorphic inline cache promotion".
+The results of the analysis are sometimes used to direct performance design decisions on the VM.
+In this section we describe how the analysis impacted a design called "Early polymorphic inline cache promotion".
 
-We designed the polymorphic inline caches (PICs) with two implementations:
+We designed the polymorphic inline caches (PICs) \cite{PICSelf} with two implementations:
 \begin{itemize}
 	\item \emph{Closed PICs:} Such caches can deal with up to 6 cases, and are basically implemented as a jump table.
-	\item \emph{Open PICs:} Such caches can deal with any number of cases, they consist of three probes searching the global look-up cache, a hash map shared with the interpreter, and fall back into a standard look-up routine if there is a cache miss.
+	\item \emph{Open PICs:} Such caches can deal with any number of cases, they consist of three probes searching the global look-up cache (a hash map shared with the interpreter) and fall back into a standard look-up routine if nothing is found after three attempts.
 \end{itemize}
@@
 
-Many VM developers developed different tools to help them being more efficient, but they rarely publish about it. We focus in this section on two related work.
+Many VM developers implemented different tools to help them working more efficiently on their VM, but they rarely publish about it. We focus in this section on two related work.
 
 \paragraph{Maxine Inspectors.} The Maxine inspectors \cite{MaxineInspector} were demonstrated at OOPSLA'18.
-They allow to inspect the running state of the VM while it runs for debugging purposes.
-One of the main difference with out design is that the Maxine VM is metacircular, they do not have a simulation and a production mode as we do.
-We believe having two different modes allows us to easily generate a production artifact while still having nice debugging features.
+They allow to inspect the running state of the Maxine VM while it runs for debugging purposes.
+One of the main difference with out design is that the Maxine VM is metacircular, hence it does not have a simulation and a production mode as we do but a single production debuggable mode.
+We believe having two different modes allows us to easily generate a production artefact while still having nice debugging features.
 Having a full metacircular VM would be interesting.
-However, so far, most VMs used in production (Java, Javascript, etc.), even after the huge recent investments in the Javascript VMs by the four major web vendors, are still compiling through the C/C++ compiler and not metacircular.
+However, so far, most VMs used in production (Java, Javascript, etc.), even after the huge recent investments in the Javascript VMs by the four major web vendors, are still compiling through the C/C++ compiler and are not metacircular.
 Hence, although a metacircular VM has interesting advantages, it is not clear it is that convenient to build a VM in such a way.
@@
 Most of the VM code is written in RPython, a restricted Python, instead of Slang, and some leftovers are written in plain C.
-The main difference is that RPython is much closer to Python than Slang is to Smalltalk, RPython allows higher level structures such as dictionaries to be used.
+The main difference is that RPython is much closer to Python than Slang is to Smalltalk.
+RPython allows higher level structures such as dictionaries to be used.
 The design decision comes with its set of advantages and drawbacks.
-The key advantage is that the RPython code feels like Python code and is relatively quite easy to read write.
+The key advantage is that the RPython code feels like Python code and is relatively quite easy to read write, unlike Slang which feels like C and is as easy to write as C.
 The main drawback is that RPython to C compilation takes way longer than the Slang to C compilation (up to 40 minutes in a recent Macbook pro for the RSqueak VM \cite{RSqueak}, instead of several seconds for Slang).
 
-Although the RPython code can be executed as normal Python code, for some reasons, the developers seem to think it is not worth to do it, mostly because executing code in such a way is very slow.
-The overall architecture of the RPython toolchain is different, which leads to a longer time to reach peak performance (though their peak performance is at least in theory better than with a template JIT as \openSmalltalkVM features).
-This time may be very significant in simulation mode.
+Although the RPython code can be executed as normal Python code, for some reasons, the developers seem to think it is not worth to do such a thing, mostly because executing code in this way is very slow.
+The overall architecture of the RPython toolchain is different, which leads to a longer time to reach peak performance (though their peak performance is at least in theory better than with our JIT).
+This time may be quite significant in simulation mode.
 In addition, RPython was originally designed for Python, which does not feature snapshot by default, so they cannot abuse snapshots to work around the simulation slow performance.

### 2018-08-15 paper.tex
added 4.1

@@
 
-TO WRITE FROM MY YOUTUBE VIDEO.
-Back-in-time debugging of machine state
-Conditional stepping.
+One idea we implemented in the recent year was to support a more aggressive JIT with speculative optimizations through a bytecode to bytecode optimizer \cite{SistaFastStartUp}.
+To be able to generate efficient bytecodes, we had to introduce new unsafe bytecodes allowing, for example, accesses in arrays without any checks (type and bounds checks) \cite{SistaBytecodeSet}.
+For each new bytecode, we introduced new templates in the JIT to generate efficient machine code for the optimized methods.
+After all the basic unit tests worked, we ran the speculative optimizer and got a crash.
+We could figure out which bytecode method was triggering the crash, but we had no idea from which template it came from, in addition, optimized methods include many inlined methods and are very large, so it's difficult to figure out where the issue comes from looking at thousands of bytecodes.
 
-Ref Sista ManLang
-\cite{SistaFastStartUp}
-END TODO
+We created a snapshot where the faulty method was executed right after start-up.
+We started the VM simulator, and set-up a conditional breakpoint so that simulator would stop when the JIT would generate that method to machine code.
+Then, when the simulator effectively stopped, we changed the conditional breakpoint to stop execution where the address corresponding to the faulty method entry in machine code would be used, either by a call from the interpreter or through inline cache relinking.
+The simulator stopped again, about to execute the machine code corresponding to the faulty method.
+We then cloned the simulator to be able to reproduce the crash again and again.
+
+Executing the faulty method led to an assertion failure.
+However, that assertion failed in a store check, telling us that the object to store into looked suspicious (outside of the heap).
+It happens that this object was read from a field on stack, and that this field held an incorrect address.
+We could not tell anymore at this point in the execution what instruction wrote on stack the invalid address.
+So we discarded the cloned simulation, and cloned a fresh simulator again, just before the execution of the faulty method in machine code.
+This time, we enabled single stepping (\emph{i.e.,} the processor simulator simulates one instruction at a time) and we added a breakpoint stopping execution when the specific field on stack would be written to an incorrect value.
+In this case, the breakpoint was checked in between each machine instruction, and execution stopped right after the machine instruction which wrote the incorrect value on stack.
+From the machine instruction address, we could map it back to the bytecode template (it mapped to the new bytecode template for inlined allocations), built a simpler method crashing the runtime and fixed the template using in-image compilation (See next subsection).
+
+The debugging process used here is exampled in a youtube video \footnote{https://www.youtube.com/watch?v=hctMBGAXVSs}.
 

### 2018-08-15 paper.tex
added fig

@@
 		\centering
-		\includegraphics[width=0.6\linewidth]{figures/VMCompilation}
+		\includegraphics[width=0.55\linewidth]{figures/VMCompilation}
 		\caption{Cog VM compilation}
@@
 
-The heap is simulated as a large contiguous byte array.
+Figure \ref{fig:SimulatedMemory} describes, from a high-level perspective, the memory used by \openSmalltalkVM at runtime and the simulated counter-part.
+Let's detail briefly the memory used by the production VM, on the top of the figure.
+On the left side, which is usually the low addresses, we can find Text, the memory section holding the native code of the VM (the interpreter, the memory manager and the JIT itself, but not the code compiled by the JIT).
+Then, Data holds initialized and uninitialized data, mainly the C variables used globally in the VM.
+On higher addresses, we can find the beginning of the memory managed by the VM.
+At start-up, the VM mmaps a memory region used for the executable code generated by the JIT (optional executable section, exists only if the JIT is enabled), for the young objects and for the old objects present in the snapshot started plus a little more space for the first tenures\footnote{A tenure is the process of promoting young objects to old objects in a generational GC.} to be performed without triggering the full garbage collector.
+Later during execution, as old space grows, new memory regions are mmap'ed on higher addresses to store other old objects.
+Usually at very high addresses, we can find the C stack.
+Room in the C stack is alloca'ed at VM start-up to hold the Smalltalk stack.
+Both stacks are disjointed and managed differently.
+
+In the simulator, the heap is simulated as a large contiguous byte array.
 References between objects are effectively indexes inside the byte array instead of pointers.
-All the C variables are simulated as Smalltalk objects.
+All the Slang variables, normally translated to C variabled, are simulated as Smalltalk objects.
 They used specific classes, such as \emph{CArrayAccessor} over normal Smalltalk classes, to emulate the C behavior (only array accesses are available in C, not high level iterator APIs, etc.).
-An interesting example is the execution stack.
-The stack is represented in \openSmalltalkVM as a double linked list of stack pages which are maintained by the VM.
-Each stack page is represented as a Smalltalk object.
+The Slang code is executed as Smalltalk code.
+The Smalltalk stack is represented in \openSmalltalkVM as a double linked list of stack pages which are maintained by the VM.
+Each stack page is represented as a Smalltalk object in the simulator.
 
-MEMORY FIGURE?
+\begin{figure*}[bth!]
+		\centering
+		\includegraphics[width=0.9\linewidth]{figures/SimulatedMemory}
+		\caption{Runtime Memory and Simulated Counter-Part}
+		\label{fig:SimulatedMemory}
+\end{figure*}
 
@@
 \subsection{In-image compilation}
+
+IN-IMAGE COMP FIG?
 
@@
 
-The debugging process used here is exampled in a youtube video \footnote{https://www.youtube.com/watch?v=hctMBGAXVSs}.
+The debugging process used here is exampled in a youtube video\footnote{https://www.youtube.com/watch?v=hctMBGAXVSs}.
 

### 2018-08-15 paper.tex
Added more figures

@@
 
+\begin{figure*}[ht!]
+		\centering
+		\includegraphics[width=0.9\linewidth]{figures/SimulatedMemory}
+		\caption{Runtime Memory and Simulated Counter-Part}
+		\label{fig:SimulatedMemory}
+\end{figure*}
+
 Figure \ref{fig:SimulatedMemory} describes, from a high-level perspective, the memory used by \openSmalltalkVM at runtime and the simulated counter-part.
@@
 Each stack page is represented as a Smalltalk object in the simulator.
-
-\begin{figure*}[bth!]
-		\centering
-		\includegraphics[width=0.9\linewidth]{figures/SimulatedMemory}
-		\caption{Runtime Memory and Simulated Counter-Part}
-		\label{fig:SimulatedMemory}
-\end{figure*}
 
@@
 
-Bindings to processor simulator libraires (Bochs for x86 and x64, Skyeye for ARMv6) were implemented so that the machine code can be simulated.
+Bindings to processor simulator libraries (Bochs for x86 and x64, Skyeye for ARMv6) were implemented so that the machine code can be simulated.
 Calls in-between slang code and direct machine code are a little bit trickier to simulate.
@@
 
-IN-IMAGE COMP FIG?
-
 Simulating the whole VM requires going through the whole start-up sequence: loading the snapshot, running code registered in the start-up sequence and resuming the user interface.
@@
 While developing the JIT, this start-up time may still be too long and move the live programming experience to an edit-compile-run cycle, which we want to avoid.
+
 To work around this problem, we implemented a tool called \emph{In-image compilation}.
-In-image compilation basically allows to call the JIT as a Smalltalk library on a given bytecode compiled method to generate the corresponding machine code and display the disassembly\footnote{Note that this apply for the baseline JIT, which translate a single bytecode method into machine code, adaptive optimizations and speculative optimizations are debugged differently.}.
-
-The JIT basically generates a sequence of machine instructions for a given pattern of bytecodes.
-We will use the term template to describe the sequence of machine instructions generated for a pattern of bytecodes, even though we do not like to reduce the JIT to a simple template-based JIT since each bytecode generates slightly different machine instructions based on register pressure, a simulated stack and
-a few heuristics.
-Since the JIT is template-based, in-image compilation is very convenient to develop and optimize each of the JIT templates.
+In-image compilation basically allows to call the JIT as a Smalltalk library on a given bytecode compiled method to generate the corresponding machine code, calls the bound processor simulator to disassemble the code and decorate the disassembly with Smalltalk specific information.
 To generate the machine code, the JIT has to access specific objects (the compiled method, the literals, known objects such as true, false or nil) as if they were in the simulated memory.
 To work around this, we built a facade, which includes mock addresses for all the object the JIT may require to generate the machine code of a given method.
+Figure \ref{fig:InImageCompiler} summarizes the process.
+Note that this technique applies for the baseline JIT, which translate a single bytecode method into machine code, adaptive optimizations and speculative optimizations are debugged differently.
+
+\begin{figure}[ht!]
+		\centering
+		\includegraphics[width=\linewidth]{figures/InImageCompiler}
+		\caption{In-image compilation}
+		\label{fig:InImageCompiler}
+\end{figure}
+
+\paragraph{Templates.} The JIT generates a sequence of machine instructions for a given pattern of bytecodes.
+We will use the term template to describe the sequence of machine instructions generated for a pattern of bytecodes, even though we do not like to reduce the JIT to a simple template-based JIT since each bytecode generates slightly different machine instructions based on register pressure, a simulated stack and
+a few heuristics. Since the JIT is template-based, in-image compilation is very convenient to develop and optimize each of the JIT templates.
 

### 2018-08-15 paper.tex
made a pass over 4.1

@@
 
-One idea we implemented in the recent year was to support a more aggressive JIT with speculative optimizations through a bytecode to bytecode optimizer \cite{SistaFastStartUp}.
-To be able to generate efficient bytecodes, we had to introduce new unsafe bytecodes allowing, for example, accesses in arrays without any checks (type and bounds checks) \cite{SistaBytecodeSet}.
-For each new bytecode, we introduced new templates in the JIT to generate efficient machine code for the optimized methods.
-After all the basic unit tests worked, we ran the speculative optimizer and got a crash.
-We could figure out which bytecode method was triggering the crash, but we had no idea from which template it came from, in addition, optimized methods include many inlined methods and are very large, so it's difficult to figure out where the issue comes from looking at thousands of bytecodes.
+In the recent years, we added support for a more aggressive JIT with speculative optimizations through a bytecode to bytecode optimizer \cite{SistaFastStartUp}, re-using the existing template JIT as a back-end.
+To be able to generate efficient bytecodes in the bytecode to bytecode optimizer, we had to introduce new unsafe bytecodes allowing, for example, accesses in arrays without any checks (type and bounds checks) \cite{SistaBytecodeSet}.
+For each new bytecode, we introduced new templates in the existing JIT to generate efficient machine code for the optimized methods using the new bytecodes.
+Once all the basic unit tests worked, we ran the VM with the speculative optimizer, which executed optimized code, and got a crash.
+We could figure out which bytecode method was triggering the crash, but we had no idea from which template the crash came from.
+In addition, optimized methods include many inlined methods, making them very large, so it was difficult to figure out where the issue comes from just by looking at thousands of bytecodes.
 
-We created a snapshot where the faulty method was executed right after start-up.
+To understand the crash, we created a snapshot where the faulty method was executed right after start-up.
 We started the VM simulator, and set-up a conditional breakpoint so that simulator would stop when the JIT would generate that method to machine code.
-Then, when the simulator effectively stopped, we changed the conditional breakpoint to stop execution where the address corresponding to the faulty method entry in machine code would be used, either by a call from the interpreter or through inline cache relinking.
+Then, when the simulator effectively stopped, we changed the conditional breakpoint to stop execution when the address corresponding to the faulty method entry in machine code would be used, either by a call from the interpreter or through inline cache relinking.
 The simulator stopped again, about to execute the machine code corresponding to the faulty method.
@@
 Executing the faulty method led to an assertion failure.
-However, that assertion failed in a store check, telling us that the object to store into looked suspicious (outside of the heap).
+However, that assertion failed in a GC store check, telling us that the object to store into looked suspicious (address outside of the heap).
 It happens that this object was read from a field on stack, and that this field held an incorrect address.
-We could not tell anymore at this point in the execution what instruction wrote on stack the invalid address.
-So we discarded the cloned simulation, and cloned a fresh simulator again, just before the execution of the faulty method in machine code.
-This time, we enabled single stepping (\emph{i.e.,} the processor simulator simulates one instruction at a time) and we added a breakpoint stopping execution when the specific field on stack would be written to an incorrect value.
-In this case, the breakpoint was checked in between each machine instruction, and execution stopped right after the machine instruction which wrote the incorrect value on stack.
-From the machine instruction address, we could map it back to the bytecode template (it mapped to the new bytecode template for inlined allocations), built a simpler method crashing the runtime and fixed the template using in-image compilation (See next subsection).
+We could not tell anymore at this point in the execution what instruction among the thousands previous ones wrote on stack the invalid address.
+So we discarded the cloned simulation, and cloned a fresh simulator again, just before the execution of the faulty method in machine code to reproduce again the crash.
+This time, we enabled single stepping (\emph{i.e.,} the processor simulator simulates one instruction at a time) and we added a breakpoint stopping execution when the specific field on stack would be written to the incorrect value found before.
+In this case, the conditional breakpoint is checked in between each machine instruction, and execution stopped right after the machine instruction which wrote the incorrect value on stack.
+From the machine instruction address, we could figure out which bytecode pattern generated the incorrect machine code (it was the new bytecode template for inlined allocations).
+From there, we built a simpler method crashing the runtime and fixed the template using in-image compilation (See next subsection).
 
-The debugging process used here is exampled in a youtube video\footnote{https://www.youtube.com/watch?v=hctMBGAXVSs}.
+The debugging process discussed here is exampled in a youtube video\footnote{https://www.youtube.com/watch?v=hctMBGAXVSs}.
 

### 2018-08-15 paper.tex
write library correctly damn it

@@
 When the changes are done, the developer takes a new snapshot, which includes its changes.
-Deployment is performed using a snapshot containing the deployed application and the required core librairies (unused librairies and development tools may be removed from the deployed snapshot to decrease memory footprint and for security purposes).
+Deployment is performed using a snapshot containing the deployed application and the required core libraries (unused libraries and development tools may be removed from the deployed snapshot to decrease memory footprint and for security purposes).
 

### 2018-08-16 paper.tex
Rename to Two Decades Of Smalltralk Development Open-Smalltalk-VM => OpenSmalltalk-VM Add .gitignore; rm --cached the generated outputs, except the finished .pdf
Work on history section, snapshots and development.  Adding references for DeutschSchiffman, Spur, etc.

@@
 
-\def\openSmalltalkVM{Open-Smalltalk-VM\xspace}
+\def\openSmalltalkVM{OpenSmalltalk-VM\xspace}
 %%%
@@
 %% Title information
-\title[Over Twenty Years of Virtual Machine Development Through Simulation]{Over Twenty Years of Virtual Machine Development and Debugging Through Simulation}
+\title[Two Decades of Smalltalk VM Development]{Two Decades of Smalltalk VM Development}
 
@@
 
-\openSmalltalkVM was originally written in Smalltalk to specify and example the virtual machine (VM) implementation.
-The same code base was changed a little bit to allow Smalltalk-to-C compilation to generate the production VM through the C compiler.
+\openSmalltalkVM is a VM for languages in the Smalltalk family which is itself written in a subset of Smalltalk that can easily be translated to C.
+Development is done in Smalltalk, an activity we call "Simulation", and the support code for which is "The Simulator".
+The production VM is derived by translating the core VM code to C, combining the generated C code with a set of platform-specific support files, and compiling with the platform's C compiler.
 Two execution models are effectively available, simulation, where the Smalltalk code is executed on top of a Smalltalk VM, and production, where the same code is compiled to executable code through the C compiler.
@@
 
-As the VM evolved, by introducing better garbage collector algorithms or a just-in-time compiler, both execution models co-evolved.
+As the VM has evolved, including adding a just-in-time compiler, better garbage collector algorithms, and better object representations, both execution models co-evolved.
 For example, the simulation infrastructure was extended with a processor simulator to simulate the code generated by the just-in-time compiler.
@@
 We discuss specifically how we use the VM simulator to develop and debug two core VM components, the garbage collector and the just-in-time compiler.
-Then, we discuss how we use the simulation infrastructure to perform analysis on the runtime, directing some design decisions we make to tune the VM performance.
+Then, we discuss how we use the simulation infrastructure to perform analysis on the runtime, directing some design decisions we have made to tune VM performance.
 
@@
 
-To specify the language and explain how to write the virtual machine (VM) for it, the Smalltalk-80 crew wrote a Smalltalk VM entirely in Smalltalk \cite{blueBook}.
-In 1995, the same team built an open-source Smalltalk and VM, Squeak \cite{SqueakByExample}.
-The VM was ported from the original specifications.
-Part of the code base was however narrowed down to a subset of Smalltalk, called \emph{Slang}, to allow Smalltalk to C compilation \cite{BackToTheFuture} and generate an efficient VM through the C compiler.
+To specify the the virtual machine (VM), the Smalltalk-80 team at Xerox PARC wrote a Smalltalk VM entirely in Smalltalk \cite{blueBook}.
+In 1995, members of the same team built an open-source Smalltalk and VM, Squeak \cite{SqueakByExample} using the code from \cite{blueBook} as a starting point.
+Part of the code base was however narrowed down to a subset of Smalltalk, called \emph{Slang}, to allow Smalltalk to C compilation \cite{BackToTheFuture}.
+Dummy Smalltalk message sends were used to embed meta information in the code, such as \emph{self var: 'foo' type: 'char *'} which have no effect during simulation but guide the translation process.
 Effectively, the VM could be both simulated by executing the Slang code on top of the Smalltalk VM or compiled through the C compiler to native code to generate the production VM.
@@
 \item an interpreter with a spaghetti stack, 
-\item a memory manager with a generational garbage collector (GC) and
-\item BitBlt, an extension for the user interface bit-based engine.
+\item a memory manager with a compact but complex object representation, a pointer-reversing tracing garbage collector and a heap divided into two generations
+\item WarpBlt, a rotation and scaling extension for the bit-based BitBLT graphics engine.
+\item external C code to support running the VM on popular platforms
 \end{itemize} 
-These three parts of the VM were written entirely in Slang. A few extra features, such as file management, were written both in Smalltalk for simulation purposes and in C for the production VM.
+The first three parts of the VM were written entirely in Slang. A few extra features, such as file management, were written both in Smalltalk for simulation purposes and in C for the production VM.
 
-Over the years, the Squeak VM evolved to give birth recently to \openSmalltalkVM\footnote{https://github.com/OpenSmalltalk/opensmalltalk-vm/}, the default VM for different Smalltalk and Smalltalk-like runtimes such as Pharo \cite{PharoByExample}, Squeak \cite{SqueakByExample}, Cuis, Croquet and NewSpeak \cite{NewspeakOopsla}.
+Over the years, the Squeak VM evolved to give birth recently to \openSmalltalkVM\footnote{https://github.com/OpenSmalltalk/opensmalltalk-vm/}, the default VM for various Smalltalk and Smalltalk-like systems such as Pharo \cite{PharoByExample}, Squeak \cite{SqueakByExample}, Cuis, Croquet and Newspeak \cite{NewspeakOopsla}.
 As the VM evolved, the simulator co-evolved as a tool to develop and debug the VM.
+One change moved the meta data that guides translation out of the executable Smalltalk into pragmas \cite{Pragmas}.
 The most significant evolution of the simulator came with the introduction of the Just-In-Time compiler (JIT).
-The existing simulator was not able to simulate the machine code generated by the JIT and had to be extended by binding multiple processor simulators (Bochs for x86 \& x64, SkyEye for ARMv6).
+The code base was liberally sprinkled with assertions.
+The spaghetti stack was mapped to a more conventional stack frame organisation in a stack zone of a few hundred k bytes organised into small pages, a scheme called context-to-stack mapping \cite{EfficientST80}, \cite{StackMappingVW}.
+Machine code generated by the JIT was executed by binding multiple processor simulators (first Bochs \cite{Bochs} for x86, SkyEye for ARMv6, Smalltalk code for 32-bit MIPS).
+A new object representation and garbage collector was added to also support 64-bits \cite{Spur}, which required refactoring the interpreter to allow the object representation to be chosen at startup.
+Bochs was used a second time for x64 support.
 
@@
 The Core VM code (the interpreter, the JIT and the memory manager), is written in Slang.
-The platform code, \emph{i.e.,} Operating System dependent code such as file management or I/O is written directly in C.
+The platform code, \emph{i.e.,} Operating System dependent code such as file management or I/O is written directly in C, with a corresponding Smalltalk version for simulation.
 %EXPLAIN scientifically in this part - no history. + FIGURE, basic with design.
@@
 The main reason why we keep using Slang over C/C++ is VM simulation.
-By interpreting the Slang code as Smalltalk code, emulating native code using an external processor simulator and simulating the memory using a large byte array, we are able to simulate the whole VM execution, allowing to develop and debug it with the Smalltalk development tools and a live programming experience.
+By interpreting the Slang code as Smalltalk code, emulating native code using an external processor simulator and simulating the memory using a large byte array, we are able to simulate the whole VM execution, allowing development and debugging with the Smalltalk development tools and a live programming experience.
 
@@
 \paragraph{Start-up from snapshots.} 
-Smalltalk is a snapshot-based language.
-A Smalltalk program is started from a snapshot, \emph{i.e.,} a memory dump of all live objects at a given point in the execution of a program.
+Smalltalk is an object \emph{system}, rather than a language.
+The entire system, including its development tools and application code is stored in a snapshot file, which is essentially a memory dump of the entire heap.
+A Smalltalk system is started by loading a snapshot file into a VM.
 The snapshot includes objects such as the classes, the compiled methods in the form of bytecodes and the running processes.
-At start-up, the VM restores the state of all objects in memory and resumes execution in the process active at snapshot time.
+At start-up, the VM restores the state of all objects in memory and resumes execution in the process active when the snapshot was created.
+A new snapshot can be made during or at the end of the session.
 
-When programming with Smalltalk, the programmers usually starts from a snapshot which contains the core libraries, the development environment and the application developed.
-Development of the application consists essentially in writing and editing code, which effectively installs, modifies and removes compiled methods and classes from the set of live objects.
-When the changes are done, the developer takes a new snapshot, which includes its changes.
+When programming with Smalltalk, the programmer usually starts from a snapshot which contains the core libraries, the development environment and the application under development.
+Development of the application consists essentially in writing and editing code, which effectively installs, modifies and removes classes and compiled methods from the set of live objects.
+Programming may be done live, as the application under development is running, for example creating or revising methods in the debugger, and continuing.
+Objects may have their shape changed on the fly as instance variables are added and removed.
+When the programmer chooses, the developer takes a new snapshot, which includes its changes, and may include application instances under development [MAYBE THIS SHOULD BE DESCRIBED BELOW IN SNAPSHOTS].
 Deployment is performed using a snapshot containing the deployed application and the required core libraries (unused libraries and development tools may be removed from the deployed snapshot to decrease memory footprint and for security purposes).
+\footnote{Code is organised in packages which are saved to an external version control system, in our case Monticello citation?
+wiresong.ca}
 
-Starting up \openSmalltalkVM from source files has never been possible.
-However, recent work~\cite{PharoBootstrap} allowed one of the Smalltalk runtime to recreate a snapshot from sources, indirectly allowing to start the VM from source files.
+Starting up \openSmalltalkVM from source files has never been supported.
+However, recent work~\cite{PharoBootstrap} allows a snapshot to be created from sources, indirectly allowing the VM to be started from source files.
+The bootstrap process uses the simulator to create and manipulate the heap and then to save the resulting snapshot.
 
@@
 As the VM can be compiled with and without the JIT, the VM can also be simulated with or without the JIT.
+Since executing machine code on the processor simulations is not fast, interpreter simulation may be faster, and startup time and memory footprint are smaller.
 

### 2018-08-16 paper.tex
changes on abstract. Left some comments for Clement

@@
 \newcommand{\sd}[1]{\color{red}\fbox{\bfseries\sffamily\scriptsize Stef:}{\sf\small$\blacktriangleright$\textit{#1}$\blacktriangleleft$}\color{black}}
-\newcommand{\sk}[1]{\color{blue}\fbox{\bfseries\sffamily\scriptsize Sophie:}{\sf\small$\blacktriangleright$\textit{#1}$\blacktriangleleft$}\color{black}}
+\newcommand{\egb}[1]{\color{blue}\fbox{\bfseries\sffamily\scriptsize Elisa:}{\sf\small$\blacktriangleright$\textit{#1}$\blacktriangleleft$}\color{black}}
 \newcommand{\cba}[1]{\color{purple}\fbox{\bfseries\sffamily\scriptsize Clement:}{\sf\small$\blacktriangleright$\textit{#1}$\blacktriangleleft$}\color{black}}
@@
 }
-\email{clement.bera@vub.ac.be}          %% \email is recommended
+\email{clement.bera@vub.be}          %% \email is recommended
 
@@
 }
-\email{egonzale@vub.ac.be}          %% \email is recommended
+\email{egonzale@vub.be}          %% \email is recommended
 
@@
 
+\egb{The title is a bit misleading.
+It sounds great for a journal paper on the \openSmalltalkVM, but probably it is best to adapt it or add a subtitle on the specific topic you touch upon in this paper}
+
+\egb{I rewrote and shortened the first paragraph since the abstract was quite unbalance towards the research context but very scarce on details about the research problem and solution.
+I left the original text in comments below}
+
 \openSmalltalkVM is a VM for languages in the Smalltalk family which is itself written in a subset of Smalltalk that can easily be translated to C.
+Development is done in Smalltalk, an activity we call ``Simulation''.
+The production VM is derived by translating the core VM code to C.
+As a result, two execution models coexist: simulation, where the Smalltalk code is executed on top of a Smalltalk VM, and production, where the same code is compiled to executable code through the C compiler.
+While simulation is used to develop and debug the VM, production is used to release the VM.
+
+%\openSmalltalkVM is a VM for languages in the Smalltalk family which is itself written in a subset of Smalltalk that can easily be translated to C.
 Development is done in Smalltalk, an activity we call "Simulation", and the support code for which is "The Simulator".
@@
 
-As the VM has evolved, including adding a just-in-time compiler, better garbage collector algorithms, and better object representations, both execution models co-evolved.
+\egb{I think you can skip the next paragraph from the abstract.
+If the point is that you need to co-evolve both execution models once the VM changes, you can do that with 1 line in the previous paragraph.
+Now it may look like you will discuss co-evolution issues, but I guess you do not}
+
+As the VM has evolved, including the addition of a just-in-time compiler, better garbage collector algorithms, and better object representations, both execution models co-evolved.
 For example, the simulation infrastructure was extended with a processor simulator to simulate the code generated by the just-in-time compiler.
 
+%TODO egb: revisit this paragraph after reading all paper.
 In this paper, we detail the VM simulation infrastructure and we report our experience developing and debugging the VM within it.
@@
 %% comma separated list
-\keywords{Just-in-Time compiler, Virtual machine, Managed runtime, Tools}  %% \keywords are mandatory in final camera-ready submission
+\keywords{Just-in-Time compiler, virtual machine, managed runtime, tools}  %% \keywords are mandatory in final camera-ready submission
 

### 2018-08-16 paper.tex
changes on introduction

@@
 \egb{I rewrote and shortened the first paragraph since the abstract was quite unbalance towards the research context but very scarce on details about the research problem and solution.
-I left the original text in comments below}
+I left the original text in comments below, and I moved some details to the introduction were they were more necessary in my opinion.}
 
-\openSmalltalkVM is a VM for languages in the Smalltalk family which is itself written in a subset of Smalltalk that can easily be translated to C.
+\openSmalltalkVM is a virtual machine (VM) for languages in the Smalltalk family (e.g.
+Squeak, Pharo) which is itself written in a subset of Smalltalk that can easily be translated to C.
 Development is done in Smalltalk, an activity we call ``Simulation''.
@@
 As a result, two execution models coexist: simulation, where the Smalltalk code is executed on top of a Smalltalk VM, and production, where the same code is compiled to executable code through the C compiler.
-While simulation is used to develop and debug the VM, production is used to release the VM.
 
@@
 
-%TODO egb: revisit this paragraph after reading all paper.
+%TODO egb: revisit this paragraph after reading all paper to refine problem statement and solution.
 In this paper, we detail the VM simulation infrastructure and we report our experience developing and debugging the VM within it.
@@
 
-To specify the the virtual machine (VM), the Smalltalk-80 team at Xerox PARC wrote a Smalltalk VM entirely in Smalltalk \cite{blueBook}.
-In 1995, members of the same team built an open-source Smalltalk and VM, Squeak \cite{SqueakByExample} using the code from \cite{blueBook} as a starting point.
-Part of the code base was however narrowed down to a subset of Smalltalk, called \emph{Slang}, to allow Smalltalk to C compilation \cite{BackToTheFuture}.
-Dummy Smalltalk message sends were used to embed meta information in the code, such as \emph{self var: 'foo' type: 'char *'} which have no effect during simulation but guide the translation process.
+To specify the virtual machine (VM), the Smalltalk-80 team at Xerox PARC wrote a Smalltalk VM entirely in Smalltalk \cite{blueBook}.
+In 1995, members of the same team built an open-source Smalltalk \egb{dialect?} and VM, Squeak \cite{SqueakByExample}, using the code from \cite{blueBook} as a starting point.
+Part of the code base was, however, narrowed down to a subset of Smalltalk, called \emph{Slang}, to allow Smalltalk to C compilation \cite{BackToTheFuture}.
+
+\egb{Rewrote the paragraph to add old info from the abstract and add some flow to the explanation.
+Maybe we should also define here what simulation and production is (since the abstract is not considered to be part of the main body of the paper)}
+
 Effectively, the VM could be both simulated by executing the Slang code on top of the Smalltalk VM or compiled through the C compiler to native code to generate the production VM. 
+The production VM can be derived by translating the core VM code to C, combining the generated C code with a set of platform-specific support files, and compiling with the platform's C compiler.
+Dummy Smalltalk message sends were used to embed meta information in the code, such as \emph{self var: 'foo' type: 'char *'} which have no effect during simulation but guide the translation process of the executable Smalltalk.
+
 At this point, the VM consisted mainly in:
@@
 \item an interpreter with a spaghetti stack, 
-\item a memory manager with a compact but complex object representation, a pointer-reversing tracing garbage collector and a heap divided into two generations
-\item WarpBlt, a rotation and scaling extension for the bit-based BitBLT graphics engine.
-\item external C code to support running the VM on popular platforms
+\item a memory manager with a compact but complex object representation: a pointer-reversing tracing garbage collector and a heap divided into two generations,
+\item WarpBlt, a rotation and scaling extension for the bit-based BitBLT graphics engine,
+\item and external C code to support running the VM on popular platforms.
 \end{itemize}
-The first three parts of the VM were written entirely in Slang. A few extra features, such as file management, were written both in Smalltalk for simulation purposes and in C for the production VM.
 
-Over the years, the Squeak VM evolved to give birth recently to \openSmalltalkVM\footnote{https://github.com/OpenSmalltalk/opensmalltalk-vm/}, the default VM for various Smalltalk and Smalltalk-like systems such as Pharo \cite{PharoByExample}, Squeak \cite{SqueakByExample}, Cuis, Croquet and Newspeak \cite{NewspeakOopsla}.
-As the VM evolved, the simulator co-evolved as a tool to develop and debug the VM.
-One change moved the meta data that guides translation out of the executable Smalltalk into pragmas \cite{Pragmas}.
-The most significant evolution of the simulator came with the introduction of the Just-In-Time compiler (JIT).
-The code base was liberally sprinkled with assertions.
+The first three components of the VM were written entirely in Slang.
+A few extra features, such as file management, were written both in Smalltalk for simulation purposes and in C for the production VM.
+
+Over the years, the Squeak VM evolved to give birth recently to \openSmalltalkVM\footnote{https://github.com/OpenSmalltalk/opensmalltalk-vm/}, the default VM for various Smalltalk-like systems such as Pharo \cite{PharoByExample}, Squeak \cite{SqueakByExample}, Cuis, Croquet and Newspeak \cite{NewspeakOopsla}.
+
+As the VM evolved, the original simulator co-evolved as a tool to develop and debug the VM. The meta data used to guide the translation process was replaced by \emph{pragmas} \cite{Pragmas}.  
+%One change moved the meta data that guides translation out of the executable Smalltalk into pragmas \cite{Pragmas}.  
+However, the most significant evolution of the simulator came with the introduction of the Just-In-Time compiler (JIT). 
+%The code base was liberally sprinkled with assertions. 
 The spaghetti stack was mapped to a more conventional stack frame organisation in a stack zone of a few hundred k bytes organised into small pages, a scheme called context-to-stack mapping \cite{EfficientST80}, \cite{StackMappingVW}.
 Machine code generated by the JIT was executed by binding multiple processor simulators (first Bochs \cite{Bochs} for x86, SkyEye for ARMv6, Smalltalk code for 32-bit MIPS).
-A new object representation and garbage collector was added to also support 64-bits \cite{Spur}, which required refactoring the interpreter to allow the object representation to be chosen at startup.
+Finally, a new object representation and garbage collector was added to also support 64-bits \cite{Spur}, which required refactoring the interpreter to allow the object representation to be chosen at startup.
 Bochs was used a second time for x64 support.
+
+\egb{at this point as reading I have all the history and context to understand the major milestones in the development of Smaltalk VMs but I still do not know what this paper is about :P.
+You need to introduce now the goal of the paper or the problem you want to explain the reader you guys tackled.
+From the brief discussions I had last week with Clement, I was thinking that the paper will focus on explaining the tools that you guys built to help debugging and developing the VM.
+You don't have this focus currently in the text.
+I will continue reviewing to see if I can get it out the rest of the paper.}
 

### 2018-08-16 paper.tex
more changes on intro, and changes on section 2 up to section 2.1.

@@
 Part of the code base was, however, narrowed down to a subset of Smalltalk, called \emph{Slang}, to allow Smalltalk to C compilation \cite{BackToTheFuture}.
-
-\egb{Rewrote the paragraph to add old info from the abstract and add some flow to the explanation.
+\egb{Changed again this paragraph after reading section 2.
 Maybe we should also define here what simulation and production is (since the abstract is not considered to be part of the main body of the paper)}
-
+Dummy Smalltalk message sends were used to embed meta information in the code, such as \emph{self var: 'foo' type: 'char *'} which have no effect during simulation but guide the translation process of the executable Smalltalk.
 Effectively, the VM could be both simulated by executing the Slang code on top of the Smalltalk VM or compiled through the C compiler to native code to generate the production VM.
-The production VM can be derived by translating the core VM code to C, combining the generated C code with a set of platform-specific support files, and compiling with the platform's C compiler.
-Dummy Smalltalk message sends were used to embed meta information in the code, such as \emph{self var: 'foo' type: 'char *'} which have no effect during simulation but guide the translation process of the executable Smalltalk.
 
@@
 Over the years, the Squeak VM evolved to give birth recently to \openSmalltalkVM\footnote{https://github.com/OpenSmalltalk/opensmalltalk-vm/}, the default VM for various Smalltalk-like systems such as Pharo \cite{PharoByExample}, Squeak \cite{SqueakByExample}, Cuis, Croquet and Newspeak \cite{NewspeakOopsla}.
-
 As the VM evolved, the original simulator co-evolved as a tool to develop and debug the VM. The meta data used to guide the translation process was replaced by \emph{pragmas} \cite{Pragmas}.  
@@
 
-As shown in Figure \ref{fig:VMCompilation}, the VM code base is written both in Slang and in C.
-The Core VM code (the interpreter, the JIT and the memory manager), is written in Slang.
-The platform code, \emph{i.e.,} Operating System dependent code such as file management or I/O is written directly in C, with a corresponding Smalltalk version for simulation.
+Figure \ref{fig:VMCompilation} depicts the main components of \openSmalltalkVM and its compilation process.
+%the VM code base is written both in Slang and in C. 
+The Core VM code consists of the interpreter, the JIT and the memory manager, and is written in Slang.
+The platform code, \emph{i.e.,} Operating System dependent code (such as file management or I/O), is written directly in C with a corresponding Smalltalk version for simulation.
 %EXPLAIN scientifically in this part - no history. + FIGURE, basic with design.
@@
 		\includegraphics[width=0.55\linewidth]{figures/VMCompilation}
-		\caption{Cog VM compilation}
+		\caption{Cog \egb{u mean openSmalltalkVM?} VM compilation}
 		\label{fig:VMCompilation}
@@
 Secondly, the C compiler (depending on the platform, LLVM or GCC), translates the C code into an executable.
-This second step may take several dozens of seconds the first time, then, depending on what the programmer changed and what the C compiler have cached, it usually only takes a few seconds to recompile the C code.
+This second step may take several dozens of seconds the first time.
+But, depending on what the programmer changed and what the C compiler have cached, it usually only takes a few seconds to recompile the C code.
 
 The VM can be compiled in two main flavours, interpreter-only or interpreter+JIT.
-Although the version with the JIT is used in production, the interpreter version is convenient for special purposes.
-For example, debugging the garbage collector or evaluating new language features without dealing can be done in the interpreter-only VM, avoiding the JIT complexity.
+Although the version with the JIT is used in production, the interpreter version is convenient for development purposes.
+For example, debugging the garbage collector or evaluating new language features can be done in the interpreter-only VM, avoiding the JIT complexity.
 
-There are multiple reasons why we keep writing the core components in Slang and not in C/C++.
-Most of them are small details.
+There are multiple reasons why the core components are written in Slang and not in C/C++, being the most important VM simulation. 
+\egb{ I commented the texts on the details why you want to keep writing in Slang because they introduced noise to the explanation, and the important part is the simulation.}
+By interpreting the Slang code as Smalltalk code, emulating native code using an external processor simulator and simulating the memory using a large byte array, it is possible to simulate the whole VM execution.
+This allows development and debugging of the VM with the Smalltalk development tools resulting in a live programming experience for VM development.
+
+%Most of them are small details.
 For instance, the Slang-to-C compiler generates C code slightly different from the Slang code using different annotations.
@@
 By interpreting the Slang code as Smalltalk code, emulating native code using an external processor simulator and simulating the memory using a large byte array, we are able to simulate the whole VM execution, allowing development and debugging with the Smalltalk development tools and a live programming experience.
+
+\egb{maybe it makes more sense to move the information of Figure 1 and the compilation process to Introduction to finish the part of VM architecture/development history, and use this last paragraph as a motivation to say that having a live programming experience for VM development helped to catch bugs, be more productive..
+and mentioned that you in this paper you describe the architecture and how it helped?}
 

### 2018-08-16 paper.tex
fixed figure with Eliot
Add a reference for Monticello.  Delete middleabstract para p[er Elisa's suggestion

@@
 
-\egb{I think you can skip the next paragraph from the abstract.
-If the point is that you need to co-evolve both execution models once the VM changes, you can do that with 1 line in the previous paragraph.
-Now it may look like you will discuss co-evolution issues, but I guess you do not}
-
-As the VM has evolved, including the addition of a just-in-time compiler, better garbage collector algorithms, and better object representations, both execution models co-evolved.
-For example, the simulation infrastructure was extended with a processor simulator to simulate the code generated by the just-in-time compiler.
-
 %TODO egb: revisit this paragraph after reading all paper to refine problem statement and solution.
@@
 To specify the virtual machine (VM), the Smalltalk-80 team at Xerox PARC wrote a Smalltalk VM entirely in Smalltalk \cite{blueBook}.
-In 1995, members of the same team built an open-source Smalltalk \egb{dialect?} and VM, Squeak \cite{SqueakByExample}, using the code from \cite{blueBook} as a starting point.
-Part of the code base was, however, narrowed down to a subset of Smalltalk, called \emph{Slang}, to allow Smalltalk to C compilation \cite{BackToTheFuture}.
-\egb{Changed again this paragraph after reading section 2.
-Maybe we should also define here what simulation and production is (since the abstract is not considered to be part of the main body of the paper)}
+In 1995, members of the same team built Squeak \cite{BackToTheFuture}, an open-source Smalltalk dialect, Squeak \cite{SqueakByExample}, and its VM \cite{BackToTheFuture}, written in Smalltalk using the code from \cite{blueBook} as a starting point.
+Part of the code base was, however, narrowed down to a subset of Smalltalk, called \emph{Slang}, to allow Smalltalk to C compilation.
+Development was done in Smalltalk, an activity we call "Simulation", and the support code for which is "The Simulator".
+The production VM is derived by translating the core VM code to C, combining the generated C code with a set of platform-specific support files, and compiling with the platform's C compiler.
+Two execution models are effectively available, simulation, where the Smalltalk code is executed on top of a Smalltalk VM, and production, where the same code is compiled to executable code through the C compiler.
+Simulation is used to develop and debug the VM.
+Production is used to release the VM.
+
 Dummy Smalltalk message sends were used to embed meta information in the code, such as \emph{self var: 'foo' type: 'char *'} which have no effect during simulation but guide the translation process of the executable Smalltalk.
-Effectively, the VM could be both simulated by executing the Slang code on top of the Smalltalk VM or compiled through the C compiler to native code to generate the production VM.
 
-At this point, the VM consisted mainly in:
+When the Squeak VM was released it consisted mainly in:
 \begin{itemize}
@@
 \item WarpBlt, a rotation and scaling extension for the bit-based BitBLT graphics engine,
-\item and external C code to support running the VM on popular platforms.
+\item external C code and makefiles to support running the VM on popular platforms.
 \end{itemize}
@@
 I will continue reviewing to see if I can get it out the rest of the paper.}
+
+In this paper we present key details of the Simulation infrastructure, give examples that demonstrate its productivity advantages, and present disadvantages that the architecture presents.
 

### 2018-08-16 paper.tex
improved opening with GT in conclusion

@@
 		\centering
-		\includegraphics[width=0.9\linewidth]{figures/SimulatedMemory}
+		\includegraphics[width=0.95\linewidth]{figures/SimulatedMemory}
 		\caption{Runtime Memory and Simulated Counter-Part}
@@
 We believe it is a powerful tool allowing to reduce our development time and to allow to fix bugs quickly.
-In the near future, we plan to extend the simulator with customizable development tools, especially the moldable inspectors and debuggers \cite{MoldableInspector, MoldableDebugger}, to have a fancy user interface on top of the current simulation model, currently quite tricky to apprehend by new developers.
+
+In the near future, we plan to extend the simulator with customizable development tools.
+Currently the tools built in top of the simulation infrastructure is mainly textual.
+Frameworks such as the moldable inspectors and debuggers \cite{MoldableInspector, MoldableDebugger} will allow us to have a more interactive interface, making it easier to apprehend by new developers.
 

### 2018-08-16 paper.tex
fixed figure with eliot
rm fig 1 and added subtitle

@@
 \title[Two Decades of Smalltalk VM Development]{Two Decades of Smalltalk VM Development}
+\subtitle{Live VM development through Simulation Tools}
 
@@
 
-Figure \ref{fig:VMCompilation} depicts the main components of \openSmalltalkVM and its compilation process.
-%the VM code base is written both in Slang and in C. 
-The Core VM code consists of the interpreter, the JIT and the memory manager, and is written in Slang.
+The main components of \openSmalltalkVM, the interpreter, the JIT and the memory manager, are written in Slang.
 The platform code, \emph{i.e.,} Operating System dependent code (such as file management or I/O), is written directly in C with a corresponding Smalltalk version for simulation.
-%EXPLAIN scientifically in this part - no history. + FIGURE, basic with design.
-
-\begin{figure}[bth!]
-		\centering
-		\includegraphics[width=0.55\linewidth]{figures/VMCompilation}
-		\caption{Cog \egb{u mean openSmalltalkVM?} VM compilation}
-		\label{fig:VMCompilation}
-\end{figure}
 
@@
 		\centering
-		\includegraphics[width=0.95\linewidth]{figures/SimulatedMemory}
+		\includegraphics[width=0.9\linewidth]{figures/SimulatedMemory}
 		\caption{Runtime Memory and Simulated Counter-Part}

### 2018-08-16 paper.tex
updates on introduction and section 2 up till 2.1

@@
 
-\egb{The title is a bit misleading.
-It sounds great for a journal paper on the \openSmalltalkVM, but probably it is best to adapt it or add a subtitle on the specific topic you touch upon in this paper}
-
-\egb{I rewrote and shortened the first paragraph since the abstract was quite unbalance towards the research context but very scarce on details about the research problem and solution.
-I left the original text in comments below, and I moved some details to the introduction were they were more necessary in my opinion.}
+\egb{Todo: update the author list to add Dan Ingalls }
 
@@
 The production VM is derived by translating the core VM code to C, combining the generated C code with a set of platform-specific support files, and compiling with the platform's C compiler.
-Two execution models are effectively available, simulation, where the Smalltalk code is executed on top of a Smalltalk VM, and production, where the same code is compiled to executable code through the C compiler.
+
+Two execution models were effectively available, simulation, where the Smalltalk code is executed on top of a Smalltalk VM, and production, where the same code is compiled to executable code through the C compiler.
 Simulation is used to develop and debug the VM.
 Production is used to release the VM.
-
 Dummy Smalltalk message sends were used to embed meta information in the code, such as \emph{self var: 'foo' type: 'char *'} which have no effect during simulation but guide the translation process of the executable Smalltalk.
@@
 %One change moved the meta data that guides translation out of the executable Smalltalk into pragmas \cite{Pragmas}.  
-However, the most significant evolution of the simulator came with the introduction of the Just-In-Time compiler (JIT). 
+But the most significant evolution of the simulator came with the introduction of the Just-In-Time compiler (JIT). 
 %The code base was liberally sprinkled with assertions. 
@@
 
-\egb{at this point as reading I have all the history and context to understand the major milestones in the development of Smaltalk VMs but I still do not know what this paper is about :P.
-You need to introduce now the goal of the paper or the problem you want to explain the reader you guys tackled.
-From the brief discussions I had last week with Clement, I was thinking that the paper will focus on explaining the tools that you guys built to help debugging and developing the VM.
-You don't have this focus currently in the text.
-I will continue reviewing to see if I can get it out the rest of the paper.}
+One of the key design decisions in \openSmalltalkVM was to keep the core components (i.e.
+the interpreter, the JIT and the memory manager) written in Slang and not in C/C++.
+By interpreting the Slang code as Smalltalk code, emulating native code using an external processor simulator and simulating the memory using a large byte array, it is possible to simulate the whole VM execution.
+This allows development and debugging of the VM with the Smalltalk development tools resulting in a live programming experience for VM development.
 
-In this paper we present key details of the Simulation infrastructure, give examples that demonstrate its productivity advantages, and present disadvantages that the architecture presents.
-
-In the following section, we explain the VM infrastructure with both the compilation pipeline to generate the production VM and the simulation infrastructure used to develop and debug the VM.
+In this paper, we present key details of the Simulation infrastructure, give examples that demonstrate its productivity advantages, and present disadvantages that the architecture presents.
+The paper is structured as follows. Section \ref{sec:VMSimulation} introduces the simulation infrastructure used to develop and debug the VM.
+%In the following section, we explain the VM infrastructure with both the compilation pipeline to generate the production VM and the simulation infrastructure used to develop and debug the VM. 
 Section \ref{sec:GCExp} and Section \ref{sec:JITExp}  reports our experience developing respectively the full GC and the JIT with our infrastructure.
-Section \ref{sec:Analysis} shows how we abuse the simulator to analyse the runtime and direct design decisions to improve performance.
+Section \ref{sec:Analysis} shows how we reuse the simulator to analyse the runtime and direct design decisions to improve performance.
 Lastly, we discuss some related work and conclude.
@@
 
-The main components of \openSmalltalkVM, the interpreter, the JIT and the memory manager, are written in Slang.
+%The main components of \openSmalltalkVM, the interpreter, the JIT and the memory manager, are written in Slang.
 The platform code, \emph{i.e.,} Operating System dependent code (such as file management or I/O), is written directly in C with a corresponding Smalltalk version for simulation.
+Before detailing the simulation infrastructure used to develop and debug the \openSmalltalkVM, we first briefly introduce the compilation pipeline to generate the production VM.
 
@@
 
-The VM can be compiled in two main flavours, interpreter-only or interpreter+JIT.
+The VM can actually be compiled in two main flavours, interpreter-only or interpreter+JIT.
 Although the version with the JIT is used in production, the interpreter version is convenient for development purposes.
@@
 
-There are multiple reasons why the core components are written in Slang and not in C/C++, being the most important VM simulation. 
-\egb{ I commented the texts on the details why you want to keep writing in Slang because they introduced noise to the explanation, and the important part is the simulation.}
-By interpreting the Slang code as Smalltalk code, emulating native code using an external processor simulator and simulating the memory using a large byte array, it is possible to simulate the whole VM execution.
-This allows development and debugging of the VM with the Smalltalk development tools resulting in a live programming experience for VM development.
-
-%Most of them are small details.
+%There are multiple reasons why the core components are written in Slang and not in C/C++, being the most important VM simulation. 
+%\egb{ I commented the texts on the details why you want to keep writing in Slang because they introduced noise to the explanation, and the important part is the simulation.}
+%%Most of them are small details.
 For instance, the Slang-to-C compiler generates C code slightly different from the Slang code using different annotations.
 It can for example duplicate the implementation of specific methods with specific constant operands to generate more efficient code in the interpreter.
-The main reason why we keep using Slang over C/C++ is VM simulation.
-By interpreting the Slang code as Smalltalk code, emulating native code using an external processor simulator and simulating the memory using a large byte array, we are able to simulate the whole VM execution, allowing development and debugging with the Smalltalk development tools and a live programming experience.
-
-\egb{maybe it makes more sense to move the information of Figure 1 and the compilation process to Introduction to finish the part of VM architecture/development history, and use this last paragraph as a motivation to say that having a live programming experience for VM development helped to catch bugs, be more productive..
-and mentioned that you in this paper you describe the architecture and how it helped?}
+%By interpreting the Slang code as Smalltalk code, emulating native code using an external processor simulator and simulating the memory using a large byte array, it is possible to simulate the whole VM execution.
+This allows development and debugging of the VM with the Smalltalk development tools resulting in a live programming experience for VM development.
 

### 2018-08-16 paper.tex
pushing latest comments on section 2 for today

@@
 Objects may have their shape changed on the fly as instance variables are added and removed.
-When the programmer chooses, the developer takes a new snapshot, which includes its changes, and may include application instances under development [MAYBE THIS SHOULD BE DESCRIBED BELOW IN SNAPSHOTS].
-Deployment is performed using a snapshot containing the deployed application and the required core libraries (unused libraries and development tools may be removed from the deployed snapshot to decrease memory footprint and for security purposes).
+\egb{there is overlap here with respect to the previous paragraph. We already knew that a snapshot can be made during a coding session, and what a snapshot includes}
+  When the programmer chooses, he can take a new snapshot, which includes his code changes, and may include application instances under development [MAYBE THIS SHOULD BE DESCRIBED BELOW IN SNAPSHOTS].
+\egb{deployment here refers to starting a Smalltalk system from the previous paragraph, right?}Deployment is performed using a snapshot containing the deployed application and the required core libraries (unused libraries and development tools may be removed from the deployed snapshot to decrease memory footprint and for security purposes).
 \footnote{Code is organised in packages which are saved to an external version control system, in our case Monticello citation?
@@
 Figure \ref{fig:SimulatedMemory} describes, from a high-level perspective, the memory used by \openSmalltalkVM at runtime and the simulated counter-part.
-Let's detail briefly the memory used by the production VM, on the top of the figure.
-On the left side, which is usually the low addresses, we can find Text, the memory section holding the native code of the VM (the interpreter, the memory manager and the JIT itself, but not the code compiled by the JIT).
+Let us detail briefly the memory used by the production VM, on the top of the figure.
+On the left side, which is usually the low addresses, we can first find Text, the memory section holding the native code of the VM (the interpreter, the memory manager and the JIT itself, but not the code compiled by the JIT).
 Then, Data holds initialized and uninitialized data, mainly the C variables used globally in the VM.
 On higher addresses, we can find the beginning of the memory managed by the VM.
-At start-up, the VM mmaps a memory region used for the executable code generated by the JIT (optional executable section, exists only if the JIT is enabled), for the young objects and for the old objects present in the snapshot started plus a little more space for the first tenures\footnote{A tenure is the process of promoting young objects to old objects in a generational GC.} to be performed without triggering the full garbage collector.
+
+At start-up, the VM mmaps a memory region used for the executable code generated by the JIT (optional executable section, exists only if the JIT is enabled), for the young objects and for the old objects present in the snapshot started plus a little more space for the first \emph{tenures}\footnote{A tenure is the process of promoting young objects to old objects in a generational GC.} to be performed without triggering the full garbage collector.
 Later during execution, as old space grows, new memory regions are mmap'ed on higher addresses to store other old objects.
 Usually at very high addresses, we can find the C stack.
-Room in the C stack is alloca'ed at VM start-up to hold the Smalltalk stack.
+Room in the C stack is allocated at VM start-up to hold the Smalltalk stack.
 Both stacks are disjointed and managed differently.
@@
 References between objects are effectively indexes inside the byte array instead of pointers.
-All the Slang variables, normally translated to C variabled, are simulated as Smalltalk objects.
+All the Slang variables, normally translated to C variables, are simulated as Smalltalk objects.
 They used specific classes, such as \emph{CArrayAccessor} over normal Smalltalk classes, to emulate the C behavior (only array accesses are available in C, not high level iterator APIs, etc.).
@@
 All the Slang code is implemented in multiple Smalltalk classes, to organise the code and to add flexibility through polymorphism.
-For example, the AbstractCompactor class has two subclasses, one implements a sweep algorithm and the other a compact one.
+For example, the \texttt{AbstractCompactor} class has two subclasses, one implementing a sweep algorithm and the other a compact one.
 For production, at Slang-to-C compilation time, all the code is compiled in a single C file.
@@
 No polymorphism is available at runtime.
-However, since polymorphism is available in the simulator, it can be abused for debugging purposes.
-Still re-using the same example, the AbstractCompactor class has also simulation specific subclasses.
+However, since polymorphism is available in the simulator, it can be reused for debugging purposes.
+Still in our example, the \texttt{AbstractCompactor} class has also simulation specific subclasses.
 Such versions typically express additional constraints in the form of assertions which can be written in plain Smalltalk without restrictions to easily express complex constraints.

### 2018-08-16 paper.tex
Minor tweaks (more keywords, cite Monticello, Spur for perf and 64-bits).

@@
 \newcommand{\egb}[1]{\color{blue}\fbox{\bfseries\sffamily\scriptsize Elisa:}{\sf\small$\blacktriangleright$\textit{#1}$\blacktriangleleft$}\color{black}}
+\newcommand{\eem}[1]{\color{green}\fbox{\bfseries\sffamily\scriptsize Eliot:}{\sf\small$\blacktriangleright$\textit{#1}$\blacktriangleleft$}\color{black}}
 \newcommand{\cba}[1]{\color{purple}\fbox{\bfseries\sffamily\scriptsize Clement:}{\sf\small$\blacktriangleright$\textit{#1}$\blacktriangleleft$}\color{black}}
@@
 %% comma separated list
-\keywords{Just-in-Time compiler, virtual machine, managed runtime, tools}  %% \keywords are mandatory in final camera-ready submission
+\keywords{Just-in-Time compiler, garbage collector, virtual machine, managed runtime, tools, live development}  %% \keywords are mandatory in final camera-ready submission
 
@@
 To specify the virtual machine (VM), the Smalltalk-80 team at Xerox PARC wrote a Smalltalk VM entirely in Smalltalk \cite{blueBook}.
-In 1995, members of the same team built Squeak \cite{BackToTheFuture}, an open-source Smalltalk dialect, Squeak \cite{SqueakByExample}, and its VM \cite{BackToTheFuture}, written in Smalltalk using the code from \cite{blueBook} as a starting point.
+In 1995, members of the same team built Squeak \cite{SqueakByExample}, an open-source Smalltalk dialect, and its VM \cite{BackToTheFuture}, written in Smalltalk using the code from \cite{blueBook} as a starting point.
 Part of the code base was, however, narrowed down to a subset of Smalltalk, called \emph{Slang}, to allow Smalltalk to C compilation.
 Development was done in Smalltalk, an activity we call "Simulation", and the support code for which is "The Simulator".
-The production VM is derived by translating the core VM code to C, combining the generated C code with a set of platform-specific support files, and compiling with the platform's C compiler.
+The production VM is derived by translating the core VM code to C, combining this generated C code with a set of platform-specific support files, and compiling with the platform's C compiler.
 
@@
 Machine code generated by the JIT was executed by binding multiple processor simulators (first Bochs \cite{Bochs} for x86, SkyEye for ARMv6, Smalltalk code for 32-bit MIPS).
-Finally, a new object representation and garbage collector was added to also support 64-bits \cite{Spur}, which required refactoring the interpreter to allow the object representation to be chosen at startup.
+Finally, a new object representation and garbage collector was added to improve performance and also to support 64-bits \cite{Spur}.
+This required refactoring the interpreter to allow the object representation to be chosen at startup \footnote{The JIT was written with this eventuality in mind.}.
 Bochs was used a second time for x64 support.
@@
 
-The VM can actually be compiled in two main flavours, interpreter-only or interpreter+JIT.
-Although the version with the JIT is used in production, the interpreter version is convenient for development purposes.
+The VM can actually be configured in two main flavours, interpreter-only or interpreter+JIT.
+Although the version with the JIT is most often used in production, the interpreter version is convenient for development purposes \footnote{and on devices that outlaw JITs}.
 For example, debugging the garbage collector or evaluating new language features can be done in the interpreter-only VM, avoiding the JIT complexity.
@@
 \egb{deployment here refers to starting a Smalltalk system from the previous paragraph, right?}Deployment is performed using a snapshot containing the deployed application and the required core libraries (unused libraries and development tools may be removed from the deployed snapshot to decrease memory footprint and for security purposes).
-\footnote{Code is organised in packages which are saved to an external version control system, in our case Monticello citation?
-wiresong.ca}
+\footnote{Code is organised in packages which are saved to an external version control system, in our case Monticello \cite{Monticello}}.
 

### 2018-08-17 paper.tex
Write the "JIT fidelity and the interfacing of machine code to the simulation objects" paragraph (follows JIT simulation).

@@
 This first step takes several seconds.
-Secondly, the C compiler (depending on the platform, LLVM or GCC), translates the C code into an executable.
-This second step may take several dozens of seconds the first time.
-But, depending on what the programmer changed and what the C compiler have cached, it usually only takes a few seconds to recompile the C code.
+Secondly, the C compiler \footnote{depending on the platform, LLVM, GCC, MSVC} translates the C code into an executable.
+This second step may take a few minutes the first time.
+But, depending on what the programmer changed and what the C compiler has cached, it usually only takes a few seconds to recompile the C code.
 
@@
 Although the version with the JIT is most often used in production, the interpreter version is convenient for development purposes \footnote{and on devices that outlaw JITs}.
-For example, debugging the garbage collector or evaluating new language features can be done in the interpreter-only VM, avoiding the JIT complexity.
+For example, debugging the garbage collector or evaluating new language features can be done in the interpreter-only VM, avoiding JIT complexity.
 
@@
 The snapshot includes objects such as the classes, the compiled methods in the form of bytecodes and the running processes.
-At start-up, the VM restores the state of all objects in memory and resumes execution in the process active when the snapshot was created.
+At start-up, the VM restores the state of all objects in memory, swizzling pointers as required, and resumes execution in the process active when the snapshot was created.
 A new snapshot can be made during or at the end of the session.
@@
 Objects may have their shape changed on the fly as instance variables are added and removed.
-\egb{there is overlap here with respect to the previous paragraph. We already knew that a snapshot can be made during a coding session, and what a snapshot includes}
-  When the programmer chooses, he can take a new snapshot, which includes his code changes, and may include application instances under development [MAYBE THIS SHOULD BE DESCRIBED BELOW IN SNAPSHOTS].
-\egb{deployment here refers to starting a Smalltalk system from the previous paragraph, right?}Deployment is performed using a snapshot containing the deployed application and the required core libraries (unused libraries and development tools may be removed from the deployed snapshot to decrease memory footprint and for security purposes).
 \footnote{Code is organised in packages which are saved to an external version control system, in our case Monticello \cite{Monticello}}.
 
-Starting up \openSmalltalkVM from source files has never been supported.
-However, recent work~\cite{PharoBootstrap} allows a snapshot to be created from sources, indirectly allowing the VM to be started from source files.
-The bootstrap process uses the simulator to create and manipulate the heap and then to save the resulting snapshot.
+  Snapshots can be used to avoid long startup times for specific bugs.
+A simulation can be run to a point where the bug is about to be induced, the system snapshotted, and then multiple analyses of the bug undertaken by loading the snapshot and resuming the simulation, short cutting the set up time for the simulator.
+Simulation is of the order of a thousand times slower than actual execution.
 
@@
 As the VM can be compiled with and without the JIT, the VM can also be simulated with or without the JIT.
-Since executing machine code on the processor simulations is not fast, interpreter simulation may be faster, and startup time and memory footprint are smaller.
+Since executing machine code on the processor simulators is not fast, the simulated interpreter may be faster \footnote{but JIT code is so much more optimized than bytecode interpretation that it may indeed be faster, depending on the code being executed}, and startup time and memory footprint are smaller.
 
@@
 At start-up, the VM mmaps a memory region used for the executable code generated by the JIT (optional executable section, exists only if the JIT is enabled), for the young objects and for the old objects present in the snapshot started plus a little more space for the first \emph{tenures}\footnote{A tenure is the process of promoting young objects to old objects in a generational GC.} to be performed without triggering the full garbage collector.
-Later during execution, as old space grows, new memory regions are mmap'ed on higher addresses to store other old objects.
-Usually at very high addresses, we can find the C stack.
-Room in the C stack is allocated at VM start-up to hold the Smalltalk stack.
-Both stacks are disjointed and managed differently.
+Later during execution, as old space grows, new memory regions are mmap'ed at higher addresses to store other old objects.
+Usually at very high addresses, we find the C stack.
+Room in the C stack is allocated at VM start-up to hold the Smalltalk stack zone.
+Both stacks are disjoint and managed differently, the stack zone being broken up into small stack pages \cite{StackMappingVW}.
 
-In the simulator, the heap is simulated as a large contiguous byte array.
-References between objects are effectively indexes inside the byte array instead of pointers.
+In the simulator, the heap is stored as a large contiguous byte array.
+References between objects are indices inside the byte array instead of pointers.
 All the Slang variables, normally translated to C variables, are simulated as Smalltalk objects.
-They used specific classes, such as \emph{CArrayAccessor} over normal Smalltalk classes, to emulate the C behavior (only array accesses are available in C, not high level iterator APIs, etc.).
+They use specific wrapper classes, such as \emph{CArrayAccessor} over normal Smalltalk classes, to emulate the C behavior (only array accesses are available in C, not high level iterator APIs, etc.).
 The Slang code is executed as Smalltalk code.
-The Smalltalk stack is represented in \openSmalltalkVM as a double linked list of stack pages which are maintained by the VM.
-Each stack page is represented as a Smalltalk object in the simulator.
+The Smalltalk stack zone is represented in \openSmalltalkVM as a double linked list of stack pages which are maintained by the VM.
+In the interpeter configuration the stack zone is a Smalltalk Array object.
+In the JIT it is in the byte array, to allow access to the stack from generated machine code.
 
@@
 
-In Simulation, one of the core feature is to be able to re-use the whole Smalltalk IDE, including the browser, the inspectors and the debugger to develop and debug the VM.
-Most new features can be develop and debug interactively, adding code to the VM at runtime, in the simulation environment, as for normal Smalltalk programs.
+\eem{this feels repetitious.  we should say this once, and once only.  where is it best said?}
+In Simulation, one of the core feature is to be able to re-use the whole Smalltalk IDE, including the browser, inspectors and debugger to develop the VM.
+Most new features can be developed interactively, adding code to the VM at runtime, in the simulation environment, as for normal Smalltalk programming.
+
+\paragraph{Miscellaneous Uses}
+
+In general, the simulator provides a toolkit for manipulating and inspecting image files.
+Starting up \openSmalltalkVM from source files has never been supported.
+However, recent work~\cite{PharoBootstrap} allows a snapshot to be created from sources, indirectly allowing the VM to be started from source files.
+The bootstrap process uses the simulator to create and manipulate the heap and then to save the resulting snapshot.
+
+The simulator is also used to create 64-bit images from 32-bit images, a transformation that involves replacing certain instances of classes by certain others, for example, the 64-bit system supports an immediate floating point type that represents a subset of 64-bit IEEE 754 floating point numbers whose exponent fits in the same 8-bit range as 32-bit IEEE 754 floats, whereas in the 32-bit system all floats are 8 byte boxed objects.
 
@@
 This second version is slower, but it allows to implement specific debugging features, such as conditional breakpoints in-between each machine instruction.
+
+\paragraph{JIT fidelity and the interfacing of machine code to the simulation objects}
+A key requirement to enable effective development of the JIT is that the code generated during simulation be as close as possible to the code generated by the production JIT.
+The code is of course not bit-identical; the address space is laid out very differently in memory to the simulation's byte array and the simulator is not subject to Address Space Layout Randomisation, a feature we augment with a synthetic clock so that time advances in lock step with code execution.
+Hence simulations are exactly repeatable.
+But we benefit from the code generated under simulation being otherwise identical to that generated in the production VM and we go to some effort to achieve this.
+The technique we use, involving illegal addresses, also allows us to interface the Smalltalk objects that comprise the Simulation, the interpreter, memory manager and jit, and variables within them, with machine code.
+To explain this let's consider how the interpreter references a stack frame.
+
+The interpreter has framePointer, stackPointer and instructionPointer variables used to interpret code.
+When machine code wants to enter the run time, for example to have the run time look up a send, or to invoke the scavenger when eden is full, it needs to record the current stack frame, writing the native stack and frame pointers into the stackPointer and framePointer variables, and then to call the relevant routine in the run time.
+The run time can then access the current machine code stack frame though these variables.
+In the production VM, the stackPointer and framePointer variables are C variables at known addresses in Data, and routines in the runtime have been compiled from C and exit in Text.
+Referencing these variables and invoking runtime routines from machine code is therefore straight-forward.
+But in the simulation they are inaccessible.
+First of all they are instance variables inside Smalltalk objects, objects that have strong encapsulation, accessible only through messages, and second of all they are alongside the simulation's byte array, not within it, therefore the processor simulator has no access to these variables.
+
+The JIT maintains a set of Dictionaries (hash maps) that map integers representing addresses to closures that get or set variables in the simulation objects, or invoke runtime routines, in both cases by sending messages to the simulation objects.
+As code is JITted and specific variables and runtime routines are referenced the JIT manufactures a unique illegal address to reference the variable or routine, and uses this as the key in the relevant map, creating and storing the relevant closure accessor as the value.
+When the processor simulator attempts to execute an instruction containing one of these illegal addresses, which will be some variety of read, write, call, jump or return instruction \footnote{activations of JITted machine code methods return to calling interpreter frames by returning to a trampoline that transfers control to the interpreter; hence the return address of a machine code activation above an interpreter frame has this trampoline as its return address}, it will attempt to raise an exception.
+Our processor simulators are invoked through a "primitive" interface, much like an FFI call, and the attempt to raise an exeption is caught by wrapper code that returns an error result from the primitve invocation; in Smalltalk terminology the primitive fails.
+The primitive failure code, Smalltalk code in the method that contains the primitive invocation, then analyses the offending instruction and constructs a Smalltalk ProcessorException object.
+This Smalltalk exception has a type (\#read, \#write, \#jump, \#call), the address of the memory location being read, written, jumped to or called, a program counter (pc) holding the instruction's address, a next pc, and an accessor for the register, if any, that is the target of the read or the source of the write.
+The processor simulator is invoked in the context of an exception handler for ProcessorException exceptions which invokes a Smalltalk method, handleSimulationTrap:, that invokes the relevant closure.
+Once the closure has completed, the processor simulator is advanced to the nextpc and simulation resumes.
+In this way, machine code generated in simulation can access arbitrary simulation objects while remaining essentially identical to the generated machine code in the production VM.
+
+\paragraph{Machine code decoration, arbitrary breakpoints and history}
+Blah, blah, blah.
 

### 2018-08-17 paper.tex
Added section on Arbitrary breakpoints, history, and decorated disassembly

@@
 
-In addition to the interpreter simulator, simulating the JIT requires to simulate the execution of native code it generates.
+In addition to the interpreter simulator, simulating the JIT requires simulate the execution of the native code it generates.
 The JIT itself is written in Slang and simulated with the Smalltalk execution model.
@@
 
-Bindings to processor simulator libraries (Bochs for x86 and x64, Skyeye for ARMv6) were implemented so that the machine code can be simulated.
-Calls in-between slang code and direct machine code are a little bit trickier to simulate.
+Bindings to processor simulator libraries (Bochs for x86 and x64, Skyeye for ARMv6, SMalltalk code for 32-bit MIPSLE) were implemented so that the machine code can be executed safely.
+All but the MIPSLE simulators are accessed through SMalltalk primitives.
+These are effectively a combination of an FFI call to invoke foreign code plus Smalltalk code that is executed if (the code invoked by) the primitive returns an error.
+The next two sections describe in detail how this scheme is used to ensure that simulation and production machine code agree, how simulation machien code is interfaced to the rest of teh simulation, and the debugging facilities for machine code that are available.
+
+%Calls in-between slang code and direct machine code are a little bit trickier to simulate.
 Calls from machine code to slang code are implemented by using multiple invalid processor instructions, leading to a trap in the processor simulator.
@@
 
-Machine code simulation can be performed in two different ways.
-The processor simulator can start simulating code until it meets an invalid instruction.
+%Machine code simulation can be performed in two different ways.
+The processor simulator can start simulating code until it meets an invalid instruction, or until the VM's internal heartbeat ticks.
 That version is convenient because it is the fastest to execute, while still catching errors such as invalid memory accesses.
@@
 In this way, machine code generated in simulation can access arbitrary simulation objects while remaining essentially identical to the generated machine code in the production VM.
+Further, ProcessorException, and a handful of other methods implemented by a processor simulator makes the interfacing of machine code to the Smalltalk part of the simulation portable and processor independent.
 
-\paragraph{Machine code decoration, arbitrary breakpoints and history}
-Blah, blah, blah.
+\paragraph{Arbitrary breakpoints, history, and decorated disassembly}
+The JIT maintains a simulation-only variable that determines whether the procerssor smulator is invoked in single-stepping mode, or run-until-exception-or-interrupt mode.
+When in single stepping mode a circular buffer remembers N previous instructions, allowing one to examine an arbitrary number of instructions (by default 160) preceeding some error.
+The arbitrary breakpoint facility is intelligent enough to only enable single stepping once code has been JITted at that address, using run mode until that point.
+
+There is a small polymorphic scheme for breakpoints.
+Booleans, integers, and arrays understand isBreakpointFor: pc.
+So if the breakPC object is false there is no breakpoint.
+If the breakPC is true then any pc is a potential breakpoint.
+If it is an integer then that pc is a potential breakpoint.
+If the pc is an Array then the pc is a breakpoint if any value in the array answers true to isBreakpointFor:.
+The scheme could trivially be extended to include intervals.
+If breakPC isBreakpointFor: pc is true then the breakBlock is sent shouldStopIfAtPC: pc.
+booleans and closures understand shouldStopIfAtPC:.
+true and false make the breakpoint unconditional or disabled respectively, and a closure evaluates itself, allowing one to specify arbitrarily complex breakpoints.
+For example, one can specify that execution should stop at a particular pc if the top two elements on the stack satisfy some criterion.
+It is much more powerful and much simpler to use than typical machine-level debuggers, and it can be extended as one is debugging (for example, adding interval support during the middle of a debugging session if required).
+
+Each processor simulator is expected to be able to disassemble its machine code.
+The Smalltalk code for each processor simulator implements a simple parser that identifies constants and field offsets embedded in machine code and passes these to routines in the JIT and the interpreter that lookup addresses, matching them to objects, in particular message selectors and classes, and to method local variable names, rendering the machine code much more readable than machine code displayed in a typical machine oriented debugger.
+For example, here is a snippet of decorated disassembly for x64:
+
+\lstdefinelanguage
+   [x64]{Assembler}     % add a "x64" dialect of Assembler
+   [x86masm]{Assembler} % based on the "x86masm" dialect
+   % with these extra keywords:
+   {morekeywords={movq, cmpq, %
+                  rax,rdx,rcx,rbx,rsi,rdi,rsp,rbp, %
+                  r8,r8d,r8w,r8b,r9,r9d,r9w,r9b, %
+                  r10,r10d,r10w,r10b,r11,r11d,r11w,r11b, %
+                  r12,r12d,r12w,r12b,r13,r13d,r13w,r13b, %
+                  r14,r14d,r14w,r14b,r15,r15d,r15w,r15b}} % etc.
+
+\lstset{language=[x64]Assembler}
+
+\begin{lstlisting}
+00002063: movq 0x800(%rbx), %rax = 'stackLimit' : 48 8B 83 00 08 00 00 
+0000206a: cmpq %rax, %rsp : 48 39 C4 
+0000206d: jb .-0x47 (0x2028=indexOf:startingAt:ifAbsent:@28) : 72 B9 
+HasBytecodePC bc 40/41:
+0000206f: movq start@20(%rbp), %rsi : 48 8B 75 14 
+00002073: movq anElement@28(%rbp), %rdi : 48 8B 7D 1C 
+00002077: movq self@-12(%rbp), %rdx : 48 8B 55 F4 
+0000207b: movq $0x0, %rcx : 48 C7 C1 00 00 00 00 
+00002082: call .-0x1B2F (0x558=ceSend2Args) : E8 D1 E4 FF FF 
+IsSendCall indexOf:startingAt: bc 44/45:
+00002087: movq %rdx, index@-20(%rbp) : 48 89 55 EC 
+\end{lstlisting}
+
+This code is the end of the entry sequence that checks the stack pointer against the stack limit for the end of the current stack page, which is also used to break out of machine code to service interrupts or events such as invoking the scavenger, etc.
+\emph{\%rbx} is used as a base register pointing at the variables in interpreter.
+Following that is the code for \emph{index := self indexOf: anElement startingAt: start} where \emph{anElement} and \emph{start} are method parameters and \emph{index} is a local variable.
+HasBytecodePC and IsSendCall are decorations that identify important locations in machine code, including object references, runtime calls, etc.
+These points are specified using metadata implemented as a simple bytecoded language, a stream of such bytes being added to the end of each machine code method.
+This metadata is parsed when the garbage collector needs to locate object references in machine code, when linked sends must be relocated when the machine code zone is compacted to make space by throwing away some number of least recently used methods, and when machine code pcs must be mapped to their corresponding bytecode pcs to preserve the illusion that the VM is a bytecode interpreter using a spaghetti stack of Context objects.
 
@@
 While developing the JIT, this start-up time may still be too long and move the live programming experience to an edit-compile-run cycle, which we want to avoid.
+Further, a particular method that exercises some new part of the JIT may be problematic to have executed suring simulation and hence not easily JITted during simulation.
 
-To work around this problem, we implemented a tool called \emph{In-image compilation}.
-In-image compilation basically allows to call the JIT as a Smalltalk library on a given bytecode compiled method to generate the corresponding machine code, calls the bound processor simulator to disassemble the code and decorate the disassembly with Smalltalk specific information.
+To work around these problems, we implemented a tool called \emph{In-image compilation}.
+In-image compilation allows invoking the JIT as a Smalltalk library on any bytecode compiled method in the host Smalltalk system (i.e.
+\emph{not} in the simulated heap) to generate the corresponding machine code.
+In-image compilation invokes the JIT on the method, calls the bound processor simulator to disassemble the code and decorate the disassembly and then dumps the output in a text wwindow.
+Indeed the above snippet is from an in-image compilation.
 To generate the machine code, the JIT has to access specific objects (the compiled method, the literals, known objects such as true, false or nil) as if they were in the simulated memory.
-To work around this, we built a facade, which includes mock addresses for all the object the JIT may require to generate the machine code of a given method.
+To implement this we built a facade, which masquerades as the simulation's memory manager, and translates the state of the bytecode compiled method in the current simulation into state that makes it appear as if it were resident in a simulaiton byte array.
+This includes mock addresses for all the object the JIT may require to generate the machine code of a given method.
 Figure \ref{fig:InImageCompiler} summarizes the process.
@@
 
+Since we also have a bytecode assembler that allows us to manufacture bytecoded methods we can in practice exercise any part of the JIT by creating a suitable bytecoded method from the assembly code.
+Although for methods of the complexity produced by the adaptive optimiser this is hardly pleasurable.
+
 \subsection{Virtual Machine Simulation Limitations}
@@
 Due to the simulation infrastructure being different from the actual hardware, code is run differently and one could think it leads to strange bugs happening only in simulation or in production.
-However, we have been using this infrastructure since 1995, and this kind of bugs are really rare and usually easy to fix.
-We have however two main limitations: simulation performance which is quite slow and calls to external C/C++/machine code which cannot be simulated.
+However, we have been using this infrastructure since 1995, and these kinds of bugs are really rare and usually easy to fix.
+We have however two main limitations: simulation performance which is quite slow and calls to external C/C++/machine code which cannot (yet) be simulated.
 

### 2018-08-17 paper.tex
Add a placeholder for Dan.
fixed references according to my maniac format. Added ref to old VMIL paper. Added Dan as a co-author

--- 2018-08-17 paper.tex
@@
 \newcommand{\todo}[1]{\color{orange}\fbox{\bfseries\sffamily\scriptsize TODO:}{\sf\small$\blacktriangleright$\textit{#1}$\blacktriangleleft$}\color{black}}
-\newcommand{\sd}[1]{\color{red}\fbox{\bfseries\sffamily\scriptsize Stef:}{\sf\small$\blacktriangleright$\textit{#1}$\blacktriangleleft$}\color{black}}
 \newcommand{\egb}[1]{\color{blue}\fbox{\bfseries\sffamily\scriptsize Elisa:}{\sf\small$\blacktriangleright$\textit{#1}$\blacktriangleleft$}\color{black}}
@@
 \email{egonzale@vub.be}          %% \email is recommended
+
+%% Author with two affiliations and emails.
+\author{Daniel Ingalls}
+\affiliation{
+  % \position{}
+	%\department{Software Languages Lab}              %% \department is recommended
+	\institution{Y Combinator Research}            %% \institution is required
+	\city{San Francisco}
+  % \state{}
+  % \postcode{}
+	\country{California}                    %% \country is recommended
+}
+\email{TOFIX danhhingalls@gmail.com}          %% \email is recommended
 
@@
 %One change moved the meta data that guides translation out of the executable Smalltalk into pragmas \cite{Pragmas}.  
-But the most significant evolution of the simulator came with the introduction of the Just-In-Time compiler (JIT). 
+But the most significant evolution of the simulator came with the introduction of the Just-In-Time compiler (JIT) \cite{CogJIT}. 
 %The code base was liberally sprinkled with assertions. 
@@
 Machine code generated by the JIT was executed by binding multiple processor simulators (first Bochs \cite{Bochs} for x86, SkyEye for ARMv6, Smalltalk code for 32-bit MIPS).
-Finally, a new object representation and garbage collector was added to improve performance and also to support 64-bits \cite{Spur}.
-This required refactoring the interpreter to allow the object representation to be chosen at startup \footnote{The JIT was written with this eventuality in mind.}.
+Finally, a new object representation and garbage collector was added to improve performance and also to support 64-bits \cite{Forwarders}.
+This required refactoring the interpreter to allow the object representation to be chosen at startup\footnote{The JIT was written with this eventuality in mind.}.
 Bochs was used a second time for x64 support.

### 2018-08-17 paper.tex

--- 2018-08-17 paper.tex
@@
 
+%% Author with single affiliation.
+\author{Daniel Ingalls}
+                                        %% can be repeated if necessary
+\affiliation{
+  %\position{Position1}
+ % \department{VM team}              %% \department is recommended
+  \institution{ARCOS}            %% \institution is required
+ % \streetaddress{Street1 Address1}
+  \city{Aptos}
+  %\state{France}
+  %\postcode{Post-Code1}
+  \country{California}                    %% \country is recommended
+}
+\email{danhhingalls@gmail.com}          %% \email is recommended
+
 %% Abstract
@@
 \begin{abstract}
-
-\egb{Todo: update the author list to add Dan Ingalls }
 

### 2018-08-17 paper.tex
Finished sec 5.2 with description of how much faster the analysis is.

@@
 The question is how many send sites would be prematurely promoted to megamorphic, or how many closed PICs have selectors for which there are open PICs.
-Analysing the question is easy in our context.
 
@@
 The JIT maintains a linked list of open PICs, and before it creates a closed PIC for a send site it will patch it to an open PIC if the list contains one for the send's selector.
+
+Analysing the question is easy in our context.
+This analysis took about an hour to perform, including writing an iterator over send sites, and then implementing the eager promotion to open PICs.
+Someshat similar analyses one of the authors performed on the second generation Deutsch Schiffman VM \cite{EfficientST80} , which is written entirely in C, when adding the same PIC scheme took several days, having to be written statically and tested with a traditional edit-compile-debug cycle.
+The performance difference is extreme.
 

### 2018-08-17 paper.tex
Fixed two typos in sec 5.2

--- 2018-08-17 paper.tex
@@
 The JIT maintains a linked list of open PICs, and before it creates a closed PIC for a send site it will patch it to an open PIC if the list contains one for the send's selector.
+
+Analysing the question is easy in our context.
+This analysis took about an hour to perform, including writing an iterator over send sites, and then implementing the eager promotion to open PICs.
+Somewhat similar analyses one of the authors performed on the second generation Deutsch Schiffman VM \cite{EfficientST80} , which is written entirely in C, when adding the same PIC scheme took several days, having to be written statically and tested with a traditional edit-compile-debug cycle.
+The productivity difference is extreme.
 

### 2018-08-17 paper.tex
changes on section 2.1 up to JIT-simulation

@@
 
-In this section, we firstly describe briefly the start-up sequence of \openSmalltalkVM using snapshots, a key aspect of the VM simulator success.
+%Simulation allows for manipulating and inspecting Smalltalk image files.
+%In general, the simulator provides a toolkit for manipulating and inspecting image files.
+%\eem{this feels repetitious.  we should say this once, and once only.  where is it best said?}
+The key idea of the Simulation is to allow developers to reuse the whole Smalltalk IDE including the browser, inspectors and debugger to develop the VM.
+Most new features can be developed interactively, adding code to the VM at runtime, in the simulation environment, as for normal Smalltalk programming.
+
+%In Simulation, one of the core feature is to be able to re-use the whole Smalltalk IDE, including the browser, inspectors and debugger to develop the VM.
+Most new features can be developed interactively, adding code to the VM at runtime, in the simulation environment, as for normal Smalltalk programming.
+
+As the VM can be compiled with and without the JIT, the VM can also be simulated with or without the JIT.
+Before detailing the simulation of the interpreter-only VM and the simulation with the JIT, we briefly describe the start-up sequence of \openSmalltalkVM using snapshots, a key aspect of the VM simulator success.
+
+%In this section, we firstly describe briefly the start-up sequence of \openSmalltalkVM using snapshots, a key aspect of the VM simulator success.
 Then, we detail the simulation of the interpreter-only VM.
@@
 The entire system, including its development tools and application code is stored in a snapshot file, which is essentially a memory dump of the entire heap.
-A Smalltalk system is started by loading a snapshot file into a VM.
+%A Smalltalk system is started by loading a snapshot file into a VM.
 The snapshot includes objects such as the classes, the compiled methods in the form of bytecodes and the running processes.
 At start-up, the VM restores the state of all objects in memory, swizzling pointers as required, and resumes execution in the process active when the snapshot was created.
-A new snapshot can be made during or at the end of the session.
+%A new snapshot can be made during or at the end of the session.
+When programming with Smalltalk, the programmer usually starts from a snapshot which contains the core libraries, the development environment and the application under development. 
+More precisely, the snapshot includes objects (such as the classes), the compiled methods in the form of bytecodes and the running processes.
+Developing applications consists essentially in writing and editing code, which effectively installs, modifies and removes classes and compiled methods from objects.
+Programming may be done live, as the application under development is running.
+%, for example creating or revising methods in the debugger, and continuing.
+For example, objects may have their shape changed on the fly as instance variables are added and removed.
+\footnote{Code is organised in packages which are saved to an external version control system, in our case Monticello \cite{Monticello}}.
+A new snapshot can be made during or at the end of the development session.
 
-When programming with Smalltalk, the programmer usually starts from a snapshot which contains the core libraries, the development environment and the application under development.
-Development of the application consists essentially in writing and editing code, which effectively installs, modifies and removes classes and compiled methods from the set of live objects.
+%When programming with Smalltalk, the programmer usually starts from a snapshot which contains the core libraries, the development environment and the application under development. 
+%Development of the application consists essentially in writing and editing code, which effectively installs, modifies and removes classes and compiled methods from the set of live objects.
 Programming may be done live, as the application under development is running, for example creating or revising methods in the debugger, and continuing.
@@
 
+In general, the simulator provides a toolkit for manipulating and inspecting snapshot files.
   Snapshots can be used to avoid long startup times for specific bugs.
-A simulation can be run to a point where the bug is about to be induced, the system snapshotted, and then multiple analyses of the bug undertaken by loading the snapshot and resuming the simulation, short cutting the set up time for the simulator.
-Simulation is of the order of a thousand times slower than actual execution.
+A simulation can be run to a point where the bug is about to be injected, the system snapshotted, and then multiple analyses of the bug undertaken by loading the snapshot and resuming the simulation, short cutting the set up time for the simulator.
+  %Simulation is of the order of a thousand times slower than actual execution.
+
+%shorteinig the miscellaneous uses.
+Besides being a key tool in VM development, the simulator can be used to indirectly allow the VM to be started from source files ~\cite{PharoBootstrap}.
+The simulator is also used to create 64-bit images from 32-bit images, a transformation that involves replacing certain instances of classes by certain others (e.g.
+in the 32-bit system all floats are 8 byte boxed objects while 64-bit systems support an immediate floating point type that represents a subset of 64-bit IEEE 754 floating point numbers).
+
+%\paragraph{Miscellaneous Uses}
+%Starting up \openSmalltalkVM from source files has never been supported.
+However, recent work~\cite{PharoBootstrap} allows a snapshot to be created from sources, indirectly allowing the VM to be started from source files.
+%The bootstrap process uses the simulator to create and manipulate the heap and then to save the resulting snapshot.
+%The simulator is also used to create 64-bit images from 32-bit images, a transformation that involves replacing certain instances of classes by certain others, for example, the 64-bit system supports an immediate floating point type that represents a subset of 64-bit IEEE 754 floating point numbers whose exponent fits in the same 8-bit range as 32-bit IEEE 754 floats, whereas in the 32-bit system all floats are 8 byte boxed objects.
 
 \paragraph{Interpreter-only simulation.} 
-As the VM can be compiled with and without the JIT, the VM can also be simulated with or without the JIT.
+%As the VM can be compiled with and without the JIT, the VM can also be simulated with or without the JIT.
  Since executing machine code on the processor simulators is not fast, the simulated interpreter may be faster \footnote{but JIT code is so much more optimized than bytecode interpretation that it may indeed be faster, depending on the code being executed}, and startup time and memory footprint are smaller.
@@
 
-\eem{this feels repetitious.  we should say this once, and once only.  where is it best said?}
-In Simulation, one of the core feature is to be able to re-use the whole Smalltalk IDE, including the browser, inspectors and debugger to develop the VM.
-Most new features can be developed interactively, adding code to the VM at runtime, in the simulation environment, as for normal Smalltalk programming.
-
-\paragraph{Miscellaneous Uses}
-
-In general, the simulator provides a toolkit for manipulating and inspecting image files.
-Starting up \openSmalltalkVM from source files has never been supported.
-However, recent work~\cite{PharoBootstrap} allows a snapshot to be created from sources, indirectly allowing the VM to be started from source files.
-The bootstrap process uses the simulator to create and manipulate the heap and then to save the resulting snapshot.
-
-The simulator is also used to create 64-bit images from 32-bit images, a transformation that involves replacing certain instances of classes by certain others, for example, the 64-bit system supports an immediate floating point type that represents a subset of 64-bit IEEE 754 floating point numbers whose exponent fits in the same 8-bit range as 32-bit IEEE 754 floats, whereas in the 32-bit system all floats are 8 byte boxed objects.
-
 \paragraph{JIT simulation.}
@@
 In addition to the interpreter simulator, simulating the JIT requires simulate the execution of the native code it generates.
-The JIT itself is written in Slang and simulated with the Smalltalk execution model.
+The JIT itself is written in Slang and simulated wi th the Smalltalk execution model.
 To simulate the machine code, the start of the byte array representing the memory holds, when the JIT is enabled, the machine code generated at runtime.

### 2018-08-17 paper.tex
changes on related work and conclusion

@@
 
-\section{Related Work and Conclusion}
+\section{Related Work}
 
-Many VM developers implemented different tools to help them working more efficiently on their VM, but they rarely publish about it. We focus in this section on two related work.
+Many VM developers implemented different tools to help them working more efficiently on their VM, but they rarely publish about it. This section focuses on the two closest related works.
 
@@
 They allow to inspect the running state of the Maxine VM while it runs for debugging purposes.
-One of the main difference with out design is that the Maxine VM is metacircular, hence it does not have a simulation and a production mode as we do but a single production debuggable mode.
-We believe having two different modes allows us to easily generate a production artefact while still having nice debugging features.
-Having a full metacircular VM would be interesting.
-However, so far, most VMs used in production (Java, Javascript, etc.), even after the huge recent investments in the Javascript VMs by the four major web vendors, are still compiling through the C/C++ compiler and are not metacircular.
-Hence, although a metacircular VM has interesting advantages, it is not clear it is that convenient to build a VM in such a way.
+One of the main difference with our design is that the Maxine VM is metacircular, hence it does not have a simulation and a production mode as we do but a single production debuggable mode.
+We believe having two different modes allows us to easily generate a production VM while still having nice debugging features.
+Having a full metacircular VM would be interesting, but it is unclear whether it is convenient to build a production VM in that way.
+So far, most production VMs (Java, Javascript, etc.)are still compiling through the C/C++ compiler and are not metacircular
+
+%However, so far, most production VMs (Java, Javascript, etc.), even after the huge recent investments in the Javascript VMs by the four major web vendors, are still compiling through the C/C++ compiler and are not metacircular.
+%Hence, although a metacircular VM has interesting advantages, it is not clear it is that convenient to build a VM in such a way.
 
@@
 Most of the VM code is written in RPython, a restricted Python, instead of Slang, and some leftovers are written in plain C.
-The main difference is that RPython is much closer to Python than Slang is to Smalltalk.
-RPython allows higher level structures such as dictionaries to be used.
-The design decision comes with its set of advantages and drawbacks.
-The key advantage is that the RPython code feels like Python code and is relatively quite easy to read write, unlike Slang which feels like C and is as easy to write as C.
+RPython is, however, much closer to Python than Slang is to Smalltalk.
+%RPython allows higher-level structures such as dictionaries to be used. 
+%The design decision comes with its set of advantages and drawbacks. 
+The key advantage of such a design choice is that the RPython code feels like Python code and is relatively quite easy to read write, unlike Slang which feels like C and is as easy to write as C.
 The main drawback is that RPython to C compilation takes way longer than the Slang to C compilation (up to 40 minutes in a recent Macbook pro for the RSqueak VM \cite{RSqueak}, instead of several seconds for Slang).
 
-Although the RPython code can be executed as normal Python code, for some reasons, the developers seem to think it is not worth to do such a thing, mostly because executing code in this way is very slow.
-The overall architecture of the RPython toolchain is different, which leads to a longer time to reach peak performance (though their peak performance is at least in theory better than with our JIT).
+Although the RPython code can be executed as normal Python code, the developers seem to think it is not worth to do such a thing, mostly because executing code in this way is very slow. 
+The overall architecture of the RPython toolchain is different, resulting in a longer time to reach peak performance (though their peak performance is at least in theory better than with our JIT).
 This time may be quite significant in simulation mode.
-In addition, RPython was originally designed for Python, which does not feature snapshot by default, so they cannot abuse snapshots to work around the simulation slow performance.
+In contrast, in \openSmalltalkVM, we can reuse the snapshots to work around the simulation slow performance.
 
-\subsection*{Conclusion}
+%In addition, RPython was originally designed for Python, which does not feature snapshot by default, so they cannot abuse snapshots to work around the simulation slow performance.
 
-We introduced and discussed the \openSmalltalkVM simulation infrastructure, used to develop and debug the VM.
-We believe it is a powerful tool allowing to reduce our development time and to allow to fix bugs quickly.
+\subsection{Conclusion}
+
+This paper introduced and discussed the \openSmalltalkVM simulation infrastructure, used to develop and debug the VM.
+In particular, we described our experiences developing and debugging the full garbage collector and the JIT compiler.
+We believe Simulation is a powerful tool allowing us to reduce the development time and to fix bugs quickly.
 
 In the near future, we plan to extend the simulator with customizable development tools.
-Currently the tools built in top of the simulation infrastructure is mainly textual.
-Frameworks such as the moldable inspectors and debuggers \cite{MoldableInspector, MoldableDebugger} will allow us to have a more interactive interface, making it easier to apprehend by new developers.
+Currently the tools built in top of the simulation infrastructure are mainly textual.
+More advanced tooling such as the moldable inspectors and debuggers \cite{MoldableInspector, MoldableDebugger} should enable a more interactive interface, making it easier to apprehend by new developers.
 

### 2018-08-17 paper.tex
reworked the overall paper structure and added annotation on what to do for the next few hours for me

@@
 %TODO egb: revisit this paragraph after reading all paper to refine problem statement and solution.
-In this paper, we detail the VM simulation infrastructure and we report our experience developing and debugging the VM within it.
-We mention some of the limitations and how we worked around them.
-We discuss specifically how we use the VM simulator to develop and debug two core VM components, the garbage collector and the just-in-time compiler.
+In this paper, we detail the VM simulation infrastructure and we report our experience developing and debugging the garbage collector and the just-in-time compiler within it.
+%We mention some of the limitations and how we worked around them.
+%We discuss specifically how we use the VM simulator to develop and debug two core VM components, the garbage collector and the just-in-time compiler. 
 Then, we discuss how we use the simulation infrastructure to perform analysis on the runtime, directing some design decisions we have made to tune VM performance.
@@
 \item a memory manager with a compact but complex object representation: a pointer-reversing tracing garbage collector and a heap divided into two generations,
-\item WarpBlt, a rotation and scaling extension for the bit-based BitBLT graphics engine,
+\item WarpBlt, a rotation and scaling extension for the bit-based BitBlt graphics engine,
 \item external C code and makefiles to support running the VM on popular platforms.
@@
 %In the following section, we explain the VM infrastructure with both the compilation pipeline to generate the production VM and the simulation infrastructure used to develop and debug the VM. 
-Section \ref{sec:GCExp} and Section \ref{sec:JITExp}  reports our experience developing respectively the full GC and the JIT with our infrastructure.
-Section \ref{sec:Analysis} shows how we reuse the simulator to analyse the runtime and direct design decisions to improve performance.
-Lastly, we discuss some related work and conclude.
+Section \label{sec:Exp} reports our experience developing the VM with the simulation infrastructure. Lastly, we discuss some of the simulation limitations, the related work and conclude.
 
-\section{Virtual Machine Infrastructure}
+\section{Virtual Machine Simulation}
 \label{sec:VMSimulation}
 
-%The main components of \openSmalltalkVM, the interpreter, the JIT and the memory manager, are written in Slang.
-The platform code, \emph{i.e.,} Operating System dependent code (such as file management or I/O), is written directly in C with a corresponding Smalltalk version for simulation.
-Before detailing the simulation infrastructure used to develop and debug the \openSmalltalkVM, we first briefly introduce the compilation pipeline to generate the production VM.
+The key idea of the Simulation is to allow developers to reuse the whole Smalltalk IDE including the browser, inspectors and debugger to develop the VM.
+Most new features can be developed interactively, adding code to the VM at runtime, in the simulation environment, as for normal Smalltalk programming.
+
+%Before detailing the simulation infrastructure used to develop and debug the \openSmalltalkVM, 
+We first briefly introduce the compilation pipeline to generate the production VM and Smalltalk snapshots, key elements to understand the design of the simulation infrastructure.
+Then, we describe how the memory layout of both the production and simulation runtimes.
+In the following subsection, we explain specific aspects of the simulation infrastructure such as the simulation of the machine code generated by the JIT through the processor simulator.
+The last subsection discusses features related to the development of the JIT, including for example breakpoints in the machine code it generates.
+
+\subsection{Context}
+
+\paragraph{VM compilation.} 
+The main components of \openSmalltalkVM, the interpreter, the JIT and the memory manager, are written in Slang.
+The platform code, \emph{i.e.,} Operating System dependent code (such as file management or I/O), is written directly in C.
 
 The VM executable is generated in a two step process.
-Firstly, the Slang-to-C compiler translates the Slang code to C code, generating a few files.
+Firstly, the Slang-to-C compiler translates the Slang code to C code, generating a few C files.
 This first step takes several seconds.
-Secondly, the C compiler \footnote{depending on the platform, LLVM, GCC, MSVC} translates the C code into an executable.
-This second step may take a few minutes the first time.
+Secondly, the C compiler, depending on the platform, LLVM, GCC or MSVC, translates all the C files into an executable.
+This second step can take up to a few minutes the first time.
 But, depending on what the programmer changed and what the C compiler has cached, it usually only takes a few seconds to recompile the C code.
@@
 The VM can actually be configured in two main flavours, interpreter-only or interpreter+JIT.
-Although the version with the JIT is most often used in production, the interpreter version is convenient for development purposes \footnote{and on devices that outlaw JITs}.
+Although the version with the JIT is most often used in production, the interpreter version is convenient for development purposes and on devices that outlaw JITs.
 For example, debugging the garbage collector or evaluating new language features can be done in the interpreter-only VM, avoiding JIT complexity.
@@
 
-\subsection{Simulating the Virtual Machine}
-
-%Simulation allows for manipulating and inspecting Smalltalk image files.
-%In general, the simulator provides a toolkit for manipulating and inspecting image files.
-%\eem{this feels repetitious.  we should say this once, and once only.  where is it best said?}
-The key idea of the Simulation is to allow developers to reuse the whole Smalltalk IDE including the browser, inspectors and debugger to develop the VM.
-Most new features can be developed interactively, adding code to the VM at runtime, in the simulation environment, as for normal Smalltalk programming.
-
-%In Simulation, one of the core feature is to be able to re-use the whole Smalltalk IDE, including the browser, inspectors and debugger to develop the VM.
-Most new features can be developed interactively, adding code to the VM at runtime, in the simulation environment, as for normal Smalltalk programming.
-
-As the VM can be compiled with and without the JIT, the VM can also be simulated with or without the JIT.
-Before detailing the simulation of the interpreter-only VM and the simulation with the JIT, we briefly describe the start-up sequence of \openSmalltalkVM using snapshots, a key aspect of the VM simulator success.
-
-%In this section, we firstly describe briefly the start-up sequence of \openSmalltalkVM using snapshots, a key aspect of the VM simulator success.
-Then, we detail the simulation of the interpreter-only VM.
-The simulation of the full VM is a superset and is explained in the last subsection.
-
-\paragraph{Start-up from snapshots.} 
+\paragraph{Snapshots.}
 Smalltalk is an object \emph{system}, rather than a language.
@@
 %, for example creating or revising methods in the debugger, and continuing.
-For example, objects may have their shape changed on the fly as instance variables are added and removed.
-\footnote{Code is organised in packages which are saved to an external version control system, in our case Monticello \cite{Monticello}}.
-A new snapshot can be made during or at the end of the development session.
+For example, objects may have their shape changed on the fly as instance variables are added and removed%\footnote{Code is organised in packages which are saved to an external version control system, in our case Monticello \cite{Monticello}}.
+. A new snapshot can be made during or at the end of the development session.
 
-%When programming with Smalltalk, the programmer usually starts from a snapshot which contains the core libraries, the development environment and the application under development. 
-%Development of the application consists essentially in writing and editing code, which effectively installs, modifies and removes classes and compiled methods from the set of live objects.
-Programming may be done live, as the application under development is running, for example creating or revising methods in the debugger, and continuing.
-Objects may have their shape changed on the fly as instance variables are added and removed.
-\footnote{Code is organised in packages which are saved to an external version control system, in our case Monticello \cite{Monticello}}.
+Snapshots can be used to avoid long startup times for specific bugs.
+The VM can be run to a point where the bug is about to be injected, the system snapshotted, and then multiple analyses of the bug undertaken by loading the snapshot and resuming execution, either in the normal VM or in the simulator, short cutting the set up time to reach the bug.
 
-In general, the simulator provides a toolkit for manipulating and inspecting snapshot files.
-  Snapshots can be used to avoid long startup times for specific bugs.
-A simulation can be run to a point where the bug is about to be injected, the system snapshotted, and then multiple analyses of the bug undertaken by loading the snapshot and resuming the simulation, short cutting the set up time for the simulator.
-  %Simulation is of the order of a thousand times slower than actual execution.
-
-%shorteinig the miscellaneous uses.
-Besides being a key tool in VM development, the simulator can be used to indirectly allow the VM to be started from source files ~\cite{PharoBootstrap}.
-The simulator is also used to create 64-bit images from 32-bit images, a transformation that involves replacing certain instances of classes by certain others (e.g.
-in the 32-bit system all floats are 8 byte boxed objects while 64-bit systems support an immediate floating point type that represents a subset of 64-bit IEEE 754 floating point numbers).
-
-%\paragraph{Miscellaneous Uses}
-%Starting up \openSmalltalkVM from source files has never been supported.
-However, recent work~\cite{PharoBootstrap} allows a snapshot to be created from sources, indirectly allowing the VM to be started from source files.
-%The bootstrap process uses the simulator to create and manipulate the heap and then to save the resulting snapshot.
-%The simulator is also used to create 64-bit images from 32-bit images, a transformation that involves replacing certain instances of classes by certain others, for example, the 64-bit system supports an immediate floating point type that represents a subset of 64-bit IEEE 754 floating point numbers whose exponent fits in the same 8-bit range as 32-bit IEEE 754 floats, whereas in the 32-bit system all floats are 8 byte boxed objects.
-
-\paragraph{Interpreter-only simulation.} 
-%As the VM can be compiled with and without the JIT, the VM can also be simulated with or without the JIT.
- Since executing machine code on the processor simulators is not fast, the simulated interpreter may be faster \footnote{but JIT code is so much more optimized than bytecode interpretation that it may indeed be faster, depending on the code being executed}, and startup time and memory footprint are smaller.
+\subsection{Memory architectures}
 
@@
 
-Figure \ref{fig:SimulatedMemory} describes, from a high-level perspective, the memory used by \openSmalltalkVM at runtime and the simulated counter-part.
+\paragraph{Production memory layout.}Figure \ref{fig:SimulatedMemory} describes, from a high-level perspective, the memory used by \openSmalltalkVM at runtime and the simulated counter-part.
 Let us detail briefly the memory used by the production VM, on the top of the figure.
@@
 Then, Data holds initialized and uninitialized data, mainly the C variables used globally in the VM.
+
 On higher addresses, we can find the beginning of the memory managed by the VM.
-
 At start-up, the VM mmaps a memory region used for the executable code generated by the JIT (optional executable section, exists only if the JIT is enabled), for the young objects and for the old objects present in the snapshot started plus a little more space for the first \emph{tenures}\footnote{A tenure is the process of promoting young objects to old objects in a generational GC.} to be performed without triggering the full garbage collector.
 Later during execution, as old space grows, new memory regions are mmap'ed at higher addresses to store other old objects.
+
 Usually at very high addresses, we find the C stack.
@@
 
-In the simulator, the heap is stored as a large contiguous byte array.
+\paragraph{Simulation memory layout.}In the simulator, the heap is stored as a large contiguous byte array.
 References between objects are indices inside the byte array instead of pointers.
@@
 The Smalltalk stack zone is represented in \openSmalltalkVM as a double linked list of stack pages which are maintained by the VM.
-In the interpeter configuration the stack zone is a Smalltalk Array object.
-In the JIT it is in the byte array, to allow access to the stack from generated machine code.
+In the interpreter-only simulator, the stack zone is a Smalltalk Array object.
+In the full simulator, the stack zone is in the byte array, to allow access to the stack from generated machine code by the processor simulator\footnote{This difference is historical, the interpreter-only was implemented first in the easiest way possible, but it did not work out with the JIT.}.
+
+\subsection{Runtime simulation}
+
+\paragraph{Modularity.}
 
@@
 
-In addition to the interpreter simulator, simulating the JIT requires simulate the execution of the native code it generates.
+In addition to the interpreter simulator, simulating the JIT requires to simulate the execution of the native code it generates.
 The JIT itself is written in Slang and simulated wi th the Smalltalk execution model.
@@
 
-Bindings to processor simulator libraries (Bochs for x86 and x64, Skyeye for ARMv6, SMalltalk code for 32-bit MIPSLE) were implemented so that the machine code can be executed safely.
-All but the MIPSLE simulators are accessed through SMalltalk primitives.
+Bindings to processor simulator libraries (Bochs for x86 and x64, Skyeye for ARMv6, Smalltalk code for 32-bit MIPSLE) were implemented so that the machine code can be executed safely.
+All but the MIPSLE simulators are accessed through Smalltalk primitives.
 These are effectively a combination of an FFI call to invoke foreign code plus Smalltalk code that is executed if (the code invoked by) the primitive returns an error.
-The next two sections describe in detail how this scheme is used to ensure that simulation and production machine code agree, how simulation machien code is interfaced to the rest of teh simulation, and the debugging facilities for machine code that are available.
+The next two sections describe in detail how this scheme is used to ensure that simulation and production machine code agree, how simulation machine code is interfaced to the rest of the simulation, and the debugging facilities for machine code that are available.
 
@@
 
-\paragraph{JIT fidelity and the interfacing of machine code to the simulation objects}
+\paragraph{Deterministic simulation.}
+
+To be able to reproduce the same bug the exact same multiple times in a row, we designed the simulator to be as deterministic as possible. The most important things to note are the following:
+\begin{itemize}
+\item Simulated memory is not subject to Address Space Layout Randomisation,
+\item A synthetic clock is used so that time advances in lock step with code execution.
+\end{itemize}
+
+\paragraph{Interfacing machine code with simulation objects.}
+
 A key requirement to enable effective development of the JIT is that the code generated during simulation be as close as possible to the code generated by the production JIT.
-The code is of course not bit-identical; the address space is laid out very differently in memory to the simulation's byte array and the simulator is not subject to Address Space Layout Randomisation, a feature we augment with a synthetic clock so that time advances in lock step with code execution.
-Hence simulations are exactly repeatable.
+The code is of course not bit-identical; the address space is laid out very differently in memory to the simulation's byte array and the simulator is not subject to Address Space Layout Randomisation.
 But we benefit from the code generated under simulation being otherwise identical to that generated in the production VM and we go to some effort to achieve this.
@@
 
+\todo{Shorten and abstract next paragraph. }
+
 The JIT maintains a set of Dictionaries (hash maps) that map integers representing addresses to closures that get or set variables in the simulation objects, or invoke runtime routines, in both cases by sending messages to the simulation objects.
@@
 
-\paragraph{Arbitrary breakpoints, history, and decorated disassembly}
-The JIT maintains a simulation-only variable that determines whether the procerssor smulator is invoked in single-stepping mode, or run-until-exception-or-interrupt mode.
-When in single stepping mode a circular buffer remembers N previous instructions, allowing one to examine an arbitrary number of instructions (by default 160) preceeding some error.
-The arbitrary breakpoint facility is intelligent enough to only enable single stepping once code has been JITted at that address, using run mode until that point.
+\subsection{Machine code debugging}
+
+\paragraph{Instruction recording.}
+
+The JIT maintains a simulation-only variable that determines whether the processor simulator is invoked in single-stepping mode, or run-until-exception-or-interrupt mode.
+When in single stepping mode a circular buffer remembers N previous instructions and associated register state, allowing one to examine an arbitrary number of instructions (by default 160) preceding some error.
+The breakpoint facility is intelligent enough to only enable single stepping once code has been JITted at that address, using run mode until that point.
+
+\paragraph{Conditional breakpoints.}
+
+\todo{Rework. maybe check interaction with section JIT debug.}
 
@@
 It is much more powerful and much simpler to use than typical machine-level debuggers, and it can be extended as one is debugging (for example, adding interval support during the middle of a debugging session if required).
+
+\paragraph{In-image compilation.}
+
+\todo{Merge 1 and 2 + some shortening}
+
+%1
 
@@
 
+\todo{RM end of next paragraph or ref}
+
 This code is the end of the entry sequence that checks the stack pointer against the stack limit for the end of the current stack page, which is also used to break out of machine code to service interrupts or events such as invoking the scavenger, etc.
@@
 
-\subsection{In-image compilation}
+%2
 
@@
 \emph{not} in the simulated heap) to generate the corresponding machine code.
-In-image compilation invokes the JIT on the method, calls the bound processor simulator to disassemble the code and decorate the disassembly and then dumps the output in a text wwindow.
+In-image compilation invokes the JIT on the method, calls the bound processor simulator to disassemble the code and decorate the disassembly and then dumps the output in a text window.
 Indeed the above snippet is from an in-image compilation.
@@
 \paragraph{Templates.} The JIT generates a sequence of machine instructions for a given pattern of bytecodes.
-We will use the term template to describe the sequence of machine instructions generated for a pattern of bytecodes, even though we do not like to reduce the JIT to a simple template-based JIT since each bytecode generates slightly different machine instructions based on register pressure, a simulated stack and
+We use the term template to describe the sequence of machine instructions generated for a pattern of bytecodes, even though we do not like to reduce the JIT to a simple template-based JIT since each bytecode generates slightly different machine instructions based on register pressure, a simulated stack and
 a few heuristics. Since the JIT is template-based, in-image compilation is very convenient to develop and optimize each of the JIT templates.
@@
 
-\subsection{Virtual Machine Simulation Limitations}
+\section{Experience reports}
+\label{sec:Exp}
 
-The simulator has several limitations.
-Due to the simulation infrastructure being different from the actual hardware, code is run differently and one could think it leads to strange bugs happening only in simulation or in production.
-However, we have been using this infrastructure since 1995, and these kinds of bugs are really rare and usually easy to fix.
-We have however two main limitations: simulation performance which is quite slow and calls to external C/C++/machine code which cannot (yet) be simulated.
+\todo{add intro about this is anecdote and examples of how we used it ; fix section }
 
-\paragraph{Performance.}
-The first limitation is due to the simulation performance.
-The interpreter-only is simulator is around 200 times slower to execute code than the normal VM.
-With the JIT and processor simulation enabled, without specific debugging options such as conditional breakpoints in between machine instructions, simulation drops to around 500 times slower than the normal VM.
-We usually enabled conditional breakpoints in-between machine instructions only when we reach a point in the simulator where the bug is about to happen since it is way slower, so the overall performance in this context is not really relevant.
+\subsection{Implementing a new full GC algorithm}
 
-This means for example that if a GC bug happens in an application 15 minutes after start-up, it will take 50 hours to reproduce in the interpreter-only simulator.
-Bugs in the jitted code are worse.
-Fortunately, we work around this problem by using snapshots and the interpreter-only simulator for GC bugs.
-In general, once we are able to reproduce a bug in the production VM, we try to snapshot the runtime just before it crashes.
-The VM simulator can then be started just before the crash and the debugging tools can be used after only several dozens of seconds.
-If the bug is unrelated to the JIT, the interpreter-only simulator can also be used and it is a little bit quicker to execute code.
-
-\paragraph{Calls to external code.}
-Although most of the GC and JIT development and debugging can be done in the simulator, specific tasks cannot be done this way.
-Basically, any calls outside of the machine code generated by the JIT and the Slang code cannot be simulated.
-For specific small parts of the VM, such as file management, we extended the simulator, effectively duplicating the code base with the C code, to support those features in simulation.
-However, there is no solution in the general case: we cannot afford to simulate both the compiled C code and the jitted code on the processor simulator, that would be horribly slow, and specific behaviors in the machine code not present in the code generated by the JIT can hardly be simulated (Access to C variables, OS variables, etc.).
-
-The main limitation we have right now is with Foreign Function Interfaces (FFI).
-We have a significant amount of bugs in FFI, often due to specific interaction between call-backs, low-level assembly FFI specific glue code and moving objects.
-Such bugs cannot be debugged with our simulation infrastructure so far and we have to rely on gdb/lldb.
-
-\section{Garbage Collection Development}
-\label{sec:GCExp}
-
+\todo{Remove Assertion references}
+\todo{Make it more like a story like the following 2 subsec}
 Recently, to evaluate new old space garbage collection algorithms we designed against standard algorithms, we implemented a Mark-Sweep in addition to the existing Mark-Compact collector.
@@
 
-\paragraph{Assertions.}To stabilise our code base and find easily production bugs, all the code based is annotated with assertions.
+%\paragraph{Assertions.}To stabilise our code base and find easily production bugs, all the code based is annotated with assertions.
 Each assertion ensures a specific state is as the VM developer would expect it to be and stops or log the incorrect result if not.
@@
 
-\section{Just-in-Time Compiler Development}
-\label{sec:JITExp}
+%We firstly discuss how we use the simulator to debug crashes in deployed applications, then how we use in-image compilation to develop the JIT itself.
 
-We firstly discuss how we use the simulator to debug crashes in deployed applications, then how we use in-image compilation to develop the JIT itself.
-
-\subsection{Debugging crashes with conditional breakpoints in machine code}
+\subsection{Debugging a crash in machine code with conditional breakpoints}
 
@@
 
-\subsection{Optimizing the templates with in-image compilation}
+\subsection{Optimizing the store templates with in-image compilation}
 
@@
 
-\section{Virtual Machine Analysis: Directing performance decisions}
-\label{sec:Analysis}
+\subsection{Machine code zone send analysis}
+
+\todo{Make following 2 subsections more like a story like the previous ones}
 
@@
 Indeed, the simulator can be stopped at any given point and arbitrary Smalltalk code can be written and evaluated similarly to the \emph{eval} Javascript construct to analyse the simulated memory, including the machine code zone, the heap or any Smalltalk object representing the VM state.
-
-\subsection{Analysis example}
 
@@
 
-\subsection{Directing the VM behavior}
+\subsection{Directing VM optimizations through analysis}
 
@@
 This analysis took about an hour to perform, including writing an iterator over send sites, and then implementing the eager promotion to open PICs.
-Somewhat similar analyses one of the authors performed on the second generation Deutsch Schiffman VM \cite{EfficientST80} , which is written entirely in C, when adding the same PIC scheme took several days, having to be written statically and tested with a traditional edit-compile-debug cycle.
+Somewhat similar analyses one of the authors performed on the second generation Deutsch Schiffman VM \cite{EfficientST80}, which is written entirely in C, when adding the same PIC scheme took several days, having to be written statically and tested with a traditional edit-compile-debug cycle.
 The productivity difference is extreme.
 
-\section{Related Work}
+\section{Discussion and Related Work}
+
+\subsection{Virtual Machine simulation limitations}
+
+\todo{Merge 1 and 2}
+
+%1
+\paragraph{Interpreter-only simulation.} 
+ Since executing machine code on the processor simulators is not fast, the simulated interpreter may be faster \footnote{but JIT code is so much more optimized than bytecode interpretation that it may indeed be faster, depending on the code being executed}, and startup time and memory footprint are smaller.
+
+ %2
+
+ The simulator has several limitations.
+Due to the simulation infrastructure being different from the actual hardware, code is run differently and one could think it leads to strange bugs happening only in simulation or in production.
+However, we have been using this infrastructure since 1995, and these kinds of bugs are really rare and usually easy to fix.
+We have however two main limitations: simulation performance which is quite slow and calls to external C/C++/machine code which cannot (yet) be simulated.
+
+\paragraph{Performance.}
+The first limitation is due to the simulation performance.
+The interpreter-only is simulator is around 200 times slower to execute code than the normal VM.
+With the JIT and processor simulation enabled, without specific debugging options such as conditional breakpoints in between machine instructions, simulation drops to around 500 times slower than the normal VM.
+We usually enabled conditional breakpoints in-between machine instructions only when we reach a point in the simulator where the bug is about to happen since it is way slower, so the overall performance in this context is not really relevant.
+
+This means for example that if a GC bug happens in an application 15 minutes after start-up, it will take 50 hours to reproduce in the interpreter-only simulator.
+Bugs in the jitted code are worse.
+Fortunately, we work around this problem by using snapshots and the interpreter-only simulator for GC bugs.
+In general, once we are able to reproduce a bug in the production VM, we try to snapshot the runtime just before it crashes.
+The VM simulator can then be started just before the crash and the debugging tools can be used after only several dozens of seconds.
+If the bug is unrelated to the JIT, the interpreter-only simulator can also be used and it is a little bit quicker to execute code.
+
+\paragraph{Calls to external code.}
+Although most of the GC and JIT development and debugging can be done in the simulator, specific tasks cannot be done this way.
+Basically, any calls outside of the machine code generated by the JIT and the Slang code cannot be simulated.
+For specific small parts of the VM, such as file management, we extended the simulator, effectively duplicating the code base with the C code, to support those features in simulation.
+However, there is no solution in the general case: we cannot afford to simulate both the compiled C code and the jitted code on the processor simulator, that would be horribly slow, and specific behaviors in the machine code not present in the code generated by the JIT can hardly be simulated (Access to C variables, OS variables, etc.).
+
+The main limitation we have right now is with Foreign Function Interfaces (FFI).
+We have a significant amount of bugs in FFI, often due to specific interaction between call-backs, low-level assembly FFI specific glue code and moving objects.
+Such bugs cannot be debugged with our simulation infrastructure so far and we have to rely on gdb/lldb.
+
+\subsection{VM Simulator and Bootstrapping facilities}
+
+In general, the simulator provides a toolkit for manipulating and inspecting snapshot files.
+  Snapshots can be used to avoid long startup times for specific bugs.
+A simulation can be run to a point where the bug is about to be injected, the system snapshotted, and then multiple analyses of the bug undertaken by loading the snapshot and resuming the simulation, short cutting the set up time for the simulator.
+  %Simulation is of the order of a thousand times slower than actual execution.
+
+%shorteinig the miscellaneous uses.
+Besides being a key tool in VM development, the simulator can be used to indirectly allow the VM to be started from source files ~\cite{PharoBootstrap}.
+The simulator is also used to create 64-bit images from 32-bit images, a transformation that involves replacing certain instances of classes by certain others (e.g.
+in the 32-bit system all floats are 8 byte boxed objects while 64-bit systems support an immediate floating point type that represents a subset of 64-bit IEEE 754 floating point numbers).
+
+\subsection{Related Work}
 
 Many VM developers implemented different tools to help them working more efficiently on their VM, but they rarely publish about it. This section focuses on the two closest related works.
+
+\todo{We need to mention something about graal, maybe high level lang for VM and debugging in general then refs}
 
@@
 
-\subsection{Conclusion}
+\section*{Conclusion}
 

### 2018-08-17 paper.tex
finished structure rewamp. Now printing and making extra pass on the whole thing

@@
 		\includegraphics[width=0.9\linewidth]{figures/SimulatedMemory}
-		\caption{Runtime Memory and Simulated Counter-Part}
+		\caption{Runtime Memory and Simulated Counter-Part\vspace{-0.2cm}}
 		\label{fig:SimulatedMemory}
@@
 
-\todo{Shorten and abstract next paragraph. }
-
+%NOTE: I almost only commented part of what Eliot wrote but did not change anything. I basically removed implementation specific details.
 The JIT maintains a set of Dictionaries (hash maps) that map integers representing addresses to closures that get or set variables in the simulation objects, or invoke runtime routines, in both cases by sending messages to the simulation objects.
-As code is JITted and specific variables and runtime routines are referenced the JIT manufactures a unique illegal address to reference the variable or routine, and uses this as the key in the relevant map, creating and storing the relevant closure accessor as the value.
-When the processor simulator attempts to execute an instruction containing one of these illegal addresses, which will be some variety of read, write, call, jump or return instruction \footnote{activations of JITted machine code methods return to calling interpreter frames by returning to a trampoline that transfers control to the interpreter; hence the return address of a machine code activation above an interpreter frame has this trampoline as its return address}, it will attempt to raise an exception.
-Our processor simulators are invoked through a "primitive" interface, much like an FFI call, and the attempt to raise an exeption is caught by wrapper code that returns an error result from the primitve invocation; in Smalltalk terminology the primitive fails.
+As code is JITted and specific variables and runtime routines are referenced, the JIT manufactures a unique illegal address to reference the variable or routine, and uses this as the key in the relevant map, creating and storing the relevant closure accessor as the value.
+When the processor simulator attempts to execute an instruction containing one of these illegal addresses, which will be some variety of read, write, call, jump or return instruction \footnote{activations of JITted machine code methods return to calling interpreter frames by returning to a trampoline that transfers control to the interpreter; hence the return address of a machine code activation above an interpreter frame has this trampoline as its return address}, it raises a ProcessorException.
+%Our processor simulators are invoked through a "primitive" interface, much like an FFI call, and the attempt to raise an exeption is caught by wrapper code that returns an error result from the primitve invocation; in Smalltalk terminology the primitive fails.
 The primitive failure code, Smalltalk code in the method that contains the primitive invocation, then analyses the offending instruction and constructs a Smalltalk ProcessorException object.
 This Smalltalk exception has a type (\#read, \#write, \#jump, \#call), the address of the memory location being read, written, jumped to or called, a program counter (pc) holding the instruction's address, a next pc, and an accessor for the register, if any, that is the target of the read or the source of the write.
-The processor simulator is invoked in the context of an exception handler for ProcessorException exceptions which invokes a Smalltalk method, handleSimulationTrap:, that invokes the relevant closure.
-Once the closure has completed, the processor simulator is advanced to the nextpc and simulation resumes.
+The processor simulator is invoked in the context of an exception handler for ProcessorException exceptions which invokes %a Smalltalk method, handleSimulationTrap:, that invokes 
+the relevant closure.
+Once the closure has completed, the processor simulator is advanced to the next pc and simulation resumes.
 In this way, machine code generated in simulation can access arbitrary simulation objects while remaining essentially identical to the generated machine code in the production VM.
@@
 \paragraph{Conditional breakpoints.}
-
-\todo{Rework. maybe check interaction with section JIT debug.}
 
@@
 
-\todo{Merge 1 and 2 + some shortening}
+Simulating the whole VM requires going through the whole start-up sequence: loading the snapshot, running code registered in the start-up sequence and resuming the user interface.
+The whole start-up takes, in the simulator, around 15 seconds on a recent Macbook pro.
+While developing the JIT, this start-up time may still be too long and move the live programming experience to an edit-compile-run cycle, which we want to avoid.
+In addition, when developing a specific part of the JIT, it is convenient to compile a well-chosen simple method exercising that specific new part.
+This is difficult to do with the full simulator since that new part of the JIT might be called by the start-up logic, leading to the erratic behavior before expected.
 
-%1
+To work around these problems, we implemented a tool called \emph{In-image compilation}.
+In-image compilation allows invoking the JIT as a Smalltalk library on any bytecode compiled method in the host Smalltalk system (i.e.
+\emph{not} in the simulated heap) to generate the corresponding machine code.
+In-image compilation invokes the JIT on the method, calls the bound processor simulator to disassemble the code, decorates the disassembly and then dumps the output in a text window.
 
-Each processor simulator is expected to be able to disassemble its machine code.
-The Smalltalk code for each processor simulator implements a simple parser that identifies constants and field offsets embedded in machine code and passes these to routines in the JIT and the interpreter that lookup addresses, matching them to objects, in particular message selectors and classes, and to method local variable names, rendering the machine code much more readable than machine code displayed in a typical machine oriented debugger.
-For example, here is a snippet of decorated disassembly for x64:
+To generate the machine code, the JIT has to access specific objects (the compiled method, the literals, known objects such as true, false or nil) as if they were in the simulated memory.
+To implement this we built a facade, which masquerades as the simulation's memory manager, and translates the state of the bytecode compiled method in the current simulation into state that makes it appear as if it were resident in a simulation byte array.
+This includes mock addresses for all the object the JIT may require to generate the machine code of a given method.
+Figure \ref{fig:InImageCompiler} summarizes the in-image compilation process.
+Note that this technique applies for the baseline JIT, which translate a single bytecode method into machine code, adaptive optimizations and speculative optimizations are debugged differently.
+
+\begin{figure}[th!]
+		\centering
+		\includegraphics[width=\linewidth]{figures/InImageCompiler}
+		\caption{In-image compilation\vspace{-0.2cm}}
+		\label{fig:InImageCompiler}
+\end{figure}
 
@@
 
+\begin{figure}[th!]
+		\centering
 \begin{lstlisting}
@@
 \end{lstlisting}
+		\caption{Decorated disassembled code\vspace{-0.2cm}}
+		\label{fig:ASMDecorated}
+\end{figure}
 
-\todo{RM end of next paragraph or ref}
+\paragraph{Decorated Disassembly.}
+Each processor simulator is expected to be able to disassemble its machine code.
+The Smalltalk code for each processor simulator implements a simple parser that identifies constants and field offsets embedded in machine code and passes these to routines in the JIT and the interpreter that lookup addresses, matching them to objects, in particular message selectors and classes, and to method local variable names, rendering the machine code much more readable than machine code displayed in a typical machine oriented debugger.
+For example, Figure \ref{fig:ASMDecorated} is a snippet of decorated disassembly for x64.
 
@@
 These points are specified using metadata implemented as a simple bytecoded language, a stream of such bytes being added to the end of each machine code method.
-This metadata is parsed when the garbage collector needs to locate object references in machine code, when linked sends must be relocated when the machine code zone is compacted to make space by throwing away some number of least recently used methods, and when machine code pcs must be mapped to their corresponding bytecode pcs to preserve the illusion that the VM is a bytecode interpreter using a spaghetti stack of Context objects.
-
-%2
-
-Simulating the whole VM requires going through the whole start-up sequence: loading the snapshot, running code registered in the start-up sequence and resuming the user interface.
-The whole start-up takes around 15 seconds on a recent Macbook pro.
-While developing the JIT, this start-up time may still be too long and move the live programming experience to an edit-compile-run cycle, which we want to avoid.
-Further, a particular method that exercises some new part of the JIT may be problematic to have executed suring simulation and hence not easily JITted during simulation.
-
-To work around these problems, we implemented a tool called \emph{In-image compilation}.
-In-image compilation allows invoking the JIT as a Smalltalk library on any bytecode compiled method in the host Smalltalk system (i.e.
-\emph{not} in the simulated heap) to generate the corresponding machine code.
-In-image compilation invokes the JIT on the method, calls the bound processor simulator to disassemble the code and decorate the disassembly and then dumps the output in a text window.
-Indeed the above snippet is from an in-image compilation.
-To generate the machine code, the JIT has to access specific objects (the compiled method, the literals, known objects such as true, false or nil) as if they were in the simulated memory.
-To implement this we built a facade, which masquerades as the simulation's memory manager, and translates the state of the bytecode compiled method in the current simulation into state that makes it appear as if it were resident in a simulaiton byte array.
-This includes mock addresses for all the object the JIT may require to generate the machine code of a given method.
-Figure \ref{fig:InImageCompiler} summarizes the process.
-Note that this technique applies for the baseline JIT, which translate a single bytecode method into machine code, adaptive optimizations and speculative optimizations are debugged differently.
-
-\begin{figure}[ht!]
-		\centering
-		\includegraphics[width=\linewidth]{figures/InImageCompiler}
-		\caption{In-image compilation}
-		\label{fig:InImageCompiler}
-\end{figure}
+This metadata is parsed when the garbage collector needs to locate object references in machine code, when linked sends must be relocated when the machine code zone is compacted to make space by throwing away some number of least recently used methods, and when machine code pcs must be mapped to their corresponding bytecode pcs for debugging purposes.
+%to preserve the illusion that the VM is a bytecode interpreter using a spaghetti stack of Context objects.
 
 \paragraph{Templates.} The JIT generates a sequence of machine instructions for a given pattern of bytecodes.
-We use the term template to describe the sequence of machine instructions generated for a pattern of bytecodes, even though we do not like to reduce the JIT to a simple template-based JIT since each bytecode generates slightly different machine instructions based on register pressure, a simulated stack and
-a few heuristics. Since the JIT is template-based, in-image compilation is very convenient to develop and optimize each of the JIT templates.
-
-Since we also have a bytecode assembler that allows us to manufacture bytecoded methods we can in practice exercise any part of the JIT by creating a suitable bytecoded method from the assembly code.
+We use the term template to describe the sequence of machine instructions generated for a pattern of bytecodes, even though we do not like to reduce the JIT to a simple template-based JIT since each bytecode generates slightly different machine instructions based on register pressure, a simulated stack and a few heuristics.
+Since the JIT is template-based, in-image compilation is very convenient to develop and optimize each of the JIT templates.
+Since we also have a bytecode assembler that allows us to manufacture bytecoded methods we can in practice exercise any part of the JIT by creating a suitable bytecode method.
 Although for methods of the complexity produced by the adaptive optimiser this is hardly pleasurable.
@@
 
-\todo{add intro about this is anecdote and examples of how we used it ; fix section }
+In this section we describe our personal experience using the simulation infrastructure to develop the VM.
+The different subsection consists in different tasks we had to do, emphasising how we used the simulator and how it helped us perform the each given task.
 
@@
 
-\todo{Remove Assertion references}
-\todo{Make it more like a story like the following 2 subsec}
-Recently, to evaluate new old space garbage collection algorithms we designed against standard algorithms, we implemented a Mark-Sweep in addition to the existing Mark-Compact collector.
+Recently, to evaluate a new old space garbage collection algorithms, we decided to implement standard algorithms to get reference benchmark results.
+Since we had already a Mark-Compact collector, we mainly introduced a Mark-Sweep non moving GC as an alternative.
 The whole implementation was done in the simulator, and only when it was working there, was it compiled to C.
-Using this process, the compiled C code was kind of working out of the box.
-To describe the implementation process, we need to discuss briefly first the assertion levels in the VM.
+Using this process, the compiled C code was working out of the box.
+%To describe the implementation process, we need to discuss briefly first the assertion levels in the VM.
 Then we will show how we debugged the algorithm.
+
+We started by sketching the new Sweep algorithm in the code base.
+Once it was partially written, we started the simulator.
+Since the algorithm was not complete, we added the missing pieces inside the debugger, installing the new code at runtime and checking against the current values of the function parameters if the function behaved as expected, as one can do it in any Smalltalk application.
+
+In addition, the simulator has an interesting property: each time it performs a GC, either a scavenge or an old space collection, it first copies the simulated memory, performs the GC there, and if no exception was raised\footnote{The VM code base includes many assertions to ensure specific state is correct at given points in the execution, such assertions are optionally compiled in the production VM, they represent the majority of exceptions raised during development or bug resolution.}, it then performs the GC on the original version.
+This means that each time the algorithm did not work correctly, the erratic behavior happened in a copy of the heap.
+Since we can clone again the original version as many times as we want, we were able, each time the memory got corrupted beyond repair, to restart the exact same GC execution on the exact same heap with each objects at the exact same addresses.
+Indeed, in most cases when the bug happens, the memory was already very corrupted and it was very difficult to track back where the bug came from without rewinding the memory state.
+We reproduced the exact same GC many times in different cases to fully understand specific bugs.
 
@@
 The VM without assertions is the production VM.
-
-\paragraph{Simulation.} Once we had partially written the new Sweep algorithm, we started the simulator.
-Since the algorithm was partially written, we had to write the missing pieces inside the debugger, installing the new code at runtime, as one can do it in any Smalltalk application.
-In addition, the simulator has an interesting property: each time it performs a GC, either a scavenge or an old space collection, it first copies the simulated memory (\emph{i.e.,} the heap), performs the GC there, and if no assertion fails, it then performs the GC on the original version.
-This means that if the algorithm was not working, and likely an assertion would fail or an error such as incorrect memory access would be raised, the erratic behavior would have happened in a copy of the heap.
-The simulator can then reproduce the exact same erratic behavior as many times as the VM developer wants duplicating again and again the same original heap.
-This is very convenient to debug specific GC bugs.
-Indeed, often in GC bugs, when the bug happens, the memory is already corrupted and it's quite difficult to track back where the bug comes from without rewinding the memory state, which is usually very difficult or impossible.
 
@@
 From the machine instruction address, we could figure out which bytecode pattern generated the incorrect machine code (it was the new bytecode template for inlined allocations).
-From there, we built a simpler method crashing the runtime and fixed the template using in-image compilation (See next subsection).
+From there, we built a simpler method crashing the runtime and fixed the template using in-image compilation.
 
@@
 
-\todo{Make following 2 subsections more like a story like the previous ones}
-
-A side-effect of VM simulation, and specifically to be able to interrupt the simulation and introspect the simulated memory and simulation specific objects, is to be able to analyse the runtime with scripts written on-the-fly.
-Indeed, the simulator can be stopped at any given point and arbitrary Smalltalk code can be written and evaluated similarly to the \emph{eval} Javascript construct to analyse the simulated memory, including the machine code zone, the heap or any Smalltalk object representing the VM state.
+The simulator can be stopped at any given point and arbitrary Smalltalk code can be written and evaluated similarly to the \emph{eval} Javascript construct.
+We used this technique to analyse the simulated memory, including the machine code zone, the heap or any Smalltalk object representing the VM state.
 
@@
 
-The results of the analysis are sometimes used to direct performance design decisions on the VM.
-In this section we describe how the analysis impacted a design called "Early polymorphic inline cache promotion".
+The results of specific analysis are sometimes used to direct performance design decisions on the VM.
+In this section we describe how the analysis on polymorphic inline caches impacted a design called "Early polymorphic inline cache promotion".
 
@@
 
+In this section we first discuss the VM simulator limits and how we work around them, then the interaction between the simulator and bootstrapping processes.
+Lastly, we mention some relevant related works.
+
 \subsection{Virtual Machine simulation limitations}
-
-\todo{Merge 1 and 2}
-
-%1
-\paragraph{Interpreter-only simulation.} 
- Since executing machine code on the processor simulators is not fast, the simulated interpreter may be faster \footnote{but JIT code is so much more optimized than bytecode interpretation that it may indeed be faster, depending on the code being executed}, and startup time and memory footprint are smaller.
-
- %2
 
@@
 The first limitation is due to the simulation performance.
-The interpreter-only is simulator is around 200 times slower to execute code than the normal VM.
-With the JIT and processor simulation enabled, without specific debugging options such as conditional breakpoints in between machine instructions, simulation drops to around 500 times slower than the normal VM.
+The interpreter-only is simulator is between a hundred and a thousand times slower to execute code than the normal VM.
+With the JIT and processor simulation enabled, without specific debugging options such as conditional breakpoints in between machine instructions, simulation performance usually drops by a factor of two, though in specific cases it is actually faster.
 We usually enabled conditional breakpoints in-between machine instructions only when we reach a point in the simulator where the bug is about to happen since it is way slower, so the overall performance in this context is not really relevant.
 
-This means for example that if a GC bug happens in an application 15 minutes after start-up, it will take 50 hours to reproduce in the interpreter-only simulator.
-Bugs in the jitted code are worse.
-Fortunately, we work around this problem by using snapshots and the interpreter-only simulator for GC bugs.
-In general, once we are able to reproduce a bug in the production VM, we try to snapshot the runtime just before it crashes.
+In practice, if a GC bug happens in an application 15 minutes after start-up, it takes around 50 hours to reproduce in the simulator.
+Fortunately, we work around this problem by using snapshots.
+Once we are able to reproduce a bug in the production VM, we try to snapshot the runtime just before it crashes.
 The VM simulator can then be started just before the crash and the debugging tools can be used after only several dozens of seconds.
-If the bug is unrelated to the JIT, the interpreter-only simulator can also be used and it is a little bit quicker to execute code.
+In addition, since the interpreter-only simulator is usually quicker at executing code, if the bug is unrelated to the JIT (typically, a GC bug), the interpreter-only simulator can be used.
 
@@
 
-Many VM developers implemented different tools to help them working more efficiently on their VM, but they rarely publish about it. This section focuses on the two closest related works.
+Many VM developers implemented different tools to help them working more efficiently on their VM, but they rarely publish about it.
+For example, the teams working on the Truffle/Graal projects \cite{OneVMToRule} have very convenient development tools for their VMs such as a tool to visualize the intermediate representation of their JIT Graal at different point in the compilation process.
+But, despite an impressive number of publications on the projects, including publications on development tools for the guest languages \cite{TrufflePolyglotDebugger} or on specific parts of the VM \cite{GraalIR2}, no publication seems to be directly related to the VM development tools.
 
-\todo{We need to mention something about graal, maybe high level lang for VM and debugging in general then refs}
+This section discusses some related work: the Maxine inspector and the RPython toolchain infrastructure.
 
@@
 Having a full metacircular VM would be interesting, but it is unclear whether it is convenient to build a production VM in that way.
-So far, most production VMs (Java, Javascript, etc.)are still compiling through the C/C++ compiler and are not metacircular
+So far, most production VMs (Java, Javascript, etc.) are still compiling through the C/C++ compiler and are not metacircular.
 

### 2018-08-17 paper.tex
fixed last pass page 1-2

@@
 To specify the virtual machine (VM), the Smalltalk-80 team at Xerox PARC wrote a Smalltalk VM entirely in Smalltalk \cite{blueBook}.
-In 1995, members of the same team built Squeak \cite{SqueakByExample}, an open-source Smalltalk dialect, and its VM \cite{BackToTheFuture}, written in Smalltalk using the code from \cite{blueBook} as a starting point.
+In 1995, members of the same team built Squeak \cite{SqueakByExample}, an open-source Smalltalk dialect, and its VM \cite{BackToTheFuture}, written in Smalltalk using the code from Smalltalk-80 \cite{blueBook} as a starting point.
 Part of the code base was, however, narrowed down to a subset of Smalltalk, called \emph{Slang}, to allow Smalltalk to C compilation.
@@
 Over the years, the Squeak VM evolved to give birth recently to \openSmalltalkVM\footnote{https://github.com/OpenSmalltalk/opensmalltalk-vm/}, the default VM for various Smalltalk-like systems such as Pharo \cite{PharoByExample}, Squeak \cite{SqueakByExample}, Cuis, Croquet and Newspeak \cite{NewspeakOopsla}.
-As the VM evolved, the original simulator co-evolved as a tool to develop and debug the VM. The meta data used to guide the translation process was replaced by \emph{pragmas} \cite{Pragmas}.  
+As the VM evolved, the original simulator co-evolved as a tool to develop and debug the VM. The metadata used to guide the translation process was replaced by \emph{pragmas} \cite{Pragmas}.  
 %One change moved the meta data that guides translation out of the executable Smalltalk into pragmas \cite{Pragmas}.  
@@
 %The code base was liberally sprinkled with assertions. 
-The spaghetti stack was mapped to a more conventional stack frame organisation in a stack zone of a few hundred k bytes organised into small pages, a scheme called context-to-stack mapping \cite{EfficientST80}, \cite{StackMappingVW}.
-Machine code generated by the JIT was executed by binding multiple processor simulators (first Bochs \cite{Bochs} for x86, SkyEye for ARMv6, Smalltalk code for 32-bit MIPS).
+The spaghetti stack was mapped to a more conventional stack frame organisation in a stack zone of a few hundred kilobytes organised into small pages, a scheme called context-to-stack mapping \cite{EfficientST80,StackMappingVW}.
+Machine code generated by the JIT was executed by binding multiple processor simulators (Bochs \cite{Bochs} for x86 and x64, SkyEye for ARMv6, Smalltalk code for 32-bit MIPS).
 Finally, a new object representation and garbage collector was added to improve performance and also to support 64-bits \cite{Forwarders}.
-This required refactoring the interpreter to allow the object representation to be chosen at startup\footnote{The JIT was written with this eventuality in mind.}.
-Bochs was used a second time for x64 support.
+This required refactoring the interpreter to allow the object representation to be chosen at start-up\footnote{The JIT was written with this eventuality in mind.}.
+%Bochs was used a second time for x64 support.
 
@@
 
-In this paper, we present key details of the Simulation infrastructure, give examples that demonstrate its productivity advantages, and present disadvantages that the architecture presents.
+In this paper, we present key details of the Simulation infrastructure, give examples that demonstrate its productivity advantages, and discuss some of its limitations.
 The paper is structured as follows. Section \ref{sec:VMSimulation} introduces the simulation infrastructure used to develop and debug the VM.
 %In the following section, we explain the VM infrastructure with both the compilation pipeline to generate the production VM and the simulation infrastructure used to develop and debug the VM. 
-Section \label{sec:Exp} reports our experience developing the VM with the simulation infrastructure. Lastly, we discuss some of the simulation limitations, the related work and conclude.
+Section \ref{sec:Exp} reports our experience developing the VM with the simulation infrastructure.
+Section \ref{sec:Disc} discuss some of the simulation infrastructure limitations, how we work around them and related works.
 
@@
 %Before detailing the simulation infrastructure used to develop and debug the \openSmalltalkVM, 
-We first briefly introduce the compilation pipeline to generate the production VM and Smalltalk snapshots, key elements to understand the design of the simulation infrastructure.
-Then, we describe how the memory layout of both the production and simulation runtimes.
+In this section, we first briefly introduce the compilation pipeline to generate the production VM and Smalltalk snapshots, key elements to understand the design of the simulation infrastructure.
+Then, we describe the memory layout of both the production and simulation runtimes.
+Next, we detail how VM execution is simulated and the interactions with the simulated memory.
 In the following subsection, we explain specific aspects of the simulation infrastructure such as the simulation of the machine code generated by the JIT through the processor simulator.
-The last subsection discusses features related to the development of the JIT, including for example breakpoints in the machine code it generates.
+The last subsection discusses simulation features related to the development of the JIT.
 
@@
 \paragraph{VM compilation.} 
-The main components of \openSmalltalkVM, the interpreter, the JIT and the memory manager, are written in Slang.
+%The main components of \openSmalltalkVM, the interpreter, the JIT and the memory manager, are written in Slang.
 The platform code, \emph{i.e.,} Operating System dependent code (such as file management or I/O), is written directly in C.
@@
 The VM executable is generated in a two step process.
-Firstly, the Slang-to-C compiler translates the Slang code to C code, generating a few C files.
-This first step takes several seconds.
-Secondly, the C compiler, depending on the platform, LLVM, GCC or MSVC, translates all the C files into an executable.
-This second step can take up to a few minutes the first time.
-But, depending on what the programmer changed and what the C compiler has cached, it usually only takes a few seconds to recompile the C code.
+Firstly, the Slang-to-C compiler translates the Slang code (interpreter, JIT and GC code) to C code, generating a few C files.
+This first step usually takes several seconds.
+Secondly, the C compiler, depending on the platform, LLVM, GCC or MSVC, translates all the C files (generated C files and platform-specific C files) into an executable.
+This second step can take up to a few minutes the first time, but usually only takes a few seconds, since unchanged C files are kept in the C compiler cache.
 
-The VM can actually be configured in two main flavours, interpreter-only or interpreter+JIT.
-Although the version with the JIT is most often used in production, the interpreter version is convenient for development purposes and on devices that outlaw JITs.
+The VM can be configured in two main flavours, interpreter-only or interpreter+JIT.
+Although the version with the JIT is the most widely used in production, the interpreter version is convenient for development purposes and on devices that outlaw JITs.
 For example, debugging the garbage collector or evaluating new language features can be done in the interpreter-only VM, avoiding JIT complexity.
@@
 
-Snapshots can be used to avoid long startup times for specific bugs.
-The VM can be run to a point where the bug is about to be injected, the system snapshotted, and then multiple analyses of the bug undertaken by loading the snapshot and resuming execution, either in the normal VM or in the simulator, short cutting the set up time to reach the bug.
+Snapshots can be used to avoid long start-up times when fixing specific bugs.
+The VM can be run to a point where the bug is about to be injected and the system can be snapshotted at this point.
+Then, multiple analyses of the bug can be undertaken by loading the snapshot and resuming execution, either in the normal VM or in the simulator, short cutting the time to reach the bug.
 
@@
 
-In the recent years, we added support for a more aggressive JIT with speculative optimizations through a bytecode to bytecode optimizer \cite{SistaFastStartUp}, re-using the existing template JIT as a back-end.
+In the recent years, we added support for a more aggressive JIT with speculative optimizations through a bytecode to bytecode optimizer \cite{SistaFaststart-up}, re-using the existing template JIT as a back-end.
 To be able to generate efficient bytecodes in the bytecode to bytecode optimizer, we had to introduce new unsafe bytecodes allowing, for example, accesses in arrays without any checks (type and bounds checks) \cite{SistaBytecodeSet}.
@@
 \section{Discussion and Related Work}
+\label{sec:Disc}
 
@@
 In general, the simulator provides a toolkit for manipulating and inspecting snapshot files.
-  Snapshots can be used to avoid long startup times for specific bugs.
+  Snapshots can be used to avoid long start-up times for specific bugs.
 A simulation can be run to a point where the bug is about to be injected, the system snapshotted, and then multiple analyses of the bug undertaken by loading the snapshot and resuming the simulation, short cutting the set up time for the simulator.

### 2018-08-17 paper.tex
pass up to 2.4

@@
 Let us detail briefly the memory used by the production VM, on the top of the figure.
-On the left side, which is usually the low addresses, we can first find Text, the memory section holding the native code of the VM (the interpreter, the memory manager and the JIT itself, but not the code compiled by the JIT).
+On the left side, which is usually the low addresses, we can first find Text, the memory section holding the native code of the VM (the compiled code of the interpreter, the memory manager and the JIT itself, but not the code compiled by the JIT).
 Then, Data holds initialized and uninitialized data, mainly the C variables used globally in the VM.
@@
 
-All the Slang code is implemented in multiple Smalltalk classes, to organise the code and to add flexibility through polymorphism.
+All the Slang code is implemented in multiple Smalltalk classes, to organise the code and to add modularity through polymorphism.
 For example, the \texttt{AbstractCompactor} class has two subclasses, one implementing a sweep algorithm and the other a compact one.
 For production, at Slang-to-C compilation time, all the code is compiled in a single C file.
-All the flexibility is removed, using the same example, the VM developer chooses at this moment if he wants to compile a VM with a sweep or compact algorithm.
+All the modularity is removed, using the same example, the VM developer chooses at this moment if he wants to compile a VM with a sweep or compact algorithm.
 No polymorphism is available at runtime.
@@
 
+\paragraph{Deterministic simulation.}
+
+To be able to reproduce the same bug the exact same multiple times in a row, we designed the simulator to be as deterministic as possible. The most important things to note are the following:
+\begin{itemize}
+\item Simulated memory is not subject to Address Space Layout Randomisation,
+\item A synthetic clock is used so that time advances in lock step with code execution.
+\end{itemize}
+
 \paragraph{JIT simulation.}
@@
 In addition to the interpreter simulator, simulating the JIT requires to simulate the execution of the native code it generates.
-The JIT itself is written in Slang and simulated wi th the Smalltalk execution model.
+The JIT itself is written in Slang and simulated with the Smalltalk execution model.
 To simulate the machine code, the start of the byte array representing the memory holds, when the JIT is enabled, the machine code generated at runtime.
@@
 Bindings to processor simulator libraries (Bochs for x86 and x64, Skyeye for ARMv6, Smalltalk code for 32-bit MIPSLE) were implemented so that the machine code can be executed safely.
-All but the MIPSLE simulators are accessed through Smalltalk primitives.
-These are effectively a combination of an FFI call to invoke foreign code plus Smalltalk code that is executed if (the code invoked by) the primitive returns an error.
-The next two sections describe in detail how this scheme is used to ensure that simulation and production machine code agree, how simulation machine code is interfaced to the rest of the simulation, and the debugging facilities for machine code that are available.
+All but the MIPSLE simulators are accessed through %Smalltalk primitives.
+These are effectively a combination of an
+FFI calls and surrounding glue code.
+% to invoke foreign code plus Smalltalk code that is executed if (the code invoked by) the primitive returns an error. 
+The next paragraphs describe in detail how this scheme is used to ensure that simulation and production machine code are as similar as possible and how simulated machine code is interfaced to the rest of the simulation.
+%, and the debugging facilities for machine code that are available.
 
@@
 
-\paragraph{Deterministic simulation.}
-
-To be able to reproduce the same bug the exact same multiple times in a row, we designed the simulator to be as deterministic as possible. The most important things to note are the following:
-\begin{itemize}
-\item Simulated memory is not subject to Address Space Layout Randomisation,
-\item A synthetic clock is used so that time advances in lock step with code execution.
-\end{itemize}
-
 \paragraph{Interfacing machine code with simulation objects.}
 
-A key requirement to enable effective development of the JIT is that the code generated during simulation be as close as possible to the code generated by the production JIT.
+A key requirement to enable effective development of the JIT is the code generated during simulation to be as close as possible to the code generated by the production JIT.
 The code is of course not bit-identical; the address space is laid out very differently in memory to the simulation's byte array and the simulator is not subject to Address Space Layout Randomisation.
 But we benefit from the code generated under simulation being otherwise identical to that generated in the production VM and we go to some effort to achieve this.
-The technique we use, involving illegal addresses, also allows us to interface the Smalltalk objects that comprise the Simulation, the interpreter, memory manager and jit, and variables within them, with machine code.
+The technique we use, involving illegal addresses, also allows us to interface with machine code the Smalltalk objects that comprise the Simulation, the interpreter, memory manager and JIT, and variables within them.
 To explain this let's consider how the interpreter references a stack frame.
@@
 The interpreter has framePointer, stackPointer and instructionPointer variables used to interpret code.
-When machine code wants to enter the run time, for example to have the run time look up a send, or to invoke the scavenger when eden is full, it needs to record the current stack frame, writing the native stack and frame pointers into the stackPointer and framePointer variables, and then to call the relevant routine in the run time.
-The run time can then access the current machine code stack frame though these variables.
+When machine code wants to enter the runtime, for example to have the runtime look up a send on a cache miss, or to invoke the scavenger when eden is full, it needs to record the current stack frame, writing the native stack and frame pointers into the stackPointer and framePointer variables, and then to call the relevant routine.
+The runtime can then access the current machine code stack frame though these variables.
 In the production VM, the stackPointer and framePointer variables are C variables at known addresses in Data, and routines in the runtime have been compiled from C and exit in Text.
@@
 As code is JITted and specific variables and runtime routines are referenced, the JIT manufactures a unique illegal address to reference the variable or routine, and uses this as the key in the relevant map, creating and storing the relevant closure accessor as the value.
-When the processor simulator attempts to execute an instruction containing one of these illegal addresses, which will be some variety of read, write, call, jump or return instruction \footnote{activations of JITted machine code methods return to calling interpreter frames by returning to a trampoline that transfers control to the interpreter; hence the return address of a machine code activation above an interpreter frame has this trampoline as its return address}, it raises a ProcessorException.
+When the processor simulator attempts to execute an instruction containing one of these illegal addresses, which will be some variety of read, write, call, jump or return instruction\footnote{activations of JITted machine code methods return to calling interpreter frames by returning to a trampoline that transfers control to the interpreter; hence the return address of a machine code activation above an interpreter frame has this trampoline as its return address}, it raises a ProcessorException.
 %Our processor simulators are invoked through a "primitive" interface, much like an FFI call, and the attempt to raise an exeption is caught by wrapper code that returns an error result from the primitve invocation; in Smalltalk terminology the primitive fails.
@@
 This Smalltalk exception has a type (\#read, \#write, \#jump, \#call), the address of the memory location being read, written, jumped to or called, a program counter (pc) holding the instruction's address, a next pc, and an accessor for the register, if any, that is the target of the read or the source of the write.
-The processor simulator is invoked in the context of an exception handler for ProcessorException exceptions which invokes %a Smalltalk method, handleSimulationTrap:, that invokes 
+The processor simulator is invoked in the context of an exception handler for such exceptions which invokes %a Smalltalk method, handleSimulationTrap:, that invokes 
 the relevant closure.

### 2018-08-17 paper.tex
minor fixes

@@
 Since the JIT is template-based, in-image compilation is very convenient to develop and optimize each of the JIT templates.
+
 Since we also have a bytecode assembler that allows us to manufacture bytecoded methods we can in practice exercise any part of the JIT by creating a suitable bytecode method.
@@
 
-In this section we describe our personal experience using the simulation infrastructure to develop the VM.
-The different subsection consists in different tasks we had to do, emphasising how we used the simulator and how it helped us perform the each given task.
+In this section we describe our personal experiences using the simulation infrastructure to develop the VM.
+The different subsections consist in different tasks we had to implement and in each case we emphasise how we used the simulator and how it helped us perform the each given task.
 
@@
 
-In the recent years, we added support for a more aggressive JIT with speculative optimizations through a bytecode to bytecode optimizer \cite{SistaFaststart-up}, re-using the existing template JIT as a back-end.
+In the recent years, we added support for a more aggressive JIT with speculative optimizations through a bytecode to bytecode optimizer \cite{SistaFastStartup}, re-using the existing template JIT as a back-end.
 To be able to generate efficient bytecodes in the bytecode to bytecode optimizer, we had to introduce new unsafe bytecodes allowing, for example, accesses in arrays without any checks (type and bounds checks) \cite{SistaBytecodeSet}.

### 2018-08-17 paper.tex
typo

@@
 
-In this section we first discuss the VM simulator limits and how we work around them, then the interaction between the simulator and bootstrapping processes.
+In this section we first discuss the VM simulator limitations and how we work around them, then the interaction between the simulator and bootstrapping processes.
 Lastly, we mention some relevant related works.

### 2018-08-17 paper.tex
typo
minor typos
fixed all the reviews from Dan

@@
 		\centering
-		\includegraphics[width=0.9\linewidth]{figures/SimulatedMemory}
+		\includegraphics[width=0.88\linewidth]{figures/SimulatedMemory}
 		\caption{Runtime Memory and Simulated Counter-Part\vspace{-0.2cm}}
@@
 On higher addresses, we can find the beginning of the memory managed by the VM.
-At start-up, the VM mmaps a memory region used for the executable code generated by the JIT (optional executable section, exists only if the JIT is enabled), for the young objects and for the old objects present in the snapshot started plus a little more space for the first \emph{tenures}\footnote{A tenure is the process of promoting young objects to old objects in a generational GC.} to be performed without triggering the full garbage collector.
+At start-up, the VM mmaps a memory region used for the executable code generated by the JIT (this optional execution exists only if the JIT is enabled), for the young objects and for the old objects present in the snapshot started plus a little more space for the first \emph{tenures}\footnote{A tenure is the process of promoting young objects to old objects in a generational GC.} to be performed without triggering the full garbage collector.
 Later during execution, as old space grows, new memory regions are mmap'ed at higher addresses to store other old objects.
@@
 The interpreter has framePointer, stackPointer and instructionPointer variables used to interpret code.
-When machine code wants to enter the runtime, for example to have the runtime look up a send on a cache miss, or to invoke the scavenger when eden is full, it needs to record the current stack frame, writing the native stack and frame pointers into the stackPointer and framePointer variables, and then to call the relevant routine.
+When machine code wants to enter the runtime, for example to have the runtime look up a send on a cache miss, or to invoke the scavenger when young space is full, it needs to record the current stack frame, writing the native stack and frame pointers into the stackPointer and framePointer variables, and then to call the relevant routine.
 The runtime can then access the current machine code stack frame though these variables.
@@
 To implement this we built a facade, which masquerades as the simulation's memory manager, and translates the state of the bytecode compiled method in the current simulation into state that makes it appear as if it were resident in a simulation byte array.
-This includes mock addresses for all the object the JIT may require to generate the machine code of a given method.
+This includes mock addresses for all the objects the JIT may require to generate the machine code of a given method.
 Figure \ref{fig:InImageCompiler} summarizes the in-image compilation process.
-Note that this technique applies for the baseline JIT, which translate a single bytecode method into machine code, adaptive optimizations and speculative optimizations are debugged differently.
+Note that this technique applies for the baseline JIT, which translate a single bytecode method into machine code; adaptive optimizations and speculative optimizations are debugged differently.
 
@@
 Since we also have a bytecode assembler that allows us to manufacture bytecoded methods we can in practice exercise any part of the JIT by creating a suitable bytecode method.
-Although for methods of the complexity produced by the adaptive optimiser this is hardly pleasurable.
+This is hardly pleasurable for methods of the complexity produced by the adaptive optimiser though.
 
@@
 
-Recently, to evaluate a new old space garbage collection algorithms, we decided to implement standard algorithms to get reference benchmark results.
+Recently, to evaluate some new old space garbage collection algorithms, we decided to implement standard algorithms to get reference benchmark results.
 Since we had already a Mark-Compact collector, we mainly introduced a Mark-Sweep non moving GC as an alternative.
@@
 Once it was partially written, we started the simulator.
-Since the algorithm was not complete, we added the missing pieces inside the debugger, installing the new code at runtime and checking against the current values of the function parameters if the function behaved as expected, as one can do it in any Smalltalk application.
+Since the algorithm was not complete, we added the missing pieces inside the debugger, installing the new code at runtime and checking against the current values of the function parameters if the function behaved as expected, as one can do in any Smalltalk application.
 
@@
 This means that each time the algorithm did not work correctly, the erratic behavior happened in a copy of the heap.
-Since we can clone again the original version as many times as we want, we were able, each time the memory got corrupted beyond repair, to restart the exact same GC execution on the exact same heap with each objects at the exact same addresses.
-Indeed, in most cases when the bug happens, the memory was already very corrupted and it was very difficult to track back where the bug came from without rewinding the memory state.
+Since we can clone again the original version as many times as we want, we were able, each time the memory got corrupted beyond repair, to restart the exact same GC execution on the exact same heap with each object at the exact same addresses.
+Indeed, in most cases when a bug happens, the memory was already corrupted and it was very difficult to track back where the bug came from without rewinding the memory state.
 We reproduced the exact same GC many times in different cases to fully understand specific bugs.
@@
 
-The debugging process discussed here is exampled in a youtube video\footnote{https://www.youtube.com/watch?v=hctMBGAXVSs}.
+The debugging process discussed here is illustrated in a youtube video\footnote{https://www.youtube.com/watch?v=hctMBGAXVSs}.
 
@@
 
-\subsection{Machine code zone send analysis}
+\subsection{Analyzing the machine code zone}
 
@@
 
-One of the first analysis we ran was on the machine code zone.
+One of the first analyses we ran was on the machine code zone.
 We stopped the simulation when the machine code zone reached 1Mb.
-We then iterated over it and investigated what was in.
+We then iterated over it and investigated what was in it.
 As show in Table \ref{tab:generalAnalysis}, 1752 methods were compiled to machine code by the JIT, 6352 sends\footnote{We use the Smalltalk terminology, send, to discuss virtual calls since we are talking about Smalltalk.} are present but 2409 of them are not linked (basically, they have never been used).
@@
 
-Further analysis, in Table \ref{tab:polyAnalysis}, confirms Urs H\"{o}lzle statement \cite{PICSelf}: around 90\% of used send sites are monomorphic, around 9\% are polymorphic (up to 6 different cases in our implementation) and the remaining \% is megamorphic.
+Further analysis, in Table \ref{tab:polyAnalysis}, confirms Urs H\"{o}lzle's statement \cite{PICSelf}: around 90\% of used send sites are monomorphic, around 9\% are polymorphic (up to 6 different cases in our implementation) and the remaining \% is megamorphic.
 
@@
 
-The code used for these analysis is detailed in the Section "Let Me Tell You All About It, Let Me Quantify" of the blog post "Build me a JIT as fast as you can"\footnote{http://www.mirandabanda.org/cogblog/2011/03/01/build-me-a-jit-as-fast-as-you-can/}.
+The code used for these analyses is detailed in the Section "Let Me Tell You All About It, Let Me Quantify" of the blog post "Build me a JIT as fast as you can"\footnote{http://www.mirandabanda.org/cogblog/2011/03/01/build-me-a-jit-as-fast-as-you-can/}.
 
@@
 Due to the simulation infrastructure being different from the actual hardware, code is run differently and one could think it leads to strange bugs happening only in simulation or in production.
-However, we have been using this infrastructure since 1995, and these kinds of bugs are really rare and usually easy to fix.
+However, we have been using this infrastructure since 1995, and these kinds of bugs are rare and usually easy to fix.
 We have however two main limitations: simulation performance which is quite slow and calls to external C/C++/machine code which cannot (yet) be simulated.
@@
 The first limitation is due to the simulation performance.
-The interpreter-only is simulator is between a hundred and a thousand times slower to execute code than the normal VM.
+The interpreter-only simulator is between a hundred and a thousand times slower to execute code than the normal VM.
 With the JIT and processor simulation enabled, without specific debugging options such as conditional breakpoints in between machine instructions, simulation performance usually drops by a factor of two, though in specific cases it is actually faster.
-We usually enabled conditional breakpoints in-between machine instructions only when we reach a point in the simulator where the bug is about to happen since it is way slower, so the overall performance in this context is not really relevant.
+We usually enable conditional breakpoints in-between machine instructions only when we reach a point in the simulator where the bug is about to happen since it is way slower, so the overall performance in this context is not really relevant.
 
@@
 Once we are able to reproduce a bug in the production VM, we try to snapshot the runtime just before it crashes.
-The VM simulator can then be started just before the crash and the debugging tools can be used after only several dozens of seconds.
+The VM simulator can then be started just before the crash and the debugging tools can be used in far less than a minute.
 In addition, since the interpreter-only simulator is usually quicker at executing code, if the bug is unrelated to the JIT (typically, a GC bug), the interpreter-only simulator can be used.
@@
 \paragraph{Calls to external code.}
-Although most of the GC and JIT development and debugging can be done in the simulator, specific tasks cannot be done this way.
+Although most of the GC and JIT development and debugging can be done in the simulator, some specific tasks cannot be done this way.
 Basically, any calls outside of the machine code generated by the JIT and the Slang code cannot be simulated.
 For specific small parts of the VM, such as file management, we extended the simulator, effectively duplicating the code base with the C code, to support those features in simulation.
-However, there is no solution in the general case: we cannot afford to simulate both the compiled C code and the jitted code on the processor simulator, that would be horribly slow, and specific behaviors in the machine code not present in the code generated by the JIT can hardly be simulated (Access to C variables, OS variables, etc.).
+However, there is no solution in the general case: we cannot afford to simulate both the compiled C code and the jitted code on the processor simulator; that would be horribly slow, and specific behaviors in the machine code not present in the code generated by the JIT can hardly be simulated (Access to C variables, OS variables, etc.).
 
-The main limitation we have right now is with Foreign Function Interfaces (FFI).
-We have a significant amount of bugs in FFI, often due to specific interaction between call-backs, low-level assembly FFI specific glue code and moving objects.
+We have a significant amount of bugs in Foreign Function Interfaces (FFI), often due to specific interaction between call-backs, low-level assembly FFI specific glue code and moving objects.
 Such bugs cannot be debugged with our simulation infrastructure so far and we have to rely on gdb/lldb.
@@
 Many VM developers implemented different tools to help them working more efficiently on their VM, but they rarely publish about it.
-For example, the teams working on the Truffle/Graal projects \cite{OneVMToRule} have very convenient development tools for their VMs such as a tool to visualize the intermediate representation of their JIT Graal at different point in the compilation process.
+For example, the teams working on the Truffle/Graal projects \cite{OneVMToRule} have very convenient development tools for their VMs such as a tool to visualize the intermediate representation of their JIT Graal at different points in the compilation process.
 But, despite an impressive number of publications on the projects, including publications on development tools for the guest languages \cite{TrufflePolyglotDebugger} or on specific parts of the VM \cite{GraalIR2}, no publication seems to be directly related to the VM development tools.
@@
 \paragraph{Maxine Inspectors.} The Maxine inspectors \cite{MaxineInspector} were demonstrated at OOPSLA'18.
-They allow to inspect the running state of the Maxine VM while it runs for debugging purposes.
-One of the main difference with our design is that the Maxine VM is metacircular, hence it does not have a simulation and a production mode as we do but a single production debuggable mode.
+They allow one to inspect the running state of the Maxine VM while it runs for debugging purposes.
+One of the main differences with our design is that the Maxine VM is metacircular, hence it does not have a simulation and a production mode as we do but a single production debuggable mode.
 We believe having two different modes allows us to easily generate a production VM while still having nice debugging features.
@@
 In the near future, we plan to extend the simulator with customizable development tools.
-Currently the tools built in top of the simulation infrastructure are mainly textual.
-More advanced tooling such as the moldable inspectors and debuggers \cite{MoldableInspector, MoldableDebugger} should enable a more interactive interface, making it easier to apprehend by new developers.
+Currently the tools built on top of the simulation infrastructure are mainly textual.
+More advanced tooling such as moldable inspectors and debuggers \cite{MoldableInspector, MoldableDebugger} should enable a more interactive interface, making it easier to apprehend by new developers.
 

### 2018-08-17 paper.tex
Comprehensive pass over the English in the paper, eliminating unnecessary commas, indefinite articles, etc. Added an anecdote as a footnote describing why the simulator was retained for JIT development. Moved description of Lemming Debugging to its own section (its cool!). Described going in the pposite direction to the in-image compilation facade. Fixed capitalisation in a citation.

@@
 The production VM is derived by translating the core VM code to C.
-As a result, two execution models coexist: simulation, where the Smalltalk code is executed on top of a Smalltalk VM, and production, where the same code is compiled to executable code through the C compiler.
+As a result, two execution models coexist: simulation, where the Smalltalk code is executed on top of a Smalltalk VM, and production, where the same code is compiled to executable code through a C compiler.
 
@@
 The production VM is derived by translating the core VM code to C, combining the generated C code with a set of platform-specific support files, and compiling with the platform's C compiler.
-Two execution models are effectively available, simulation, where the Smalltalk code is executed on top of a Smalltalk VM, and production, where the same code is compiled to executable code through the C compiler.
+Two execution models are effectively available, simulation, where the Smalltalk code is executed on top of a Smalltalk VM, and production, where the same code is compiled to executable code through a C compiler.
 Simulation is used to develop and debug the VM.
@@
 
-To specify the virtual machine (VM), the Smalltalk-80 team at Xerox PARC wrote a Smalltalk VM entirely in Smalltalk \cite{blueBook}.
+To specify their virtual machine (VM), the Smalltalk-80 team at Xerox PARC wrote a Smalltalk VM entirely in Smalltalk \cite{blueBook}.
 In 1995, members of the same team built Squeak \cite{SqueakByExample}, an open-source Smalltalk dialect, and its VM \cite{BackToTheFuture}, written in Smalltalk using the code from Smalltalk-80 \cite{blueBook} as a starting point.
@@
 The spaghetti stack was mapped to a more conventional stack frame organisation in a stack zone of a few hundred kilobytes organised into small pages, a scheme called context-to-stack mapping \cite{EfficientST80,StackMappingVW}.
+A JIT was written.
 Machine code generated by the JIT was executed by binding multiple processor simulators (Bochs \cite{Bochs} for x86 and x64, SkyEye for ARMv6, Smalltalk code for 32-bit MIPS).
-Finally, a new object representation and garbage collector was added to improve performance and also to support 64-bits \cite{Forwarders}.
+Support for multiple bytecode sets was added, initially to support Newspeak, and more recently to support adaptive optimization.
+Finally, a new object representation and garbage collector was added to improve performance and also to support 64 bits \cite{Forwarders}.
 This required refactoring the interpreter to allow the object representation to be chosen at start-up\footnote{The JIT was written with this eventuality in mind.}.
 %Bochs was used a second time for x64 support.
+Slang was also extended with type inference to allow Smalltalk code that is independent of word size by virtue of Smalltalk's infinite precision arithmetic to be used in both 32 and 64 bit contexts.
 
@@
 By interpreting the Slang code as Smalltalk code, emulating native code using an external processor simulator and simulating the memory using a large byte array, it is possible to simulate the whole VM execution.
-This allows development and debugging of the VM with the Smalltalk development tools resulting in a live programming experience for VM development.
+This allows development and debugging of the VM with the Smalltalk development tools resulting in a live programming experience for VM development.\footnote{Arriving at the startup that first funded the OpenSmalltalk-VM, the first author had considerable experience maintaining the 2nd generation Deutsch Schiffmann VM written in C, and was sceptical that the Squeak Simulator would be viable for implementing a JIT.
+The startup wanted incremental development so as to reap value early and reduce risk, so it asked first for an interpreter using context-to-stack mapping.
+Writing this initial deliverable within the simulator was so much more productive and pleasurable than his previous work in C that he happily decided to stay with the simulator and invest in the necessary machine code simulation infrastructure.}
 
@@
 Secondly, the C compiler, depending on the platform, LLVM, GCC or MSVC, translates all the C files (generated C files and platform-specific C files) into an executable.
-This second step can take up to a few minutes the first time, but usually only takes a few seconds, since unchanged C files are kept in the C compiler cache.
+This second step can take up to a few minutes the first time, but usually only takes a few seconds, since object files for unchanged C files do not need to be recreated.
 
@@
 More precisely, the snapshot includes objects (such as the classes), the compiled methods in the form of bytecodes and the running processes.
-Developing applications consists essentially in writing and editing code, which effectively installs, modifies and removes classes and compiled methods from objects.
+Developing applications consists essentially in writing and editing code, which installs, modifies and removes classes and compiled methods from objects.
 Programming may be done live, as the application under development is running.
@@
 Snapshots can be used to avoid long start-up times when fixing specific bugs.
-The VM can be run to a point where the bug is about to be injected and the system can be snapshotted at this point.
-Then, multiple analyses of the bug can be undertaken by loading the snapshot and resuming execution, either in the normal VM or in the simulator, short cutting the time to reach the bug.
+The VM can be run to a point where the bug is about to be encountered and the system can be snapshotted at this point.
+Then multiple analyses of the bug can be undertaken by loading the snapshot and resuming execution, either in the normal VM or in the simulator, short cutting the time to reach the bug.
 
@@
 
-\paragraph{Production memory layout.}Figure \ref{fig:SimulatedMemory} describes, from a high-level perspective, the memory used by \openSmalltalkVM at runtime and the simulated counter-part.
+\paragraph{Production memory layout.}Figure \ref{fig:SimulatedMemory} describes, from a high-level perspective, the memory used by \openSmalltalkVM in the prducton VM and its simulated counter-part.
 Let us detail briefly the memory used by the production VM, on the top of the figure.
@@
 
-On higher addresses, we can find the beginning of the memory managed by the VM.
-At start-up, the VM mmaps a memory region used for the executable code generated by the JIT (this optional execution exists only if the JIT is enabled), for the young objects and for the old objects present in the snapshot started plus a little more space for the first \emph{tenures}\footnote{A tenure is the process of promoting young objects to old objects in a generational GC.} to be performed without triggering the full garbage collector.
+At higher addresses we find the beginning of the memory managed by the VM.
+At start-up the VM mmaps a memory region used for the executable code generated by the JIT (this optional execution exists only if the JIT is enabled), for new space (initially empty) and for the old objects present in the snapshot plus a little more space for the first \emph{tenures}\footnote{A tenure is the process of promoting young objects to old objects in a generational GC.} to be performed without triggering the full garbage collector.
 Later during execution, as old space grows, new memory regions are mmap'ed at higher addresses to store other old objects.
 
-Usually at very high addresses, we find the C stack.
+Usually at a very high addresse we find the C stack.
 Room in the C stack is allocated at VM start-up to hold the Smalltalk stack zone.
@@
 The Slang code is executed as Smalltalk code.
-The Smalltalk stack zone is represented in \openSmalltalkVM as a double linked list of stack pages which are maintained by the VM.
+The Smalltalk stack zone is represented in \openSmalltalkVM as a doubly linked cycle of stack pages which are maintained by the VM.
 In the interpreter-only simulator, the stack zone is a Smalltalk Array object.
-In the full simulator, the stack zone is in the byte array, to allow access to the stack from generated machine code by the processor simulator\footnote{This difference is historical, the interpreter-only was implemented first in the easiest way possible, but it did not work out with the JIT.}.
+In the full simulator, the stack zone is in the byte array, to allow access to the stack from generated machine code by the processor simulator\footnote{This difference is historical, the interpreter-only version was implemented first in the easiest way possible, but out with the JIT the Smalltalk stack must be accessible to the processor simulator.}.
 
@@
 
-To be able to reproduce the same bug the exact same multiple times in a row, we designed the simulator to be as deterministic as possible. The most important things to note are the following:
+To be able to reproduce the same bug exactly multiple times in a row, we designed the simulator to be as deterministic as possible. The most important features are:
 \begin{itemize}
@@
 
-In addition to the interpreter simulator, simulating the JIT requires to simulate the execution of the native code it generates.
+In addition to the interpreter simulator, simulating the JIT requires simulating execution of the native code it generates.
 The JIT itself is written in Slang and simulated with the Smalltalk execution model.
-To simulate the machine code, the start of the byte array representing the memory holds, when the JIT is enabled, the machine code generated at runtime.
+When the JIT is enabled, the start of the byte array representing the memory is used to hold the machine code it generates at runtime.
 
@@
 FFI calls and surrounding glue code.
+Each processor simulator is invoked via calls that supply the byte array as a parameter, and effectively the byte array is the processor simulator's memory, while the processor simulator contains its register state which is accessed via Smalltalk's support for external memory access.
+\footnote{http://www.squeaksource.com/Alien}
 % to invoke foreign code plus Smalltalk code that is executed if (the code invoked by) the primitive returns an error. 
@@
 
-A key requirement to enable effective development of the JIT is the code generated during simulation to be as close as possible to the code generated by the production JIT.
+A key requirement to enable effective development of the JIT is that the code generated during simulation be as close as possible to the code generated by the production JIT.
 The code is of course not bit-identical; the address space is laid out very differently in memory to the simulation's byte array and the simulator is not subject to Address Space Layout Randomisation.
 But we benefit from the code generated under simulation being otherwise identical to that generated in the production VM and we go to some effort to achieve this.
-The technique we use, involving illegal addresses, also allows us to interface with machine code the Smalltalk objects that comprise the Simulation, the interpreter, memory manager and JIT, and variables within them.
+The technique we use, involving illegal addresses, also allows us to interface generated machine code with the Smalltalk objects that comprise the Simulation, the interpreter, memory manager and JIT, and variables within them.
 To explain this let's consider how the interpreter references a stack frame.
 
-The interpreter has framePointer, stackPointer and instructionPointer variables used to interpret code.
-When machine code wants to enter the runtime, for example to have the runtime look up a send on a cache miss, or to invoke the scavenger when young space is full, it needs to record the current stack frame, writing the native stack and frame pointers into the stackPointer and framePointer variables, and then to call the relevant routine.
+The interpreter has \emph{framePointer}, \emph{stackPointer} and \emph{instructionPointer} variables used to interpret code.
+When machine code wants to enter the runtime, for example to have the runtime look up a send on a cache miss, or to invoke the scavenger when young space is full, it needs to record the current stack frame, writing the native stack and frame pointers into the \emph{stackPointer} and \emph{framePointer} variables, and then to call the relevant routine.
 The runtime can then access the current machine code stack frame though these variables.
-In the production VM, the stackPointer and framePointer variables are C variables at known addresses in Data, and routines in the runtime have been compiled from C and exit in Text.
-Referencing these variables and invoking runtime routines from machine code is therefore straight-forward.
+In the production VM, the \emph{stackPointer} and \emph{framePointer} are C variables in Data, and routines in the runtime have been compiled from C and exist in Text, and all have fixed addresses.
+Referencing these variables and invoking these runtime routines from machine code is therefore straight-forward.
 But in the simulation they are inaccessible.
@@
 The JIT maintains a set of Dictionaries (hash maps) that map integers representing addresses to closures that get or set variables in the simulation objects, or invoke runtime routines, in both cases by sending messages to the simulation objects.
-As code is JITted and specific variables and runtime routines are referenced, the JIT manufactures a unique illegal address to reference the variable or routine, and uses this as the key in the relevant map, creating and storing the relevant closure accessor as the value.
+As code is JITted and specific variables and runtime routines are referenced, the JIT manufactures a unique illegal address to reference each variable or routine, and uses this as the key in the relevant map, creating and storing a suitable closure accessor as the value.
 When the processor simulator attempts to execute an instruction containing one of these illegal addresses, which will be some variety of read, write, call, jump or return instruction\footnote{activations of JITted machine code methods return to calling interpreter frames by returning to a trampoline that transfers control to the interpreter; hence the return address of a machine code activation above an interpreter frame has this trampoline as its return address}, it raises a ProcessorException.
@@
 This Smalltalk exception has a type (\#read, \#write, \#jump, \#call), the address of the memory location being read, written, jumped to or called, a program counter (pc) holding the instruction's address, a next pc, and an accessor for the register, if any, that is the target of the read or the source of the write.
-The processor simulator is invoked in the context of an exception handler for such exceptions which invokes %a Smalltalk method, handleSimulationTrap:, that invokes 
+The processor simulator is always invoked in the context of an exception handler for such exceptions which invokes %a Smalltalk method, \emph{handleSimulationTrap:}, that invokes 
 the relevant closure.
@@
 In this way, machine code generated in simulation can access arbitrary simulation objects while remaining essentially identical to the generated machine code in the production VM.
-Further, ProcessorException, and a handful of other methods implemented by a processor simulator makes the interfacing of machine code to the Smalltalk part of the simulation portable and processor independent.
+Further, ProcessorException, and a handful of other methods implemented by a processor simulator makes the interfacing of machine code to the Smalltalk part of the simulation processor independent.
 
@@
 There is a small polymorphic scheme for breakpoints.
-Booleans, integers, and arrays understand isBreakpointFor: pc.
-So if the breakPC object is false there is no breakpoint.
-If the breakPC is true then any pc is a potential breakpoint.
+Booleans, integers, and arrays understand \emph{isBreakpointFor: pc}.
+So if the \emph{breakPC} variable is \emph{false} there is no breakpoint.
+If \emph{breakPC} is \emph{true} then any pc is a potential breakpoint.
 If it is an integer then that pc is a potential breakpoint.
-If the pc is an Array then the pc is a breakpoint if any value in the array answers true to isBreakpointFor:.
+If the pc is an \emph{Array} then the pc is a breakpoint if any value in the array answers true to \emph{isBreakpointFor:}.
 The scheme could trivially be extended to include intervals.
-If breakPC isBreakpointFor: pc is true then the breakBlock is sent shouldStopIfAtPC: pc.
-booleans and closures understand shouldStopIfAtPC:.
-true and false make the breakpoint unconditional or disabled respectively, and a closure evaluates itself, allowing one to specify arbitrarily complex breakpoints.
+If \emph{breakPC isBreakpointFor: pc} is \emph{true} then the \emph{breakBlock} is sent \emph{shouldStopIfAtPC: pc}.
+booleans and closures understand \emph{shouldStopIfAtPC:}.
+\emph{true} and \emph{false} make the breakpoint unconditional or disabled respectively, and a closure evaluates itself, allowing one to specify arbitrarily complex breakpoints.
 For example, one can specify that execution should stop at a particular pc if the top two elements on the stack satisfy some criterion.
-It is much more powerful and much simpler to use than typical machine-level debuggers, and it can be extended as one is debugging (for example, adding interval support during the middle of a debugging session if required).
+It is much more powerful and much simpler to use than typical machine-level debuggers, closures can be created at any point in any tool with a text interface, such as the debugger, and the scheme can be extended as one is debugging (for example, adding interval support during the middle of a debugging session if required).
 
@@
 
-Simulating the whole VM requires going through the whole start-up sequence: loading the snapshot, running code registered in the start-up sequence and resuming the user interface.
-The whole start-up takes, in the simulator, around 15 seconds on a recent Macbook pro.
-While developing the JIT, this start-up time may still be too long and move the live programming experience to an edit-compile-run cycle, which we want to avoid.
+Simulating the whole VM requires going through the entire Smalltalk system start-up sequence: loading the snapshot, running code registered in the start-up sequence and resuming the user interface.
+In the simulator start-up takes around 15 seconds on a recent MacBook Pro.
+While developing the JIT, this start-up time may still be too long and degrade the live programming experience to an edit-compile-run cycle, which we prefer to avoid.
 In addition, when developing a specific part of the JIT, it is convenient to compile a well-chosen simple method exercising that specific new part.
@@
 
-To generate the machine code, the JIT has to access specific objects (the compiled method, the literals, known objects such as true, false or nil) as if they were in the simulated memory.
+To generate the machine code, the JIT has to access specific objects (the compiled method, the literals, known objects such as \emph{true}, \emph{false} or \emph{nil}) as if they were in the simulated memory.
 To implement this we built a facade, which masquerades as the simulation's memory manager, and translates the state of the bytecode compiled method in the current simulation into state that makes it appear as if it were resident in a simulation byte array.
@@
 \end{figure}
+
+We also go in the opposite direction to the facade.
+Smalltalk provides a pretty printer for bytecoded methods.
+The class \emph{VMCompiledMethodProxy} is polymorphic with \emph{CompiledMethod} and makes methods in the simulation's heap appear to be ordinary \emph{CompiledMethods} in the Smalltalk system, allowing the pretty printer, and other tools, to be applied to methods in the simulation.
+This scheme can be extended as required; potentially any object in the simulation can be reified to be inspected using the Smalltalk tool suite.
 
@@
 Each processor simulator is expected to be able to disassemble its machine code.
-The Smalltalk code for each processor simulator implements a simple parser that identifies constants and field offsets embedded in machine code and passes these to routines in the JIT and the interpreter that lookup addresses, matching them to objects, in particular message selectors and classes, and to method local variable names, rendering the machine code much more readable than machine code displayed in a typical machine oriented debugger.
+The Smalltalk code for each processor simulator implements a simple parser that identifies constants and field offsets embedded in that disassembly and passes these to routines in the JIT and the interpreter that lookup addresses, matching them to objects, in particular message selectors and classes, and to method local variable names, rendering the disassembly much more readable than that displayed in a typical machine oriented debugger.
 For example, Figure \ref{fig:ASMDecorated} is a snippet of decorated disassembly for x64.
@@
 These points are specified using metadata implemented as a simple bytecoded language, a stream of such bytes being added to the end of each machine code method.
-This metadata is parsed when the garbage collector needs to locate object references in machine code, when linked sends must be relocated when the machine code zone is compacted to make space by throwing away some number of least recently used methods, and when machine code pcs must be mapped to their corresponding bytecode pcs for debugging purposes.
+This metadata is parsed when the garbage collector needs to locate object references in machine code, when linked sends must be relocated when the machine code zone is compacted to make space by throwing away some number of least recently used methods, and when machine code pcs must be mapped to their corresponding bytecode pcs since at the Smalltalk level only bytecode pcs are seen.
 %to preserve the illusion that the VM is a bytecode interpreter using a spaghetti stack of Context objects.
@@
 
-\subsection{Implementing a new full GC algorithm}
+\subsection{Initial JIT Development and New Back Ends}
+When the JIT was first being developed and was at a state where bytecoded methods could actually be translated to machine code, development fell into a very productive and enjoyable pattern.
+As a bytecode was encountered for which the template had yet to be written execution would stop in the debugger, which would create a skeleton implementation for the template method from the MessageNotUnderstood exception.
+The programmer would implement the body of the method and execution would continue.
+Were the new template definition to be incorrect and cause a bug during JIT generation, rather than when the generated machine code was run, then the debugger could be used to wind execution back to the beginning of the JITting of the current method and resume execution, repeating the cycle until disassembly looked good, somewhat like lemming debugging above.
+As a new method was encountered and JITted yet more templates would be encountered.
+Similarly as the abstract machine instruction to concrete machine instruction mappings were added these could be implemented in the debugger, an experience repeated when new back ends were added.
+Much of the JIT was written in this interactive live style.
+This style encourages tool creation because one can create, extend and polish tools in the middle of debugging sessions, enhancing them to make sense of ones current predicament.
+Immediate feedback makes tool investment cheap and cheerful.
+
+\subsection{Debugging New GC Algorithms and Lemming Debugging}
+
+Debugging graph rewriting garbage collectors can be hard; bugs leave the heap in a scrambled state, and figuring out what happened when one is left with a corrupted graph, often too broken to traverse and make sense of, or apply tools to, is often effectively impossible.
+But because the simulation is simply an object graph it is easy to copy the entire simulation.
+Therefore the simulator has been extended to by default take a copy of itself on every scavenge or old space reclamation, and to run the collection in the copy, performing a leak check afterwards (if enabled).
+Hence if a bug occurs or validation fails it will fail in the copy while the original remains as is, yet to be corrupted.
+We then take a fresh copy and use it to debug, repeating the process as many times as we like, an approach we call Lemming Debugging.
+One throws the copy off the cliff and if it gets to the bottom without incident it is descarded and the simulation continues in the original.
+This approach was key in rapid development of the new memory manager that supports both 32 and 64 bits \cite{Forwarders}.
 
@@
 
-In addition, the simulator has an interesting property: each time it performs a GC, either a scavenge or an old space collection, it first copies the simulated memory, performs the GC there, and if no exception was raised\footnote{The VM code base includes many assertions to ensure specific state is correct at given points in the execution, such assertions are optionally compiled in the production VM, they represent the majority of exceptions raised during development or bug resolution.}, it then performs the GC on the original version.
+%In addition, the simulator has an interesting property: each time it performs a GC, either a scavenge or an old space collection, it first copies the simulated memory, performs the GC there, and if no exception was raised\footnote{The VM code base includes many assertions to ensure specific state is correct at given points in the execution, such assertions are optionally compiled in the production VM, they represent the majority of exceptions raised during development or bug resolution.}, it then performs the GC on the original version.
 This means that each time the algorithm did not work correctly, the erratic behavior happened in a copy of the heap.
@@
 
-%\paragraph{Assertions.}To stabilise our code base and find easily production bugs, all the code based is annotated with assertions.
-Each assertion ensures a specific state is as the VM developer would expect it to be and stops or log the incorrect result if not.
+%\paragraph{Assertions.}To stabilise our code base and find production bugs as early as possible, lots of the code base is extended with assertions.
+Each assertion ensures a specific state is as the VM developer expects and stops or logs the incorrect result if not.
 We have a multi-level assertion system.
@@
 The VM without assertions is the production VM.
+Some, "expensive assertions" are so costly that they are optional, controlled with a flag in both C and simulation.
 
@@
 In the recent years, we added support for a more aggressive JIT with speculative optimizations through a bytecode to bytecode optimizer \cite{SistaFastStartup}, re-using the existing template JIT as a back-end.
-To be able to generate efficient bytecodes in the bytecode to bytecode optimizer, we had to introduce new unsafe bytecodes allowing, for example, accesses in arrays without any checks (type and bounds checks) \cite{SistaBytecodeSet}.
+To be able to generate efficient bytecodes in the bytecode to bytecode optimizer, we had to introduce new unsafe bytecodes allowing, for example, accesses in arrays without type and bounds checks \cite{SistaBytecodeSet}.
 For each new bytecode, we introduced new templates in the existing JIT to generate efficient machine code for the optimized methods using the new bytecodes.
@@
 
-In this section we first discuss the VM simulator limitations and how we work around them, then the interaction between the simulator and bootstrapping processes.
-Lastly, we mention some relevant related works.
+In this section we first discuss the VM simulator's limitations and how we work around them, and then the interaction between the simulator and bootstrapping processes.
+Lastly, we mention some relevant related work.
 
@@
 With the JIT and processor simulation enabled, without specific debugging options such as conditional breakpoints in between machine instructions, simulation performance usually drops by a factor of two, though in specific cases it is actually faster.
-We usually enable conditional breakpoints in-between machine instructions only when we reach a point in the simulator where the bug is about to happen since it is way slower, so the overall performance in this context is not really relevant.
+We usually enable conditional breakpoints in-between machine instructions only when we reach a point in the simulator where the bug is about to happen since single stepping is much slower, so the overall performance in this context is not really relevant.
 
-In practice, if a GC bug happens in an application 15 minutes after start-up, it takes around 50 hours to reproduce in the simulator.
+In practice, if a GC bug happens in an application 15 minutes after start-up, it takes around 50 hours to reproduce in the simulator!!
 Fortunately, we work around this problem by using snapshots.
@@
 The VM simulator can then be started just before the crash and the debugging tools can be used in far less than a minute.
-In addition, since the interpreter-only simulator is usually quicker at executing code, if the bug is unrelated to the JIT (typically, a GC bug), the interpreter-only simulator can be used.
+In addition, since the interpreter-only simulator is usually quicker at executing code, and garbage collection simpler because machine code is not scanned for object references, if the bug is unrelated to the JIT (typically, a GC bug), the interpreter-only simulator can be used.
 
@@
 Basically, any calls outside of the machine code generated by the JIT and the Slang code cannot be simulated.
-For specific small parts of the VM, such as file management, we extended the simulator, effectively duplicating the code base with the C code, to support those features in simulation.
+For specific small parts of the VM, such as file management and socket support, we extended the simulator, effectively duplicating the code base with the C code, to support those features in simulation.
 However, there is no solution in the general case: we cannot afford to simulate both the compiled C code and the jitted code on the processor simulator; that would be horribly slow, and specific behaviors in the machine code not present in the code generated by the JIT can hardly be simulated (Access to C variables, OS variables, etc.).
@@
 
-%shorteinig the miscellaneous uses.
+%shortening the miscellaneous uses.
 Besides being a key tool in VM development, the simulator can be used to indirectly allow the VM to be started from source files ~\cite{PharoBootstrap}.
@@
 
-Many VM developers implemented different tools to help them working more efficiently on their VM, but they rarely publish about it.
+Many VM developers have implemented different tools to help them work more efficiently on their VM, but they rarely publish about it.
 For example, the teams working on the Truffle/Graal projects \cite{OneVMToRule} have very convenient development tools for their VMs such as a tool to visualize the intermediate representation of their JIT Graal at different points in the compilation process.
-But, despite an impressive number of publications on the projects, including publications on development tools for the guest languages \cite{TrufflePolyglotDebugger} or on specific parts of the VM \cite{GraalIR2}, no publication seems to be directly related to the VM development tools.
+But, despite an impressive number of publications on the projects, including publications on development tools for the guest languages \cite{TrufflePolyglotDebugger} or on specific parts of the VM \cite{GraalIR2}, no publication seems to be directly related to their VM development tools.
 
@@
 
-%However, so far, most production VMs (Java, Javascript, etc.), even after the huge recent investments in the Javascript VMs by the four major web vendors, are still compiling through the C/C++ compiler and are not metacircular.
+%However, so far, most production VMs (Java, Javascript, etc.), even after the huge recent investments in the Javascript VMs by the four major web vendors, are still written in C/C++ and are not metacircular.
 %Hence, although a metacircular VM has interesting advantages, it is not clear it is that convenient to build a VM in such a way.
@@
 The key advantage of such a design choice is that the RPython code feels like Python code and is relatively quite easy to read write, unlike Slang which feels like C and is as easy to write as C.
-The main drawback is that RPython to C compilation takes way longer than the Slang to C compilation (up to 40 minutes in a recent Macbook pro for the RSqueak VM \cite{RSqueak}, instead of several seconds for Slang).
+The main drawback is that RPython to C compilation takes much longer than the Slang to C compilation (up to 40 minutes in a recent Macbook pro for the RSqueak VM \cite{RSqueak}, instead of several seconds for Slang).
 
@@
 
-%In addition, RPython was originally designed for Python, which does not feature snapshot by default, so they cannot abuse snapshots to work around the simulation slow performance.
+%In addition, RPython was originally designed for Python, which does not feature snapshot by default, so they cannot abuse snapshots to work around slow simulation performance.
 
@@
 This paper introduced and discussed the \openSmalltalkVM simulation infrastructure, used to develop and debug the VM.
-In particular, we described our experiences developing and debugging the full garbage collector and the JIT compiler.
-We believe Simulation is a powerful tool allowing us to reduce the development time and to fix bugs quickly.
+In particular, we described our experiences when developing and debugging the full garbage collector and the JIT compiler.
+We have found that Simulation is a powerful tool allowing us to reduce development time and to fix bugs quickly.
 

### 2018-08-17 paper.tex
Feenk=>feenk
I missed one: the simulator slow performance => slow simulator performance.
minor disambiguation.

@@
  % \department{VM team}              %% \department is recommended
-  \institution{Feenk}            %% \institution is required
+  \institution{feenk}            %% \institution is required
  % \streetaddress{Street1 Address1}
@@
 
-This code is the end of the entry sequence that checks the stack pointer against the stack limit for the end of the current stack page, which is also used to break out of machine code to service interrupts or events such as invoking the scavenger, etc.
+This code is the end of the entry sequence that checks the stack pointer against the stack limit for the end of the current stack page, a check which is also used to break out of machine code to service interrupts or events such as invoking the scavenger, etc.
 \emph{\%rbx} is used as a base register pointing at the variables in interpreter.
@@
 This time may be quite significant in simulation mode.
-In contrast, in \openSmalltalkVM, we can reuse the snapshots to work around the simulation slow performance.
+In contrast, in \openSmalltalkVM, we can reuse the snapshots to work around slow simulation performance.
 

### 2018-08-17 paper.tex
nuke a spurious "out"

@@
 In the interpreter-only simulator, the stack zone is a Smalltalk Array object.
-In the full simulator, the stack zone is in the byte array, to allow access to the stack from generated machine code by the processor simulator\footnote{This difference is historical, the interpreter-only version was implemented first in the easiest way possible, but out with the JIT the Smalltalk stack must be accessible to the processor simulator.}.
+In the full simulator, the stack zone is in the byte array, to allow access to the stack from generated machine code by the processor simulator\footnote{This difference is historical, the interpreter-only version was implemented first in the easiest way possible, out with the JIT the Smalltalk stack must be accessible to the processor simulator.}.
 

### 2018-08-17 paper.tex
Adding a brief Acknowledgements section for those that have done JIT work.
Oops.  A comment was eliding an important sentence about type inference in Slang.
Acknowledge Ryan's contributing the MIPSLE simulator. Fix some capitalisation.

@@
 
-\subsection{Initial JIT Development and New Back Ends}
+\subsection{Initial JIT development and new back ends}
 When the JIT was first being developed and was at a state where bytecoded methods could actually be translated to machine code, development fell into a very productive and enjoyable pattern.
@@
 
-\subsection{Debugging New GC Algorithms and Lemming Debugging}
+\subsection{Debugging new GC algorithms and Lemming Debugging}
 
@@
   %% extraction tools.
+\section*{Acknowledgements}
+The authors would like to thank Ryan Macnak who write the MIPSLE sprocessor simulator and MIPSLE back end.
 %  This material is based upon work supported by the

### 2018-08-17 paper.tex
Optional footnote for Lemming Debugging ;-)

@@
 Hence if a bug occurs or validation fails it will fail in the copy while the original remains as is, yet to be corrupted.
-We then take a fresh copy and use it to debug, repeating the process as many times as we like, an approach we call Lemming Debugging.
+We then take a fresh copy and use it to debug, repeating the process as many times as we like, an approach we call Lemming Debugging \footnote{http://www.romainbrasier.fr/404/}.
 One throws the copy off the cliff and if it gets to the bottom without incident it is descarded and the simulation continues in the original.

### 2018-08-18 paper.tex
Fixed a typo in Acknowledgements
typos

--- 2018-08-17 paper.tex
@@
 By interpreting the Slang code as Smalltalk code, emulating native code using an external processor simulator and simulating the memory using a large byte array, it is possible to simulate the whole VM execution.
-This allows development and debugging of the VM with the Smalltalk development tools resulting in a live programming experience for VM development.\footnote{Arriving at the startup that first funded the OpenSmalltalk-VM, the first author had considerable experience maintaining the 2nd generation Deutsch Schiffmann VM written in C, and was sceptical that the Squeak Simulator would be viable for implementing a JIT.
+This allows development and debugging of the VM with the Smalltalk development tools resulting in a live programming experience for VM development\footnote{Arriving at the startup that first funded the OpenSmalltalk-VM, the first author had considerable experience maintaining the 2nd generation Deutsch Schiffmann VM written in C, and was sceptical that the Squeak Simulator would be viable for implementing a JIT.
 The startup wanted incremental development so as to reap value early and reduce risk, so it asked first for an interpreter using context-to-stack mapping.
-Writing this initial deliverable within the simulator was so much more productive and pleasurable than his previous work in C that he happily decided to stay with the simulator and invest in the necessary machine code simulation infrastructure.}
+Writing this initial deliverable within the simulator was so much more productive and pleasurable than his previous work in C that he happily decided to stay with the simulator and invest in the necessary machine code simulation infrastructure.}.
 
@@
 Hence if a bug occurs or validation fails it will fail in the copy while the original remains as is, yet to be corrupted.
-We then take a fresh copy and use it to debug, repeating the process as many times as we like, an approach we call Lemming Debugging.
+We then take a fresh copy and use it to debug, repeating the process as many times as we like, an approach we call Lemming Debugging%\footnote{http://www.romainbrasier.fr/404/}
+.
 One throws the copy off the cliff and if it gets to the bottom without incident it is descarded and the simulation continues in the original.
@@
 \section*{Acknowledgements}
-The authors would like to thank Ryan Macnak who write the MIPSLE sprocessor simulator and MIPSLE back end.
+The authors would like to thank Ryan Macnak who write the MIPSLE processor simulator and MIPSLE back end.
 %  This material is based upon work supported by the

### 2018-08-18 paper.tex
More typo correction.  Use \emph for all code. inor tweaks to language. Add a footnote detauling the amazingly meta experience of adding socket primitive support to the VM.

@@
 The production VM is derived by translating the core VM code to C.
-As a result, two execution models coexist: simulation, where the Smalltalk code is executed on top of a Smalltalk VM, and production, where the same code is compiled to executable code through a C compiler.
+As a result, two execution models coexist: simulation, where the Smalltalk code is executed on top of a Smalltalk VM, and production, where the same code is compiled to an executable through a C compiler.
 
@@
 %TODO egb: revisit this paragraph after reading all paper to refine problem statement and solution.
-In this paper, we detail the VM simulation infrastructure and we report our experience developing and debugging the garbage collector and the just-in-time compiler within it.
+In this paper, we detail the VM simulation infrastructure and we report our experience developing and debugging the garbage collector and the just-in-time compiler (JIT) within it.
 %We mention some of the limitations and how we worked around them.
@@
 %Before detailing the simulation infrastructure used to develop and debug the \openSmalltalkVM, 
-In this section, we first briefly introduce the compilation pipeline to generate the production VM and Smalltalk snapshots, key elements to understand the design of the simulation infrastructure.
-Then, we describe the memory layout of both the production and simulation runtimes.
-Next, we detail how VM execution is simulated and the interactions with the simulated memory.
+In this section we first briefly introduce the compilation pipeline to generate the production VM and Smalltalk snapshots, key elements to understand the design of the simulation infrastructure.
+Then we describe the memory layout of both the production and simulation runtimes.
+Next we detail how VM execution is simulated and the interactions with the simulated memory.
 In the following subsection, we explain specific aspects of the simulation infrastructure such as the simulation of the machine code generated by the JIT through the processor simulator.
@@
 This first step usually takes several seconds.
-Secondly, the C compiler, depending on the platform, LLVM, GCC or MSVC, translates all the C files (generated C files and platform-specific C files) into an executable.
+Secondly, the C compiler (depending on the platform LLVM, GCC or MSVC) translates all the C files (generated C files and platform-specific C files) into an executable.
 This second step can take up to a few minutes the first time, but usually only takes a few seconds, since object files for unchanged C files do not need to be recreated.
 
-The VM can be configured in two main flavours, interpreter-only or interpreter+JIT.
-Although the version with the JIT is the most widely used in production, the interpreter version is convenient for development purposes and on devices that outlaw JITs.
+The VM can be configured in two main flavours, interpreter-only or interpreter+JIT \footnote{combining an interpreter and a JIT has performance and complexity advantages, as well as complexity disadvantages.
+But discussion is beyond the scope of this paper.}.
+Although the version with the JIT is the most widely used in production, the interpreter version is convenient for development purposes and essential on devices that outlaw JITs.
 For example, debugging the garbage collector or evaluating new language features can be done in the interpreter-only VM, avoiding JIT complexity.
@@
 More precisely, the snapshot includes objects (such as the classes), the compiled methods in the form of bytecodes and the running processes.
-Developing applications consists essentially in writing and editing code, which installs, modifies and removes classes and compiled methods from objects.
+Developing applications consists essentially in writing and editing code, which installs, modifies and removes classes and compiled methods to and from the class hierarchy.
 Programming may be done live, as the application under development is running.
@@
 
-Snapshots can be used to avoid long start-up times when fixing specific bugs.
-The VM can be run to a point where the bug is about to be encountered and the system can be snapshotted at this point.
+Snapshots can also be used to avoid long start-up times when fixing specific bugs.
+The VM can be run to a point where the bug is about to manifest and a snapshot taken.
 Then multiple analyses of the bug can be undertaken by loading the snapshot and resuming execution, either in the normal VM or in the simulator, short cutting the time to reach the bug.
@@
 
-\paragraph{Production memory layout.}Figure \ref{fig:SimulatedMemory} describes, from a high-level perspective, the memory used by \openSmalltalkVM in the prducton VM and its simulated counter-part.
+\paragraph{Production memory layout.}Figure \ref{fig:SimulatedMemory} describes, from a high-level perspective, the memory used by \openSmalltalkVM in the production VM and its simulated counter-part.
 Let us detail briefly the memory used by the production VM, on the top of the figure.
-On the left side, which is usually the low addresses, we can first find Text, the memory section holding the native code of the VM (the compiled code of the interpreter, the memory manager and the JIT itself, but not the code compiled by the JIT).
-Then, Data holds initialized and uninitialized data, mainly the C variables used globally in the VM.
+On the left, low address, side, we can first find Text, the section holding the native code of the VM (the compiled code of the interpreter, the memory manager and the JIT itself, but not the code compiled by the JIT).
+Then, Data holds initialized and uninitialized data, including the VM's global variables.
 
 At higher addresses we find the beginning of the memory managed by the VM.
-At start-up the VM mmaps a memory region used for the executable code generated by the JIT (this optional execution exists only if the JIT is enabled), for new space (initially empty) and for the old objects present in the snapshot plus a little more space for the first \emph{tenures}\footnote{A tenure is the process of promoting young objects to old objects in a generational GC.} to be performed without triggering the full garbage collector.
+At start-up the VM mmaps a memory region used for the executable code generated by the JIT (this optional section exists only if the JIT is enabled), for new space (initially empty) and for the old objects present in the snapshot plus a little more space for the first \emph{tenures}\footnote{A tenure is the process of promoting young objects to old objects in a generational GC.} to be performed without triggering the full garbage collector.
 Later during execution, as old space grows, new memory regions are mmap'ed at higher addresses to store other old objects.
 
-Usually at a very high addresse we find the C stack.
+Usually at a very high address we find the C stack.
 Room in the C stack is allocated at VM start-up to hold the Smalltalk stack zone.
@@
 \paragraph{Simulation memory layout.}In the simulator, the heap is stored as a large contiguous byte array.
-References between objects are indices inside the byte array instead of pointers.
-All the Slang variables, normally translated to C variables, are simulated as Smalltalk objects.
-They use specific wrapper classes, such as \emph{CArrayAccessor} over normal Smalltalk classes, to emulate the C behavior (only array accesses are available in C, not high level iterator APIs, etc.).
+References between objects are indices into the byte array instead of pointers.
+All the Slang variables, normally translated to C variables, are instance variables of Smalltalk objects.
+When required, they use specific wrapper classes, such as \emph{CArrayAccessor}, over normal Smalltalk classes, to emulate the C behavior (only array accesses are available in C, not high level iterator APIs, etc.).
 The Slang code is executed as Smalltalk code.
 The Smalltalk stack zone is represented in \openSmalltalkVM as a doubly linked cycle of stack pages which are maintained by the VM.
-In the interpreter-only simulator, the stack zone is a Smalltalk Array object.
-In the full simulator, the stack zone is in the byte array, to allow access to the stack from generated machine code by the processor simulator\footnote{This difference is historical, the interpreter-only version was implemented first in the easiest way possible, out with the JIT the Smalltalk stack must be accessible to the processor simulator.}.
+In the interpreter-only simulator, the stack zone is a Smalltalk \emph{Array} object.
+In the full simulator, the stack zone is in the byte array, to allow access to the stack from generated machine code\footnote{This difference is historical, the interpreter-only version was implemented first in the easiest way possible, but with the JIT the Smalltalk stack must be accessible to the processor simulator.}.
 
@@
 
-All the Slang code is implemented in multiple Smalltalk classes, to organise the code and to add modularity through polymorphism.
-For example, the \texttt{AbstractCompactor} class has two subclasses, one implementing a sweep algorithm and the other a compact one.
-For production, at Slang-to-C compilation time, all the code is compiled in a single C file.
+The Slang code is implemented in multiple Smalltalk classes, to organise the code, to add modularity through polymorphism, and reuse through inheritance.
+For example, the \emph{AbstractCompactor} class has two subclasses, one implementing a sweep algorithm and the other a compaction one.
+For production, at Slang-to-C compilation time, all the code is compiled into a single C file\footnote{one C file contains the interpreter and memory manager, with one JIT file per back end}.
 All the modularity is removed, using the same example, the VM developer chooses at this moment if he wants to compile a VM with a sweep or compact algorithm.
@@
 However, since polymorphism is available in the simulator, it can be reused for debugging purposes.
-Still in our example, the \texttt{AbstractCompactor} class has also simulation specific subclasses.
+Still in our example, the \emph{AbstractCompactor} class has also simulation specific subclasses.
 Such versions typically express additional constraints in the form of assertions which can be written in plain Smalltalk without restrictions to easily express complex constraints.
@@
 
-In addition to the interpreter simulator, simulating the JIT requires simulating execution of the native code it generates.
+In addition to the interpreter simulator, simulating the JIT requires simulating execution of the machine code it generates.
 The JIT itself is written in Slang and simulated with the Smalltalk execution model.
@@
 
-Bindings to processor simulator libraries (Bochs for x86 and x64, Skyeye for ARMv6, Smalltalk code for 32-bit MIPSLE) were implemented so that the machine code can be executed safely.
+Bindings to processor simulator libraries (Bochs for x86 and x64, Skyeye for ARMv6, Smalltalk code for 32-bit MIPSLE) were implemented so that machine code can be executed safely.
 All but the MIPSLE simulators are accessed through %Smalltalk primitives.
@@
 % to invoke foreign code plus Smalltalk code that is executed if (the code invoked by) the primitive returns an error. 
-The next paragraphs describe in detail how this scheme is used to ensure that simulation and production machine code are as similar as possible and how simulated machine code is interfaced to the rest of the simulation.
+The next paragraphs describe in detail how this scheme is used to ensure that simulation and production machine code are as similar as possible and how machine code is interfaced to the rest of the simulation.
 %, and the debugging facilities for machine code that are available.
@@
 The runtime can then access the current machine code stack frame though these variables.
-In the production VM, the \emph{stackPointer} and \emph{framePointer} are C variables in Data, and routines in the runtime have been compiled from C and exist in Text, and all have fixed addresses.
+In the production VM, the \emph{stackPointer} and \emph{framePointer} are C variables in Data, and routines in the runtime have been compiled from C and exist in Text; all have fixed addresses.
 Referencing these variables and invoking these runtime routines from machine code is therefore straight-forward.
@@
 As code is JITted and specific variables and runtime routines are referenced, the JIT manufactures a unique illegal address to reference each variable or routine, and uses this as the key in the relevant map, creating and storing a suitable closure accessor as the value.
-When the processor simulator attempts to execute an instruction containing one of these illegal addresses, which will be some variety of read, write, call, jump or return instruction\footnote{activations of JITted machine code methods return to calling interpreter frames by returning to a trampoline that transfers control to the interpreter; hence the return address of a machine code activation above an interpreter frame has this trampoline as its return address}, it raises a ProcessorException.
+When the processor simulator attempts to execute an instruction containing one of these illegal addresses, which will be some variety of read, write, call, jump or return instruction\footnote{activations of JITted machine code methods return to calling interpreter frames by returning to a trampoline that transfers control to the interpreter; hence the return address of a machine code activation above an interpreter frame has this trampoline as its return address}, it will return an error from the FFI call.
+Support code analyses the failure and raises a \emph{ProcessorException} which contains all necessary details, the type, pc, next pc and so on.
 %Our processor simulators are invoked through a "primitive" interface, much like an FFI call, and the attempt to raise an exeption is caught by wrapper code that returns an error result from the primitve invocation; in Smalltalk terminology the primitive fails.
-The primitive failure code, Smalltalk code in the method that contains the primitive invocation, then analyses the offending instruction and constructs a Smalltalk ProcessorException object.
+The primitive failure code, Smalltalk code in the method that contains the primitive invocation, then analyses the offending instruction and constructs a Smalltalk \emph{ProcessorException} object.
 This Smalltalk exception has a type (\#read, \#write, \#jump, \#call), the address of the memory location being read, written, jumped to or called, a program counter (pc) holding the instruction's address, a next pc, and an accessor for the register, if any, that is the target of the read or the source of the write.
@@
 In this way, machine code generated in simulation can access arbitrary simulation objects while remaining essentially identical to the generated machine code in the production VM.
-Further, ProcessorException, and a handful of other methods implemented by a processor simulator makes the interfacing of machine code to the Smalltalk part of the simulation processor independent.
+Further, \emph{ProcessorException}, and a handful of other methods implemented by a processor simulator makes the interfacing of machine code to the Smalltalk part of the simulation processor independent.
 
@@
 For example, one can specify that execution should stop at a particular pc if the top two elements on the stack satisfy some criterion.
-It is much more powerful and much simpler to use than typical machine-level debuggers, closures can be created at any point in any tool with a text interface, such as the debugger, and the scheme can be extended as one is debugging (for example, adding interval support during the middle of a debugging session if required).
+It is much more powerful and much simpler to use than typical machine-level debuggers, closures can be created at any point in any tool with a text interface, such as the debugger, and the scheme can be extended as one is debugging (for example, adding interval support during the middle of a debugging session).
 
@@
 In addition, when developing a specific part of the JIT, it is convenient to compile a well-chosen simple method exercising that specific new part.
-This is difficult to do with the full simulator since that new part of the JIT might be called by the start-up logic, leading to the erratic behavior before expected.
+This is difficult to do with the full simulator since that new part of the JIT might be called by the start-up logic, leading to erratic behavior earlier than expected.
 
 To work around these problems, we implemented a tool called \emph{In-image compilation}.
-In-image compilation allows invoking the JIT as a Smalltalk library on any bytecode compiled method in the host Smalltalk system (i.e.
+In-image compilation allows the JIT to be invoked as a Smalltalk library on any bytecode compiled method in the host Smalltalk system (i.e.
 \emph{not} in the simulated heap) to generate the corresponding machine code.
@@
 The Smalltalk code for each processor simulator implements a simple parser that identifies constants and field offsets embedded in that disassembly and passes these to routines in the JIT and the interpreter that lookup addresses, matching them to objects, in particular message selectors and classes, and to method local variable names, rendering the disassembly much more readable than that displayed in a typical machine oriented debugger.
-For example, Figure \ref{fig:ASMDecorated} is a snippet of decorated disassembly for x64.
+For example, Figure \ref{fig:ASMDecorated} is a snippet of decorated disassembly for x64, in this case from an in-image compilation.
 
@@
 These points are specified using metadata implemented as a simple bytecoded language, a stream of such bytes being added to the end of each machine code method.
-This metadata is parsed when the garbage collector needs to locate object references in machine code, when linked sends must be relocated when the machine code zone is compacted to make space by throwing away some number of least recently used methods, and when machine code pcs must be mapped to their corresponding bytecode pcs since at the Smalltalk level only bytecode pcs are seen.
+This metadata is parsed when the garbage collector needs to locate object references in machine code, when linked sends must be relocated as the machine code zone is compacted to make space for new code by throwing away some number of least recently used methods, and when machine code pcs must be mapped to their corresponding bytecode pcs since at the Smalltalk level only bytecode pcs are seen.
 %to preserve the illusion that the VM is a bytecode interpreter using a spaghetti stack of Context objects.
@@
 
-In this section we describe our personal experiences using the simulation infrastructure to develop the VM.
-The different subsections consist in different tasks we had to implement and in each case we emphasise how we used the simulator and how it helped us perform the each given task.
+In this section we describe some memorable and representative personal experiences using the simulation infrastructure to develop the VM.
+The different subsections describe various tasks we undertook and in each case we emphasise how we used the simulator and how it helped us perform each given task.
 
 \subsection{Initial JIT development and new back ends}
-When the JIT was first being developed and was at a state where bytecoded methods could actually be translated to machine code, development fell into a very productive and enjoyable pattern.
-As a bytecode was encountered for which the template had yet to be written execution would stop in the debugger, which would create a skeleton implementation for the template method from the MessageNotUnderstood exception.
+When the JIT was first being developed and was at a state where bytecoded methods could actually be translated to machine code, development fell into a very productive and enjoyable rhythm.
+As a bytecode was encountered for which the template had yet to be written execution would stop in the debugger, which would create a skeleton implementation for the template method from the \emph{MessageNotUnderstood} exception.
 The programmer would implement the body of the method and execution would continue.
-Were the new template definition to be incorrect and cause a bug during JIT generation, rather than when the generated machine code was run, then the debugger could be used to wind execution back to the beginning of the JITting of the current method and resume execution, repeating the cycle until disassembly looked good, somewhat like lemming debugging above.
+Were the new template definition to be incorrect and cause a bug during JIT generation, rather than when the generated machine code was run, then the debugger could be used to wind execution back to the beginning of the JITting of the current method and resume execution, repeating the cycle until disassembly looked good (somewhat like lemming debugging below).
 As a new method was encountered and JITted yet more templates would be encountered.
@@
 Debugging graph rewriting garbage collectors can be hard; bugs leave the heap in a scrambled state, and figuring out what happened when one is left with a corrupted graph, often too broken to traverse and make sense of, or apply tools to, is often effectively impossible.
-But because the simulation is simply an object graph it is easy to copy the entire simulation.
-Therefore the simulator has been extended to by default take a copy of itself on every scavenge or old space reclamation, and to run the collection in the copy, performing a leak check afterwards (if enabled).
+Therefore one is often stuck trying to create a reproducible case in a system that is essentially chaotic: the slightest change in the heap might change the bug; any variability in timing or user input can result in a different heap and hence in the bug morphing or going into hiding.
+
+Here, simulation comes to the rescue.
+Because the simulation is simply an object graph it is easy to copy the entire simulation.
+Therefore the simulator has been extended to by default take a copy of itself on every scavenge or old space reclamation, and to first run the collection in the copy, performing a leak check afterwards (if enabled).
 Hence if a bug occurs or validation fails it will fail in the copy while the original remains as is, yet to be corrupted.
@@
 One throws the copy off the cliff and if it gets to the bottom without incident it is descarded and the simulation continues in the original.
+There is no need to create a reproducible case for bugs encountered during simulation\footnote{Alas not for bugs in the production VM.
+But Lemming Debugging \emph{could} be implemented in C using fork to create the copy in a child process, continuing only if the child successfully completes its GC.
+Our leak checker works both in Simulation and the production VM.}; the reproducible case is at hand.
 This approach was key in rapid development of the new memory manager that supports both 32 and 64 bits \cite{Forwarders}.
@@
 Since we had already a Mark-Compact collector, we mainly introduced a Mark-Sweep non moving GC as an alternative.
-The whole implementation was done in the simulator, and only when it was working there, was it compiled to C.
-Using this process, the compiled C code was working out of the box.
-%To describe the implementation process, we need to discuss briefly first the assertion levels in the VM.
-Then we will show how we debugged the algorithm.
+The whole implementation was done in the simulator, and only when it was working there was it compiled to C.
+Using this process, the compiled C code worked first time.
 
@@
 Since the algorithm was not complete, we added the missing pieces inside the debugger, installing the new code at runtime and checking against the current values of the function parameters if the function behaved as expected, as one can do in any Smalltalk application.
-
-%In addition, the simulator has an interesting property: each time it performs a GC, either a scavenge or an old space collection, it first copies the simulated memory, performs the GC there, and if no exception was raised\footnote{The VM code base includes many assertions to ensure specific state is correct at given points in the execution, such assertions are optionally compiled in the production VM, they represent the majority of exceptions raised during development or bug resolution.}, it then performs the GC on the original version.
-This means that each time the algorithm did not work correctly, the erratic behavior happened in a copy of the heap.
-Since we can clone again the original version as many times as we want, we were able, each time the memory got corrupted beyond repair, to restart the exact same GC execution on the exact same heap with each object at the exact same addresses.
-Indeed, in most cases when a bug happens, the memory was already corrupted and it was very difficult to track back where the bug came from without rewinding the memory state.
-We reproduced the exact same GC many times in different cases to fully understand specific bugs.
 
@@
 In the recent years, we added support for a more aggressive JIT with speculative optimizations through a bytecode to bytecode optimizer \cite{SistaFastStartup}, re-using the existing template JIT as a back-end.
-To be able to generate efficient bytecodes in the bytecode to bytecode optimizer, we had to introduce new unsafe bytecodes allowing, for example, accesses in arrays without type and bounds checks \cite{SistaBytecodeSet}.
+To be able to generate efficient code in the bytecode to bytecode optimizer, we introduced new unsafe bytecodes allowing, for example, accessing of arrays without type and bounds checks \cite{SistaBytecodeSet}.
 For each new bytecode, we introduced new templates in the existing JIT to generate efficient machine code for the optimized methods using the new bytecodes.
@@
 We could figure out which bytecode method was triggering the crash, but we had no idea from which template the crash came from.
-In addition, optimized methods include many inlined methods, making them very large, so it was difficult to figure out where the issue comes from just by looking at thousands of bytecodes.
+In addition, optimized methods include many inlined methods, making them very large, so it was difficult to figure out where the issue came from just by looking at thousands of bytecodes.
 
 To understand the crash, we created a snapshot where the faulty method was executed right after start-up.
-We started the VM simulator, and set-up a conditional breakpoint so that simulator would stop when the JIT would generate that method to machine code.
-Then, when the simulator effectively stopped, we changed the conditional breakpoint to stop execution when the address corresponding to the faulty method entry in machine code would be used, either by a call from the interpreter or through inline cache relinking.
+We started the VM simulator, and set-up a conditional breakpoint so that simulator would stop when the JIT generated machine code for that method.
+Then, when the simulator reached the breakpoint, we changed the conditional breakpoint to stop execution when the address corresponding to the faulty method entry in machine code would be used, either by a call from the interpreter or through inline cache relinking.
 The simulator stopped again, about to execute the machine code corresponding to the faulty method.
@@
 It happens that this object was read from a field on stack, and that this field held an incorrect address.
-We could not tell anymore at this point in the execution what instruction among the thousands previous ones wrote on stack the invalid address.
+We could not tell anymore at this point in the execution what instruction among the thousands of previous ones wrote the invalid address to the stack.
 So we discarded the cloned simulation, and cloned a fresh simulator again, just before the execution of the faulty method in machine code to reproduce again the crash.
-This time, we enabled single stepping (\emph{i.e.,} the processor simulator simulates one instruction at a time) and we added a breakpoint stopping execution when the specific field on stack would be written to the incorrect value found before.
+This time we enabled single stepping (\emph{i.e.,} the processor simulator simulates one instruction at a time) and we added a breakpoint stopping execution when the specific field on stack would be written to the incorrect value found before.
 In this case, the conditional breakpoint is checked in between each machine instruction, and execution stopped right after the machine instruction which wrote the incorrect value on stack.
@@
 A few years ago, we added support in the VM for read-only objects \cite{ROObject}.
-Read-only objects were critical performance-wise for specific customers using them in the context of object databases.
-To maximize the performance, we changed the templates in the JIT compiler for the different memory stores.
+Read-only objects were critical performance-wise for specific customers using the system in the context of object databases.
+To maximize their performance, we changed the templates in the JIT compiler for the different memory stores.
 %Since the JIT has to generate very quickly machine code, it goes through the code three times: (1) it scans the bytecode, (2) it generates abstract instructions from the bytecode and (3) it generates the native instructions from the abstract instructions.
@@
 It is possible, in in-image compilation, to use the Smalltalk debugger on the JIT code itself to inspect the JIT state and fix the JIT code on-the-fly without any major compilation pause.
-Once we went through the few store templates (there are a few different templates for optimizations purposes, for example, storing a constant integer does not require a garbage collector write barrier check), we've just had to evaluate performance and correctness through benchmarks and tests to validate the implementation.
+Once we went through the few store templates (there are a few different templates for optimization purposes, for example, storing a constant integer does not require a garbage collector write barrier check), we've just had to evaluate performance and correctness through benchmarks and tests to validate the implementation.
 
@@
 We then iterated over it and investigated what was in it.
-As show in Table \ref{tab:generalAnalysis}, 1752 methods were compiled to machine code by the JIT, 6352 sends\footnote{We use the Smalltalk terminology, send, to discuss virtual calls since we are talking about Smalltalk.} are present but 2409 of them are not linked (basically, they have never been used).
+As show in Table \ref{tab:generalAnalysis}, 1752 methods were compiled to machine code by the JIT, 6352 sends\footnote{We use the Smalltalk terminology, send, to discuss virtual calls since we are talking about Smalltalk.} are present but 2409 of them are not linked (they have never been executed).
 
@@
 
-Further analysis, in Table \ref{tab:polyAnalysis}, confirms Urs H\"{o}lzle's statement \cite{PICSelf}: around 90\% of used send sites are monomorphic, around 9\% are polymorphic (up to 6 different cases in our implementation) and the remaining \% is megamorphic.
+Further analysis, in Table \ref{tab:polyAnalysis}, confirms Urs H\"{o}lzle's statement \cite{PICSelf}: around 90\% of used send sites are monomorphic, around 9\% are polymorphic (up to 6 different cases in our implementation) and the remaining \% are megamorphic.
 
@@
 
-The results of specific analysis are sometimes used to direct performance design decisions on the VM.
-In this section we describe how the analysis on polymorphic inline caches impacted a design called "Early polymorphic inline cache promotion".
+The results of specific analyses are sometimes used to direct performance design decisions on the VM.
+In this section we describe how the analysis on polymorphic inline caches impacted a hack called "Early polymorphic inline cache promotion".
 
@@
 	\item \emph{Closed PICs:} Such caches can deal with up to 6 cases, and are basically implemented as a jump table.
-	\item \emph{Open PICs:} Such caches can deal with any number of cases, they consist of three probes searching the global look-up cache (a hash map shared with the interpreter) and fall back into a standard look-up routine if nothing is found after three attempts.
+	\item \emph{Open PICs:} Such caches can deal with any number of cases, they consist of three probes searching the global look-up cache (a hash map shared with the interpreter), falling back to a standard look-up routine on miss.
 \end{itemize}
 
-One idea we had was to promote a monomorphic inline cache straight to an open PIC if available, and create the closed PIC only if no open PIC is available for the given selector.
+One idea we had was to promote a monomorphic inline cache directly to an open PIC if available, and create the closed PIC only if no open PIC is available for the given selector.
 The benefit is avoiding lots of code space modifications and an allocation.
 The downside is replacing faster closed PIC dispatch with slower open PIC dispatch.
-The question is how many send sites would be prematurely promoted to megamorphic, or how many closed PICs have selectors for which there are open PICs.
+The question was how many send sites would prematurely be promoted to megamorphic, or how many closed PICs have selectors for which there are open PICs?
 
-The analysis result showed that 17\% of polymorphic send sites would get prematurely promoted.
-So we have implemented a simple sharing scheme.
-The JIT maintains a linked list of open PICs, and before it creates a closed PIC for a send site it will patch it to an open PIC if the list contains one for the send's selector.
+The analysis showed that 17\% of polymorphic send sites would get prematurely promoted.
+So we implemented a simple sharing scheme; the JIT maintains a linked list of open PICs, and before it creates a closed PIC for a send site it will patch it to an open PIC if the list contains one for the send's selector.
 
-Analysing the question is easy in our context.
-This analysis took about an hour to perform, including writing an iterator over send sites, and then implementing the eager promotion to open PICs.
-Somewhat similar analyses one of the authors performed on the second generation Deutsch Schiffman VM \cite{EfficientST80}, which is written entirely in C, when adding the same PIC scheme took several days, having to be written statically and tested with a traditional edit-compile-debug cycle.
+Analysing the question was easy in our context.
+This analysis took about an hour to perform, including writing an iterator over send sites.
+Somewhat similar analyses performed on the second generation Deutsch Schiffman VM \cite{EfficientST80}, which is written entirely in C, when adding the same PIC scheme, took several days, having to be written statically and tested with a traditional edit-compile-debug cycle.
 The productivity difference is extreme.
@@
 
-In this section we first discuss the VM simulator's limitations and how we work around them, and then the interaction between the simulator and bootstrapping processes.
-Lastly, we mention some relevant related work.
+In this section we first discuss some VM simulator limitations and how we work around them.
+Then we cover the interaction between the simulator and bootstrapping processes.
+Lastly we mention some relevant related work.
 
@@
 \paragraph{Performance.}
-The first limitation is due to the simulation performance.
+The first limitation is due to simulation performance.
 The interpreter-only simulator is between a hundred and a thousand times slower to execute code than the normal VM.
-With the JIT and processor simulation enabled, without specific debugging options such as conditional breakpoints in between machine instructions, simulation performance usually drops by a factor of two, though in specific cases it is actually faster.
+With the JIT and processor simulation enabled, without specific debugging options such as conditional breakpoints in between machine instructions, simulation performance usually drops by a factor of two, although in specific cases it is actually faster.
 We usually enable conditional breakpoints in-between machine instructions only when we reach a point in the simulator where the bug is about to happen since single stepping is much slower, so the overall performance in this context is not really relevant.
@@
 Basically, any calls outside of the machine code generated by the JIT and the Slang code cannot be simulated.
-For specific small parts of the VM, such as file management and socket support, we extended the simulator, effectively duplicating the code base with the C code, to support those features in simulation.
+For specific small parts of the VM, such as file management and socket support
+\footnote{Adding socket simulation support was as meta as it gets.
+The Socket primitive support in the Simulator was written above the socket primitives in the production VM.
+An application that used the socket support, the Smalltalk system updating itself from a remote repository, was used to stress the socket primitives.
+Errors in the simulated primitives under development caused primitive failures in the simulation, causing a debugger to open up in the simulation.
+We interacted with this debugger to debug the socket primitives we were writing for the simulator!},
+we extended the simulator, effectively duplicating the code base with the C code, to support those features in simulation.
 However, there is no solution in the general case: we cannot afford to simulate both the compiled C code and the jitted code on the processor simulator; that would be horribly slow, and specific behaviors in the machine code not present in the code generated by the JIT can hardly be simulated (Access to C variables, OS variables, etc.).
@@
 
-This paper introduced and discussed the \openSmalltalkVM simulation infrastructure, used to develop and debug the VM.
-In particular, we described our experiences when developing and debugging the full garbage collector and the JIT compiler.
+This paper introduced and discussed the \openSmalltalkVM's simulation infrastructure, used to develop and debug the VM.
+In particular, we described our experiences when developing and debugging the garbage collector and the JIT compiler.
 We have found that Simulation is a powerful tool allowing us to reduce development time and to fix bugs quickly.
@@
 Currently the tools built on top of the simulation infrastructure are mainly textual.
-More advanced tooling such as moldable inspectors and debuggers \cite{MoldableInspector, MoldableDebugger} should enable a more interactive interface, making it easier to apprehend by new developers.
+More advanced tooling such as moldable inspectors and debuggers \cite{MoldableInspector, MoldableDebugger} should enable a more interactive interface, making it easier to apprehend by new developers, and further improving productivity.
 
@@
 \section*{Acknowledgements}
-The authors would like to thank Ryan Macnak who write the MIPSLE processor simulator and MIPSLE back end.
+The authors would like to thank Ryan Macnak who wrote the MIPSLE processor simulator and back end, as well as much of the Newspeak send machinery.
 %  This material is based upon work supported by the
